---
title: "homework_problem_set_2"
author: "JP"
date: "9/1/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Tell me what each of these symbols mean. Try this without looking at your notes.

$$ \sigma $$
**Answer**

$$ \Sigma $$
**Answer**

$$ \mu $$
**Answer**

$$ \overline{X} $$
**Answer**

$$ S\;or\;s\;or\;S_{X}\;or\;s_{X} $$
**Answer**

$$ f $$
**Answer**

$$ N $$
**Answer**

2. If I were to give you the following scores/values, what would the standard deviation be?

10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10

**Answer**

3. What is the range indirectly useful for finding?

**Answer**

4. What is the difference between variance and standard deviation? What is similar about variance and standard deviation?

**Answer**

5. How do we get around the sum of deviations equaling zero? What are we doing to calculate for variance and standard deviation?

**Answer**

6. What is the area under the curve/proportion at or below the following z-scores?

```{r, echo = FALSE, eval = TRUE}
data.frame(z = c(1.4, -.4, 3.1, .8, -1.7, 2.57, 1.67, 1.11, 0, -2.99))
```

**Answer** For the following z-scores, the area under the curve is:
1.40 = 
-.4 =
3.1 =
.8 =
-1.70 =
2.57 =
1.67 =
1.11 =
0 =
-2.99 =

7. What is the area between the mean and the following z-scores?
```{r, echo = FALSE, eval = TRUE}
data.frame(z = c(1.4, -.4, 3.1, .8, -1.7, 2.57, 1.67, 1.11, 0, -2.99))
```

**Answer** For the following z-scores, the area between the mean and these scores are:
1.40 = 
-.4 =
3.1 =
.8 =
-1.70 =
2.57 =
1.67 =
1.11 =
0 =
-2.99 =

8. What is the percentile for each of the following z-scores?

```{r, echo = FALSE, eval = TRUE}
data.frame(z = c(1.4, -.4, 3.1, .8, -1.7, 2.57, 1.67, 1.11, 0, -2.99))
```

**Answer** For the following z-scores, the percentiles are:
1.40 = 
-.4 =
3.1 =
.8 =
-1.70 =
2.57 =
1.67 =
1.11 =
0 =
-2.99 =

```{r, include = FALSE}
#REMEMBER TO RUN THIS SECTION ALL AT ONCE SO YOU GET THE SAME NUMBERS AS I DO
set.seed(090121)

numbers = rnorm(n = 15, mean = 9.47, sd = 3.18)

N = 15
```

```{r}
numbers
```


$$ s^2_{x} = \frac{\Sigma(X - \overline{X})^2}{N - 1} $$

$$ s_{x} = \sqrt{\frac{\Sigma(X - \overline{X})^2}{N - 1}} $$
1. Calculate the variance and the standard deviation of the estimated population using the two formulas above.

$$ s^2_{x} = \frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N - 1} $$

$$ s_{x} = \sqrt{\frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N - 1}} $$

2. Use the two formulas above to calculate the variance and standard deviation. 

# Range

```{r}
data_example <- data.frame(Maximum = c(30, 35, 40, 98, 47, 51, 61, 91, 41, 50),
                       Minimum = c(20, 29, 7, 64, 40, 34, 50, 14, 10, 12))

data_example

```

# z-Score

$$ z = \frac{X - \overline{X}}{S_{X}} $$

3. Calculate the z-score using the formula above 

```{r}

get_z <- data.frame(observation = c(12, 14, 18, 5, 28 ,30, 14, 6, 14, 15),
                mean = c(15, 16, 24, 7, 21, 24, 10, 8, 9, 4),
                sd = c(3, 2, 4, 3.8, 1.4, 5.1, .9, .47, .21, .5))

get_z

```

# Raw Score From z-Score

$$ X = (z)(S_{X}) + \overline{X} $$

4. Calculate the raw score from the z-score using the formula above. 

```{r}
get_raw <- data.frame(z = c(1, 2, 3, -3, -2, -1, -3.7, 3.1, 1.5, 1.96),
                mean = c(14, 16, 30, 13, 18, 19, 5, 16, 27, 20),
                sd = c(2.1, 3.1, 1.4, 3, 1.2, .2, .9, .47, .16, .67))

get_raw

```

# Standard Error of the Mean

$$ \sigma_{\overline{X}} = \frac{\sigma_{X}}{\sqrt{N}} $$

5. Calculate the standard error of the mean using the formula above. 

```{r}
get_sample_distribution <- data.frame(each_sample_mean = c(17, 15, 24, 61, 9, 10, 14, 18, 30, 4),
                                  mu = c(12, 16, 14, 22, 30, 40, 32, 54, 43, 8),
                                  sigma_x = c(3, 2, 4, 1.4, .2, .6, .7, .1, 2.4, 2.9), #this is not sigma xbar
                                  N = c(30, 15, 60, 17, 15, 34, 43, 60, 58, 81))

get_sample_distribution

```


$$ z = \frac{\overline{X} - \mu}{\sigma_{\overline{X}}} $$

```{r}
get_sample_distribution
```

6. Using the formula above, calculate the z score from the standard error of the mean (**Note** remember that you will be using the standard deviation of the sample mean and not the standard deviation of each sample.)

7. Calculate the estimated population variance and standard deviation using either the formulas used for question 1 or the formulas used for question 2. 

```{r, include = FALSE}
#REMEMBER TO RUN THIS SECTION ALL AT ONCE SO YOU GET THE SAME NUMBERS AS I DO
set.seed(090221)

numbers2 = rnorm(n = 15, mean = 4.14, sd = 1.11)

N = 15

```

```{r}
numbers2
```

