---
title: "PSY 3307"
subtitle: "Z-Scores"
author: "Jonathan A. Pedroza, MS, MA"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(xaringanthemer)
style_duo(primary_color = "#1F4257", secondary_color = "#F97B64")
library(tidyverse)
library(kableExtra)
library(ggpubr)

coffee <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-07/coffee_ratings.csv') %>% 
  mutate(species = as.factor(species),
         process = recode(processing_method, "Washed / Wet" = "washed",
                          "Semi-washed / Semi-pulped" = "not_washed",
                          "Pulped natural / honey" = "not_washed",
                          "Other" = "not_washed",
                          "Natural / Dry" = "not_washed",
                          "NA" = NA_character_),
         process = as.factor(process),
         species = as.factor(species),
         country_of_origin = as.factor(country_of_origin),
         variety = as.factor(variety)) %>% 
  drop_na(process, color)

data <- read_csv("https://raw.githubusercontent.com/jpedroza1228/dissertation/master/final_data/ca_county.csv") %>% 
  janitor::clean_names() %>% 
  select(-x1, -x) %>% 
  filter(county_name != "california")

options(scipen = 999)
theme_set(theme_minimal())

xvalues <- tibble(x = c(-3, 3))

dnorm_one_sd <- function(x){
  norm_one_sd <- dnorm(x)
  # Have NA values outside interval x in [-1, 1]:
  norm_one_sd[x <= -1 | x >= 1] <- NA
  return(norm_one_sd)
}


dnorm_two_sd <- function(x){
  norm_two_sd <- dnorm(x)
  # Have NA values outside interval x in [-1, 1]:
  norm_two_sd[x <= -2 | x >= 2] <- NA
  return(norm_two_sd)
}

dnorm_three_sd <- function(x){
  norm_three_sd <- dnorm(x)
  # Have NA values outside interval x in [-1, 1]:
  norm_three_sd[x <= -3 | x >= 3] <- NA
  return(norm_three_sd)
}

area_one_sd <- round(pnorm(1) - pnorm(-1), 4)
area_one_sd

area_two_sd <- round(pnorm(2) - pnorm(-2), 4)
area_two_sd

area_three_sd <- round(pnorm(3) - pnorm(-3), 4)
area_three_sd

```

# z-Scores

* **relative standing**

* **z-score** is a statistic that indicates the distance a score is from the mean when measured in standard deviation units
  - positive values are above the mean
  - negative values are below the mean
  
* z-score is the location on a distribution as well as the distance from the mean

---

# z-score Example

* Get a score of 90 to the mean of 60
  - deviate of 30

* Standard deviation of 10

* Now means that the +30 score is 3 SD above the mean
  +30 score doesn't tell us much
  +3 sd tells us much more
  - also tells us that the z-score is 3
  
---

# Calculating the z-score

* For a sample 

$$ z = \frac{X - \overline{X}}{S_{X}} $$

* For a population

$$ z = \frac{X - \mu}{\sigma_{X}} $$

---

# Calculating the raw score from a z-score

$$ X = (z)(S_{X}) + \overline{X} $$

---

```{r}
z = 1.4
S_X = 1.14
mean = 3.1

(1.4)*(1.14) + 3.1

```


---

# z-distribution to interpret scores

* **z-distribution** is transforming all raw scores into z-scores

* a z-score of 0 is the mean

* the z-distribution also represents that everything has been standardized
  - every variable is now comparable and on the same scale

---

# Some Reminders

* A z-distribution will always have the same shape as the raw score distribution

* the mean is and will always be zero

* the standard deviation is 1, even when the standard deviation of the raw scores is a different value (e.g., 10, 15, 100)

* z-scores that are greater in either the positive or negative direction mean that values are less likely to occur

---

# Comparing Apples to Oranges

* When using the z-distribution, every variable is put on the same scale

* often referred to as **standardized scores**

---

# z-distributino for Relative Frequency

* relative frequency of z-scores will be the same on all normal z-distributions

* Knowing that +1SD is above the mean, whatever raw scores fall within this part of the distribution will occur 34% of the time

---

```{r, echo = FALSE, eval = TRUE}

ggplot(xvalues, aes(x = x)) + stat_function(fun = dnorm) + 
  stat_function(fun = dnorm_one_sd, geom = "area", fill = "orange", alpha = 0.7) +
  stat_function(fun = dnorm_two_sd, geom = "area", fill = "blue", alpha = 0.5) +
  stat_function(fun = dnorm_three_sd, geom = "area", fill = "green", alpha = 0.3) +
  geom_text(x = .5, y = 0.25, size = 4, fontface = "bold",
            label = paste0((area_one_sd * 100)/2, "%")) +
    geom_text(x = -.5, y = 0.25, size = 4, fontface = "bold",
            label = paste0((area_one_sd * 100)/2, "%")) +
  geom_text(x = 1.5, y = 0.05, size = 4, fontface = "bold",
            label = paste0(round((pnorm(2) - pnorm(1))*100, 2), "%")) +
  geom_text(x = -1.5, y = 0.05, size = 4, fontface = "bold",
            label = paste0(round((pnorm(-1) - pnorm(-2))*100, 2), "%")) +
  geom_text(x = 2.5, y = 0.008, size = 4, fontface = "bold",
            label = paste0(round((pnorm(3) - pnorm(2))*100, 2), "%")) +
  geom_text(x = -2.5, y = 0.008, size = 4, fontface = "bold",
            label = paste0(round((pnorm(-2) - pnorm(-3))*100, 2), "%")) +
  scale_x_continuous(breaks = c(-3:3)) + 
  labs(x = "\nNumber of Standard Deviations From the Mean (z)", y = "f(z) \n", title = "Standard Normal Distribution \n",
       caption = "Thanks to\nhttps://steemit.com/programming/@dkmathstats/creating-normal-distribution-plots-with-r-programming\nfor the assistance with the code.") +
  geom_vline(xintercept = 0, color = "black", linetype = 2, alpha = .3) + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="blue", size = 12),
        axis.title.y = element_text(face="bold", colour="blue", size = 12))
```

---

# Standard Normal Curve

* **Standard Normal Curve** is a perfect normal z-distribution that serves as the model of any approximately normal z-distribution

* Approach is most common with
  - large sample (or population)
  - interval or ratio scores
  - come close to forming a normal distribution

* While z-scores past +-3SD are possible, they occur .0013 of the time, which is why we often only look at +-3SD

---

```{r, echo = FALSE, eval = TRUE}

ggplot(xvalues, aes(x = x)) + stat_function(fun = dnorm) + 
  stat_function(fun = dnorm_one_sd, geom = "area", fill = "orange", alpha = 0.7) +
  stat_function(fun = dnorm_two_sd, geom = "area", fill = "blue", alpha = 0.5) +
  stat_function(fun = dnorm_three_sd, geom = "area", fill = "green", alpha = 0.3) +
  geom_text(x = 0, y = 0.25, size = 4, fontface = "bold",
            label = paste0("<----------------", area_one_sd * 100, "%", "-------------->")) +
  geom_text(x = 0, y = 0.05, size = 4, fontface = "bold",
            label = paste0("<---------------------------------------", area_two_sd * 100, "%", "-------------------------------------->")) +
  geom_text(x = 0, y = 0.008, size = 4, fontface = "bold",
            label = paste0("<--------------------------------------------------------", area_three_sd * 100, "%", "-------------------------------------------------------->")) +
  scale_x_continuous(breaks = c(-3:3)) + 
  labs(x = "\n z", y = "f(z) \n", title = "Standard Normal Distribution \n") +
  geom_vline(xintercept = 0, color = "black", linetype = 2, alpha = .3) + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="blue", size = 12),
        axis.title.y = element_text(face="bold", colour="blue", size = 12))

```

---

# Relative Frequency Using the z-Distribution

* Calculate the z-score for one observation

* Multiply by N

* Answer will be the number of observations between the mean and the z-score you calculated

* Then you can add up the other half to see all the observations that are at or below the score that the one observation had

---

# Example

```{r}
z_score = -1

N = 100

# 
.3413*100

50 - 34.13

#15.87% of scores were at or below the z-score we calculated

```

---

# Using the z-Table

* In your Book at the end, you'll see the z-distribution

* There are two ways to show this
  - Area between mean and the z-score along with area beyond z in tail
  - z-scores with the tenths position on the column and the hundredths position on the top row
  
---

# Using z-Scores to Describe Sample Means

* **Sampling Distribution of Means** is the frequency distribution showing all possible sample means when samples are drawn from a population

* Every time a population is sampled from, they will always be slightly different, with some means being higher and some being lower

* **Central Limit Theorem** statistical concept that defines the mean, standard deviation, and shape of a sampling distribution
  - Always follows some rules
    + A sampling distribution is always an approximately normal distribution
    + mean of sampling distribution equals mean of underlying raw score population used to create the sampling distribution
      + \mu is the mean of the means
    + standard deviation of the sampling distribution is related to the standard deviation of the raw score population
    
---

# Central Limit Theorem

* The use of this concept is that we can assume that our sample is representative of the population without having to sample the whole population

---

# The Standard Error of the Mean

$$ \sigma_{\overline{X}} = \frac{\sigma_{X}}{\sqrt{N}} $$

* **Standard Error of the Mean** is the standard deviation of the sampling distribution of means

* \sigma_{\overline{X}} will be used to describe the true standard error of the mean
  - which is the population standard deviation of the population of sample means

---

# Example

```{r}
# population standard deviation of raw scores is 15
# population mean/mu is 25
sigma_raw = 15

N = 100

15/sqrt(N)

# 1.5
```

* Out of the sampling distribution, the individual sample means differ from the population mean of 25 by an average of 1.5 points

---

# z-Score from a Sample Mean

$$ z = \frac{\overline{X} - \mu}{\sigma_{\overline{X}}} $$

* Using information from the previous slide
  - \mu = 25
  - \overline{X} = 21
  - \sigma_{\overline{X}} = 1.5
  
```{r}
(21 - 25)/1.5

```

---

# Relative Frequency of Sample Means

* Because a sampling distribution is always an approximation of a normal distribution, transforming all sampling means into z-scores means it is a normal z-distribution

---

```{r, echo = FALSE, eval = TRUE}

ggplot(xvalues, aes(x = x)) + stat_function(fun = dnorm) + 
  stat_function(fun = dnorm_one_sd, geom = "area", fill = "orange", alpha = 0.7) +
  stat_function(fun = dnorm_two_sd, geom = "area", fill = "blue", alpha = 0.5) +
  stat_function(fun = dnorm_three_sd, geom = "area", fill = "green", alpha = 0.3) +
  geom_text(x = 0, y = 0.25, size = 4, fontface = "bold",
            label = paste0("<----------------", area_one_sd * 100, "%", "-------------->")) +
  geom_text(x = 0, y = 0.05, size = 4, fontface = "bold",
            label = paste0("<---------------------------------------", area_two_sd * 100, "%", "-------------------------------------->")) +
  geom_text(x = 0, y = 0.008, size = 4, fontface = "bold",
            label = paste0("<--------------------------------------------------------", area_three_sd * 100, "%", "-------------------------------------------------------->")) +
  scale_x_continuous(breaks = c(-3:3)) + 
  labs(x = "\n z", y = "f(z) \n", title = "Standard Normal Distribution \n") +
  geom_vline(xintercept = 0, color = "black", linetype = 2, alpha = .3) + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="blue", size = 12),
        axis.title.y = element_text(face="bold", colour="blue", size = 12))

```

