---
title: "Paired-Samples t-test"
format: beamer
editor: visual
execute:
  echo: true
---

## Repeated Measures

-   with a paired-samples t-test, we are interested in the sampling distribution of difference scores

    -   JP: My mistake, I told you that it was for all t-tests last week

-   since we are looking at pairs of scores for the same participants, we want the difference between the two scores

## Paired-Samples t-test Equation

-   when you want to test different conditions for all participants in a within-subjects design, you'd use a **paired-samples t-test**

    -   Ex: reduce salt intake --> blood pressure --> blood pressure medication --> blood pressure

$$
t = \frac{\overline{D} - \mu_{D}}{\sigma_{\overline{D}}}
$$

$$
t = \frac{\overline{D}}{\sigma_{\overline{D}}}
$$

## Paired-Samples t-test Equation

-   this equation compares the mean difference between our samples to the difference that we would expect to find between population means divided by the standard error of the difference

-   with the assumption that the differences between the population means is consistent with the null hypothesis, the difference between the two population means should be zero

-   due to sampling variation, it is possible to get differences between the condition means that are large but it will happen infrequently

## Paired-Samples t-test Equation

-   the standard deviation of the sampling distribution of differences between means is the **standard error of differences**

    -   small standard error suggests difference between the means is close to the population mean

    -   large standard error suggests that differences between means will be more spread out/distant

## Paired-Samples t-test Equation

-   we should expect that the differences between means would be centered around zero

$$
t = \frac{\overline{D}}{\sigma_{\overline{D}}} = \frac{\overline{D}}{\frac{S_{D}}{\sqrt{N}}}
$$

## Example

```{r}
data <- data.frame(participant = 1:5,
                   t1 = c(10, 8, 7, 9, 10),
                   t2 = c(5, 4, 2, 5, 3))
data
```

## Paired-Samples t-test Equation

-   First, we must calculate the difference between every participants' first time point and second time point

```{r}
10 - 5
8 - 4
7 - 2
9 - 5 
10 - 3
```

## Paired-Samples t-test Equation

```{r}
data$difference <- data$t1 - data$t2
data
```

## Paired-Samples t-test Equation

-   then we'll get the average of the difference scores

```{r}
(5 + 4 + 5 + 4 + 7)/5
```

-   now, we have an average score of 5, so we now need our sum of squares, df, variance/standard deviation to get a standard error of the differences

## Paired-Samples t-test Equation

```{r}
(5 - 5)^2 + (4 - 5)^2 + (5 - 5)^2 + (4 - 5)^2 + (7 - 5)^2
```

-   sum of squares is 6

-   df is 5 - 1= 4

## Paired-Samples t-test Equation

```{r}
6/4
```

-   variance is 1.5

## Paired-Samples t-test Equation

```{r}
sqrt(1.5)
```

-   standard deviation is 1.22

## Paired-Samples t-test Equation

-   now we can fill in the rest of our formula

$$
t = \frac{5}{\frac{1.22}{\sqrt{5}}}
$$

## Paired-Samples t-test Equation

```{r}
sqrt(5)
```

$$
t = \frac{5}{\frac{1.22}{2.24}}
$$

## Paired-Samples t-test Equation

```{r}
1.22/2.24
```

$$
t = \frac{5}{0.54}
$$

## Paired-Samples t-test Equation

```{r}
5/.54
```

$$
t = 9.26
$$

## Paired-Samples t-test Equation

-   could also use the variance with slight changes

$$
t = \frac{5}{\sqrt{\frac{1.5}{5}}}
$$

## Effect Sizes

-   when your t-obtained value passes the t-critical value, you have a significant finding

    -   ***this does not tell you the size of the relationship***

-   you can convert your t-test into an r value (used for correlation, is standardized, we'll get there soon)

$$
r = \sqrt{\frac{t^2}{t^2 + df}}
$$

## Effect Sizes

-   for correlations

    -   .1 = small

    -   .3 = moderate

    -   .5 = large

## Example - Independent Samples

-   from the last set of slides

-   t = -.93

-   df = 10

$$
r = \sqrt{\frac{-.93^2}{-.93^2 + 10}}
$$

## Example - Independent Samples

```{r}
-.93*-.93
```

$$
r = \sqrt{\frac{.86}{.86 + 10}}
$$

## Example - Independent Samples

```{r}
.86+10
```

$$
r = \sqrt{\frac{.86}{10.86}}
$$

## Example - Independent Samples

```{r}
.86/10.86
```

$$
r = \sqrt{.08}
$$

## Example - Independent Samples

```{r}
sqrt(.08)
```

$$
r = .28
$$

## Example - Paired Samples

-   t = 9.26

-   df = 4

$$
r = \sqrt{\frac{-9.26^2}{-9.26^2 + 4}}
$$

## Example - Paired Samples

```{r}
(-9.26)^2
```

$$
r = \sqrt{\frac{85.75}{85.75 + 4}}
$$

## Example - Paired Samples

```{r}
95.75+4
```

$$
r = \sqrt{\frac{85.75}{89.75}}
$$

## Example - Paired Samples

```{r}
85.75/89.75
```

$$
r = \sqrt{.96}
$$

## Example - Paired Samples

```{r}
sqrt(.96)
```

$$
r = .98
$$

## Effect Sizes

-   instead, you could calculate cohen's d, using the two means and the standard deviation of the control group

$$
\hat{d} = \frac{\overline{X}_{1} - \overline{X}_{2}}{S_{2}} 
$$

## Example - Independent Samples

-   X1 = 4.5

-   X2 = 5.5

-   S2 = 1.87

$$
\hat{d} = \frac{4.5 - 5.5}{1.87} 
$$

## Example - Independent Samples

```{r}
4.5-5.5
```

$$
\hat{d} = \frac{-1}{1.87} 
$$

## Example - Independent Samples

```{r}
-1/1.87
```

$$
\hat{d} = -.53 
$$

## Example - Independent Samples

$$
\hat{d} = .53 
$$

## Example - Paired Samples

$$
\hat{d} = \frac{\overline{X}_{1} - \overline{X}_{2}}{S_{2}} 
$$

-   We didn't calculate the means for both conditions

    -   now we have to calculate the means for both conditions

## Example - Paired Samples

```{r}
(10+8+7+9+10)/5 #t1

(5+4+2+5+3)/5 #t2
```

## Example - Paired Samples

-   we also now need our sd for both conditions

```{r}
# t1
(10 - 8.8)^2 + (8 - 8.8)^2 + (7 - 8.8)^2 + (9 - 8.8)^2 + (10 - 8.8)^2

6.8/4

sqrt(1.7)
```

## Example - Paired Samples

```{r}
# t2
(5 - 3.8)^2 + (4 - 3.8)^2 + (2 - 3.8)^2 + (5 - 3.8)^2 + (3 - 3.8)^2

6.8/4

sqrt(1.7)
```

$$
\hat{d} = \frac{8.8 - 3.8}{1.30} 
$$

## Example - Paired Samples

```{r}
8.8-3.8
```

$$
\hat{d} = \frac{5}{1.30} 
$$

## Example - Paired Samples

```{r}
5/1.3

```

$$
\hat{d} = 3.85
$$

## Independent Samples t-test in SPSS

-   Steps

    -   Test assumptions of your outcome by the two groups in your IV

    -   Analyze --> Compare Means --> Independent Samples t-test

        -   Test Variable = Outcome (must be continuous)

        -   Grouping Variable = IV (must only have two groups)

    -   *Defining Groups* is used to create two groups, these should correspond to the values of your IV

    -   you can then get confidence intervals and bootstrapping if you'd like

    -   

        -   if you don't trust your noramlity, you could bootstrap your t-test to get bootstrapped confidence intervals

        -   it will re-estimate your standard error using the number of samples you asked for

-   don't worry about reading the Bayesian test of two independent means (I'm not going to test you on that)

## Paired Samples t-test in SPSS

-   Steps

    -   Test assumptions of your outcomes

    -   Analyze --> Compare Means --> Paired-Samples t-test

        -   We put both conditions/pairs on the same row for *Pair 1*

        -   you can put more pairs of variables underneath and run all your paired t-tests all at once

    -   you can then get confidence intervals and bootstrapping if you'd like

        -   if you don't trust your normality, you could bootstrap your t-test to get bootstrapped confidence intervals

        -   it will re-estimate your standard error using the number of samples you asked for
