---
title: "Research Methods"
subtitle: "PSY 3307"
author: "Jonathan A. Pedroza, PhD"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(xaringanthemer)
style_duo(primary_color = '#56B4E9', secondary_color = 'black')
```

# Agenda

- Greek & Roman
- Variables
- Scales
- Measurement Error
- Steps in Research
- Theories
- Hypotheses
- Validity & Reliability
- Research Design
  - Correlational Designs
  - Experimental Designs
- Methods of Data Collection
- Types of Variation (Generally)
- Randomization

---


# Learning Greek & Roman

- Greek symbols = population

- Roman symbols = sample

$$\Sigma = Sum$$

$$\bar{x} = sample\;mean$$

$$\mu = population\;mean$$

---

# Learning Greek & Roman

$$s^2 = variance$$

$$s = standard\;deviation$$

$$\sigma = population\;standard\;deviation$$

$$\sigma^2 = population\;variance$$

---

# Learning Greek & Roman

$$ N = Total\;Observations\;(Either\;Sample\;or\;Population)$$

$$ f = Frequency $$

$$ n = observations\;for\;categories\;in\;survey $$

---

# Review

- **Variable**
  - anything that can produce different scores; scores that vary

- **Relationship**
  - the connection between two variables

- Three different types of relationships
  - **positive relationship** is when one variable increases, the second variable increases
  - **negative relationship** is when one variable increases, the second variable decreases
  - **no relationship** shows no clear connection between two variables
  
---


# Poll: Is it a Variable?

- Height
- Age
- Class Standing
- Name

---

# Relationships

```{r, eval = TRUE, echo = FALSE, out.width = "400px", message = FALSE, warning = FALSE}
set.seed(082421)
no <- rnorm(1000, mean = 0, sd = 1)
  
no_tibble <- tibble(value1 = no,
                    value2 = no) %>% 
  janitor::clean_names()

scatterplot1 <- read_csv("https://raw.githubusercontent.com/jpedroza1228/dissertation/master/final_data/ca_county.csv") %>% 
  janitor::clean_names() %>% 
  select(-x1, -x) %>% 
  filter(release_year == 2018 & 
           county_name != "california") %>%
  ggplot(aes(access_to_exercise_opportunities, physical_inactivity)) +
  geom_point(color = "purple", alpha = .5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_light()


pos <- no_tibble %>% 
  ggplot(aes(value1, value2)) + 
  geom_point(color = "purple", alpha = .5) +
  theme_light()

nothing <- read_csv("https://raw.githubusercontent.com/jpedroza1228/dissertation/master/final_data/ca_county.csv") %>% 
  janitor::clean_names() %>% 
  select(-x1, -x) %>% 
  filter(release_year == 2018 & 
           county_name != "california") %>% 
  ggplot(aes(some_college, air_pollution_particulate_matter)) + 
  geom_point(color = "purple", alpha = .5) +
  theme_light()

ggpubr::ggarrange(scatterplot1, pos, nothing, nrow = 1)
```

---

# Descriptive & Inferential Statistics

* **Descriptive Statistics** used for organizing and summarizing your data. 

* Tells the reader how the data looks.

* **Inferential Statistics** used to estimate or infer if there is a relationship in your sample data that represents the population studying

* Tells the reader if there is a relationship.

---

# Examples of Descriptive & Inferential Statistics

1. The average amount of physical inactivity in California counties
2. The association between parks and physical activity engaged 
3. The *difference* in the number of liquor stores between San Diego and Los Angeles counties.  
4. The number of males and females in PSY 3307. 

---

# Types of Research

- **Qualitative**
  - testing theories using language
  
- **Quantitative**
  - testing theories with numbers

---

# Poll: Types of Research

- Finding the average age of students that take PSY 3307 
- Conducting interviews with people on why they won't get vaccinated
- Focus groups on improving graduation rates at CPP
- Testing to see if drinking more coffee makes you better at math

---

# Variables

- **Independent variables** are often seen as the cause of some effect
  - Experimental designs = cause
  - Correlational designs = predictor
  - **Predictor variable** predicts an outcome; IV --> DV

- **Dependent variables** are variables that are affected by the independent variable(s)
  - Outcome based on changes from the independent variable
  - Also called **outcome variable**

- Independent variables = IV

- Dependent variables = DV

- This class, most analyses will be focused on one IV and one DV
  - More advanced statistics can have multiples of both

---

# Conditions

* **Conditions** are the categories of the IV to measure DV scores in an experiment

* Example: Examining test scores of students in a class. Half of the class gets a study guide while the other half is given nothing.

IV = Study guide
DV = Test scores
Conditions: 1 = Group receiving study guide; 2 = Group receiving nothing (control group)

* For experiments, you can say "IV had an effect on DV" or the "IV caused DV" but for correlational students that is a big NO NO. Every time JP hears/reads a correlational study that describes the effect of IV on DV he cries a little inside.

---

# Levels of Measurement

- What is being measured and the numbers that represent what is being measured is the **level of measurement**

- **Categorical variables**
  - entities are divided into distinct categories
  - variables that do not allow fractions
  
- Ex: Getting an 94.5 on a exam

- **Continuous variables**
  - a score for each person/distinct score
  - variables that can be measured with decimals
  
- Ex: Being a 4th year compared to a 3rd year student

---

# Categorical Variables

- **Binary variables**
  - there are only two categories (sex)

- **Nominal variables**
  - categorize into groups; categories are not ordered (race/ethnicity groups)
  - can be binary and nominal

- **Ordinal variables**
  - ordered categories
  - categories indicate rank order (race order)

---

# Continuous Variables

- **Interval variables**
  - equal intervals on the variable represent equal differences in measurement
  - zero is not an absolute zero

- **Ratio variables**
  - similar to interval
  - true zero (zero is the smallest number)

---

# Measurement Scale Examples

1. California Counties
2. Temperature measured in Celcius
2b. Temperature measured in Kelvin
3. Olympic Medal Winners (Gold, Silver, Bronze)
4. Money
5. Income Brackets [0-10k, 11-20k, 21-30k, ...]
6. Miles per hour (mpg)
7. 10 most common baby names
8. Grades (A, B, C, D, F)
9. Grades (100-95, 94-90, 89-85, 84-80, ...)
10. Grades (100-0%)

---

# Measurement Error

- **Measurement error** is the discrepancy between the actual value you're trying to measure and the value you obtained

- A participant's ID states a weight of 130lbs

- Their actual weight is 134lbs

- There is a measurement error of 4lbs

- To assess measurement error, we rely on validity and reliability

---

# Steps of Conducting Statistics Within the Research Process

- Initial observation(s)
- Theory creation
- Generate hypothesis
  - identify variables of interest
- Generate predictions
- Collect data
- Analyze data
- Interpret data

---

# Initial Observation

- Something in the real world
  - JP Example: What causes Type-2 diabetes in the United States?
  - **Research Question** is a question about a concept you are interested in answering
  
- read other research
  - helps you get a better idea of what the field is focusing on
  - `psycnet`, `pubmed` (medical), `google scholar`

---

# Theory & Hypotheses

- A **theory** is a general principle or set of principles that explain findings about a topic to then move forward with new hypotheses to be created

- A **hypothesis** is a prediction from a theory
  - plural term is **hypotheses**

- **Falsification** refers to the act of disproving a theory or hypothesis

---

# Theory & Hypothesis Examples

- The Immigrant Paradox
  - Those that immigrate into the United States are healthier than assimilated individuals
  
- Theory: Immigrants are healthier because those that immigrate are healthy enough to immigrate
  
- Hypothesis: Those immigrating are younger than assimilated individuals
  - Immigrants' age is related to how healthy they are

- Possible explanation: Immigrants may look healthier in certain behaviors (healthier eating), but may be unhealthier in other aspects (mental health, physical activity)

---

# Another Example

- Example:
  - IV: Race/Ethnicity
  - DV: Depression

- Racial/ethnic groups will vary in depression levels (two-tailed hypothesis)
- White participants will have higher depression than Latinx/a/o participants (one-tailed hypothesis)

---

# Try It Out Yourself

- Example:
  - IV: Age Groups (Baby, Minor, Young Adult, Adult, Older Adult)
  - DV: Amount of naps in a day
  
- Hypothesis:

---

# Validity & Reliability

- if an instrument is measured how it is supposed to be than it is **valid** or shows **validity**

- if a measure is consistent then it is deemed **reliable** or shows **reliability**

---
class: center, middle
background-image: url(https://upload.wikimedia.org/wikipedia/commons/5/5d/Reliability_and_validity.svg)
background-size: contain

---

# Validity & Reliability

- **criterion validity** is when your instrument should measure what it is supposed to while agreeing with a "gold standard" objective measure
  - Self-reported stress compared to measuring stress physiologically 
  
- **concurrent validity** is assessing your measure is valid by comparing it to another measure
  - Waist Circumference compared to Body Mass Index (weight/height [kg/m^2])
  - type of criterion validity
  
- **preditive validity** is when you are measuring observations at a later point
  - much more complex validity to achieve
  - better measure of criterion validity
  
---

# Validity & Reliability

- **content validity** is examining to see whether your instrument measures the full range of a construct

- Depression
  - not eating
  - not sleeping
  - sleeping too much
  - eating too much
  - sadness
  - lack of energy
  
---

# Validity & Reliability

- two main measures of reliability

- **Test-retest reliability**
  - testing participants with the same measure twice 
  - in theory, they should score the same

- **Internal reliability/consistency** or **Cronbach's alpha**
  - assesses the consistency the findings from each item

$$\alpha = Cronbach's\;Alpha$$

---

# Designs

- **Design** is how a study is organized

- Experimental (manipulate your independent variable to see what effect it has on the dependent variable)
  - True Experiment
  - Quasi Experiment
  - Natural Experiment
  
- Correlational (don't manipulate anything; just look at the relationship between two variables)
  - Correlation
  - Regression

---

# Correlational Designs

- measured either through:
  - **cross-sectional** where measures and other data are gathered at one time point
  - **longitudinal** gathered data over several time points with the same participants or with different cohorts of participants

- issue of **tertium quid**, or the issue of a third (or fourth) variable that may be present that may actually be impacting the relationship/association you were originally interested in

- Example: school --> stress
  - burnout (**confounding variable**) is an example of a extraneous variable that is affecting the main relationship
  
---

# Experimental Designs

- true experiments imply causality; x (IV) causes y (DV)
  - IV is manipulated
  - participants are randomly selected and randomly put into groups

- there are conditions in the IV that are compared when testing for the DV
  - mindfulness -- > attention
  - meditation --> attention
  - control --> attention
  
---

# Methods of Data Collection

- Types of collecting data

- **between-subjects**
  - different groups get different conditions/manipulations

- **within-subjects** or **repeated measures**
  - every participant gets each condition
  - multiple time points and gets the same condition each time

---

# Types of Variation

- **Unsystematic Variation**
  - variation in scores based on unknown factors/reasons
  - participant with a headache may not perform similarly to peers or other time points

- **Systematic Variation**
  - differences in performance in conditions or time points based on experimental manipulation
  - one condition being encouraging words prior to taking an exam compared to a control

---

# Randomization

- **randomization** is the act of separating groups in an objective way
  - it eliminates most other sources of systematic variation
  - creates a strong experiment

---

# Randomization with Within-Subjects Designs

- systematic variation is often seen through **practice effects** and **boredom effects**

- practice effects are when participants perform differently because they remember aspects of the experiment from the previous time point
  - taking an exam when it is the same exact exam

- boredom effects are when participants are too tired or bored from the first time point
  - an exam is long
  
- while impossible to completely remove, **counterbalancing** can reduce these effects
  - randomly having participants take condition 1 before condition 2 or vice versa
  
---

# Randomization with Between-Subjects Designs

- best option is to simply randomly put participants in the conditions/groups