---
title: "z-scores"
subtitle: "PSY 3307"
author: "Jonathan A. Pedroza, PhD"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, include = FALSE}
library(tidyverse)

theme_set(theme_light())
options(scipen = 999)

xvalues <- tibble(x = c(-3, 3))

dnorm_one_sd <- function(x){
  norm_one_sd <- dnorm(x)
  # Have NA values outside interval x in [-1, 1]:
  norm_one_sd[x <= -1 | x >= 1] <- NA
  return(norm_one_sd)
}


dnorm_two_sd <- function(x){
  norm_two_sd <- dnorm(x)
  # Have NA values outside interval x in [-1, 1]:
  norm_two_sd[x <= -2 | x >= 2] <- NA
  return(norm_two_sd)
}

dnorm_three_sd <- function(x){
  norm_three_sd <- dnorm(x)
  # Have NA values outside interval x in [-1, 1]:
  norm_three_sd[x <= -3 | x >= 3] <- NA
  return(norm_three_sd)
}

area_one_sd <- round(pnorm(1) - pnorm(-1), 4)
area_one_sd

area_two_sd <- round(pnorm(2) - pnorm(-2), 4)
area_two_sd

area_three_sd <- round(pnorm(3) - pnorm(-3), 4)
area_three_sd
```


# Agenda

- distributions
- z-scores
- using z-scores

---

# Distributions

- raw distribution is where the scores of a distribution are on whatever scale of the variable you are looking at
  - Ex: looking at number of minutes studying for all students in your class/sample

- z-score distribution is a distribution where all the scores are in standard deviation units

- probability distribution is the distribution ranges from 0 to 1 of the likelihood of something happening (0 = not happening, 1 = event will definitely happen)
  - **Probability Density Functions** is a mathematical formula to specify the distributions

---

# z-score

### Sample z-score

$$z = \frac{X - \overline{X}}{S}$$

### Population z-score

$$z = \frac{X - \overline{X}}{\sigma}$$

---

# Getting raw scores from a z-score

$$X = (z)(S) + \overline{X}$$

---

# z-scores 

- **z-score** is a statistic that indicates the distance a score is from the mean when measured in standard deviation units
  - positive values are above the mean
  - negative values are below the mean
  
---

# z-score Example

- Get a score of 90 to the mean of 60
  - deviate of 30

- Standard deviation of 10
 
- Now means that the +30 score is 3 SD above the mean
  -  30 score doesn't tell us much
  - 3 sd tells us much more
  - also tells us that the z-score is 3

---

# z-distribution to interpret scores

- **z-distribution** is transforming all raw scores into z-scores

- a z-score of 0 is the mean

- the z-distribution also represents that everything has been standardized
  - every variable is now comparable and on the same scale
  
---

# JP Reminders

- A z-distribution will always have the same shape as the raw score distribution

- the mean is and will always be zero

- the standard deviation is 1, even when the standard deviation of the raw scores is a different value (e.g., 10, 15, 100)

- z-scores that are greater in either the positive or negative direction mean that values are less likely to occur

---

# Comparing Apples to Oranges

- When using the z-distribution, every variable is put on the same scale

- often referred to as **standardized scores**

---

```{r, echo = FALSE, eval = TRUE}
library(tidyverse)
ggplot(xvalues, aes(x = x)) + stat_function(fun = dnorm) + 
  stat_function(fun = dnorm_one_sd, geom = "area", fill = "orange", alpha = 0.7) +
  stat_function(fun = dnorm_two_sd, geom = "area", fill = "blue", alpha = 0.5) +
  stat_function(fun = dnorm_three_sd, geom = "area", fill = "green", alpha = 0.3) +
  geom_text(x = .5, y = 0.25, size = 4, fontface = "bold",
            label = paste0((area_one_sd * 100)/2, "%")) +
    geom_text(x = -.5, y = 0.25, size = 4, fontface = "bold",
            label = paste0((area_one_sd * 100)/2, "%")) +
  geom_text(x = 1.5, y = 0.05, size = 4, fontface = "bold",
            label = paste0(round((pnorm(2) - pnorm(1))*100, 2), "%")) +
  geom_text(x = -1.5, y = 0.05, size = 4, fontface = "bold",
            label = paste0(round((pnorm(-1) - pnorm(-2))*100, 2), "%")) +
  geom_text(x = 2.5, y = 0.008, size = 4, fontface = "bold",
            label = paste0(round((pnorm(3) - pnorm(2))*100, 2), "%")) +
  geom_text(x = -2.5, y = 0.008, size = 4, fontface = "bold",
            label = paste0(round((pnorm(-2) - pnorm(-3))*100, 2), "%")) +
  scale_x_continuous(breaks = c(-3:3)) + 
  labs(x = "\nNumber of Standard Deviations From the Mean (z)", y = "f(z) \n", title = "Standard Normal Distribution \n",
       caption = "Thanks to\nhttps://steemit.com/programming/@dkmathstats/creating-normal-distribution-plots-with-r-programming\nfor the assistance with the code.") +
  geom_vline(xintercept = 0, color = "black", linetype = 2, alpha = .3) + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="blue", size = 12),
        axis.title.y = element_text(face="bold", colour="blue", size = 12))
```

---

# Standard Normal Curve

- **Standard Normal Curve** is a perfect normal z-distribution that serves as the model of any approximately normal z-distribution

- Approach is most common with
  - large sample (or population)
  - interval or ratio scores
  - come close to forming a normal distribution

- While z-scores past +-3SD are possible, they occur .0013 of the time, which is why we often only look at +-3SD

---

```{r, echo = FALSE, eval = TRUE}

ggplot(xvalues, aes(x = x)) + stat_function(fun = dnorm) + 
  stat_function(fun = dnorm_one_sd, geom = "area", fill = "orange", alpha = 0.7) +
  stat_function(fun = dnorm_two_sd, geom = "area", fill = "blue", alpha = 0.5) +
  stat_function(fun = dnorm_three_sd, geom = "area", fill = "green", alpha = 0.3) +
  geom_text(x = 0, y = 0.25, size = 4, fontface = "bold",
            label = paste0("<----------------", area_one_sd * 100, "%", "-------------->")) +
  geom_text(x = 0, y = 0.05, size = 4, fontface = "bold",
            label = paste0("<---------------------------------------", area_two_sd * 100, "%", "-------------------------------------->")) +
  geom_text(x = 0, y = 0.008, size = 4, fontface = "bold",
            label = paste0("<--------------------------------------------------------", area_three_sd * 100, "%", "-------------------------------------------------------->")) +
  scale_x_continuous(breaks = c(-3:3)) + 
  labs(x = "\n z", y = "f(z) \n", title = "Standard Normal Distribution \n") +
  geom_vline(xintercept = 0, color = "black", linetype = 2, alpha = .3) + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="blue", size = 12),
        axis.title.y = element_text(face="bold", colour="blue", size = 12))

```

---

# Relative Frequency Using the z-Distribution

- Calculate the z-score for one observation

- Multiply by N

- Answer will be the number of observations between the mean and the z-score you calculated

- Then you can add up the other half to see all the observations that are at or below the score that the one observation had

---

# Example

```{r}
z_score = -1

N = 100

# 
.3413*100

50 - 34.13

#15.87% of scores were at or below the z-score we calculated

```

---

# Using the z-Table

- In your Book at the end, you'll see the z-distribution

- There are two ways to show this
  - Area between mean and the z-score along with area beyond z in tail
  - z-scores with the tenths position on the column and the hundredths position on the top row

---

# Using z-Scores to Describe Sample Means

- **Sampling Distribution of Means** is the frequency distribution showing all possible sample means when samples are drawn from a population

- Every time a population is sampled from, they will always be slightly different, with some means being higher and some being lower

- **Central Limit Theorem** statistical concept that defines the mean, standard deviation, and shape of a sampling distribution
  - Always follows some rules
    + A sampling distribution is always an approximately normal distribution
    + mean of sampling distribution equals mean of underlying raw score population used to create the sampling distribution
      + \mu is the mean of the means
    + standard deviation of the sampling distribution is related to the standard deviation of the raw score population
    
---

# Central Limit Theorem

- The use of this concept is that we can assume that our sample is representative of the population without having to sample the whole population
