---
title: "Spring 2022 Course Syllabus"
subtitle: "PSY 3307-01/3307A-01: Statistics for Psychology"
author: "Jonathan Pedroza PhD"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{css my-header-colors, echo = FALSE}
.page-header {
  color: #F0E442;
  text-align: center;
  background-color: #005C55;
  background-image: linear-gradient(120deg, #7DA37B, #005C55);
}

a {
  color: #009E73; #color of linked inline code
  font-weight: bold;
  text-decoration: none;
}
  a:hover {
    text-decoration: underline;
    }

.main-content {
  word-wrap: break-word; }
  .main-content :first-child {
    margin-top: 0; }
  @media screen and (min-width: 64em) {
    .main-content {
      max-width: 64rem;
      padding: 2rem 6rem;
      margin: 0 auto;
      font-size: 1.1rem; } }
  @media screen and (min-width: 42em) and (max-width: 64em) {
    .main-content {
      padding: 2rem 4rem;
      font-size: 1.1rem; } }
  @media screen and (max-width: 42em) {
    .main-content {
      padding: 2rem 1rem;
      font-size: 1rem; } }
  .main-content img {
    max-width: 100%; }
  .main-content h1,
  .main-content h2,
  .main-content h3,
  .main-content h4,
  .main-content h5,
  .main-content h6 {
    margin-top: 2rem;
    margin-bottom: 1rem;
    font-weight: normal;
    color: #3F7F68; } # header color
  .main-content p {
    margin-bottom: 1em; }
  .main-content code {
    padding: 2px 4px;
    font-family: Consolas, "Liberation Mono", Menlo, Courier, monospace;
    font-size: 0.9rem;
    color: #005C55; #package color/whatever is in `` brackets
    background-color: #7DA37B; #background color of ``brackets
    border-radius: 0.3rem; }
  .main-content pre {
    padding: 0.8rem;
    margin-top: 0;
    margin-bottom: 1rem;
    font: 1rem Consolas, "Liberation Mono", Menlo, Courier, monospace;
    color: #567482;
    word-wrap: normal;
    background-color: #f3f6fa; #background color of code chunks
    border: solid 1px #005C55; #this is the border of the r code chunks
    border-radius: 0.3rem; }
    .main-content pre > code {
      padding: 0;
      margin: 0;
      font-size: 0.9rem;
      color: #5D6D7E;
      word-break: normal;
      white-space: pre;
      background: transparent;
      border: 0; }
  .main-content .highlight {
    margin-bottom: 1rem; }
    .main-content .highlight pre {
      margin-bottom: 0;
      word-break: normal; }
  .main-content .highlight pre,
  .main-content pre {
    padding: 0.8rem;
    overflow: auto;
    font-size: 0.9rem;
    line-height: 1.45;
    border-radius: 0.3rem;
    -webkit-overflow-scrolling: touch; }
  .main-content pre code,
  .main-content pre tt {
    display: inline;
    max-width: initial;
    padding: 0;
    margin: 0;
    overflow: initial;
    line-height: inherit;
    word-wrap: normal;
    background-color: transparent;
    border: 0; }
    .main-content pre code:before, .main-content pre code:after,
    .main-content pre tt:before,
    .main-content pre tt:after {
      content: normal; }
  .main-content ul,
  .main-content ol {
    margin-top: 0; }
  .main-content blockquote {
    padding: 0 1rem;
    margin-left: 0;
    color: #819198;
    border-left: 0.3rem solid #dce6f0; }
    .main-content blockquote > :first-child {
      margin-top: 0; }
    .main-content blockquote > :last-child {
      margin-bottom: 0; }
  .main-content table {
    display: block;
    width: 100%;
    overflow: auto;
    word-break: normal;
    word-break: keep-all;
    -webkit-overflow-scrolling: touch; }
    .main-content table th {
      font-weight: bold; }
    .main-content table th,
    .main-content table td {
      padding: 0.5rem 1rem;
      border: 1px solid #e9ebec; }
  .main-content dl {
    padding: 0; }
    .main-content dl dt {
      padding: 0;
      margin-top: 1rem;
      font-size: 1rem;
      font-weight: bold; }
    .main-content dl dd {
      padding: 0;
      margin-bottom: 1rem; }
  .main-content hr {
    height: 2px;
    padding: 0;
    margin: 1rem 0;
    background-color: #5E9171; #line to break up sections
    border: 0;
}
```

## Meeting Time

### Lecture: **T/Th 2:30pm-3:45pm**

### Activity: **T/Th 4pm-4:50pm**

-----------------------------------

### Instructor: Jonathan A. Pedroza PhD

### Pronouns: he/him/his

### Preferred Name(s): JP, Jon

### Email: japedroza@cpp.edu

-----------------------------------

### Office: Virtual

### Office Hours: **T 10am-12pm**, email if time does not work

-----------------------------------

## Required Materials: Basic scientific calculator

## Optional Materials: [CRAN R for calculations](https://ftp.osuosl.org/pub/cran/)

### Required Text: [Field, A. (2017). *Discovering Statistics Using IBM SPSS Statistics. 5th Edition.* Sage Publications Ltd.](https://www.google.com/search?q=andy+field+discovering+statistics+5th+edition&rlz=1C1CHBF_enUS926US926&sxsrf=AOaemvLlNNbsNCmLa8NIna-w6WzDv8Acyw%3A1642800531521&ei=kyXrYaejH5CmmAXrkbLwBA&oq=andy+field+5th+edition+discovering+statistics&gs_lcp=Cgdnd3Mtd2l6EAEYADIGCAAQFhAeMgYIABAWEB4yBggAEBYQHjIGCAAQFhAeMgYIABAWEB4yBggAEBYQHjoHCAAQRxCwAzoFCAAQgARKBAhBGABKBAhGGABQ5QRY7RhggyNoAXACeACAAbUBiAGuGpIBBDAuMjOYAQCgAQHIAQjAAQE&sclient=gws-wiz)

## SPSS: [CPP Virtual Lab](https://www.cpp.edu/it/virtual-software-lab/index.shtml)

## OneDrive: [OneDrive Recordings](https://livecsupomona-my.sharepoint.com/:f:/g/personal/japedroza_cpp_edu/EkfOZj9UHtFDjvZbwdtdz_IBNGBS7vExpOzqfvO8aa_1SQ?e=Wc6TAl)

-----------------------------------

## Corequisites: PSY 3307 & PSY 3307A

## Prerequisites: STA 1200 & PSY 2204

-----------------------------------

## Zoom Information: 

### Lecture & Activity: https://cpp.zoom.us/j/83196561554

### Meeting ID: 831 9656 1554

-----------------------------------

# Course Description

Analytic techniques and inferential statistics useful to behavioral scientists. z and t-tests, introduction to analysis of variance, correlational designs, and selected non-parametric statistics. Selection, application, and interpretation of appropriate statistics for analysis of behavioral data.

-----------------------------------

# Course Structure

This course will be conducted online (i.e., no face-to-face meetings) in a synchronous format (i.e., lectures and activities scheduled at specific times). Course content will be provided via Canvas. Lectures and activities will be conducted via Zoom. Lectures will include presentations with worked example problems, followed by practice problems. Recordings of each presentation portion will be posted on Canvas. Activities will include problem set discussion, practice problems, and SPSS tutorials. 

-----------------------------------

# Course Objectives

Upon completion of this course, students will be able to: 

* Understand and calculate descriptive statistics
* Understand and calculate inferential statistics
* Use SPSS to compute statistics
* Read and interpret basic statistics used in the psychology literature
* Develop a foundation for success in higher level statistics and research methods courses

-----------------------------------

# Grading Policy

**Overview**: 

A total of 500 points will be available for this course based on 3 exams, 12 problem sets, and 4 SPSS assignments.

```{r, include = FALSE}
table <- tibble::tibble(Assignments_Tests = c("Exam", "Problem Sets", "SPSS Assignments", "Total Class Points"),
                        Amount = c("3", "12", "4", ""),
                        Points_Each = c("100",  "10", "20", ""),
                        Total_Points = c("300", "120", "80", "500"))

library(tidyverse)
library(kableExtra)

```
 
```{r, echo = FALSE, eval = TRUE}
kbl(table) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"),
                position = "center")
```
 
**Final Grade:** 
Your final grade will be the same for the lecture and activity based on the following scale: 

```{r, include = FALSE}
table2 <- tibble::tibble(Letter_Grade = c("A", "A-",
                                          "B+", "B", "B-",
                                          "C+", "C", "C-",
                                          "D+", "D", "D-",
                                          "F"),
                         Points = c("457.5 - 500", "447.5-457",
                                    "437.5-447", "407.5-437", "397.5-407",
                                    "387.5-397", "357.5-387", "347.5-357",
                                    "337.5-347", "307.5-337", "297.5-307",
                                    "0-297.5"),
                         Percentages = c("91.5-100", "89.5-91.4",
                                    "87.5-89.4", "81.5-87.4", "79.5-81.4",
                                    "77.5-79.4", "71.5-77.4", "69.5-71.4",
                                    "67.5-69.4", "61.5-67.4", "59.5-61.4",
                                    "0-59.5"))
```

```{r, echo = FALSE, eval = TRUE}
kbl(table2) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"),
                position = "center")
```

-----------------------------------

# **Assignments**

## **Exams:** 

There will be `3` take-home exams, each worth up to `100 points`. Exams will contain vignettes and statistical problems to solve. To earn full credit, you must show all steps taken in a problem to arrive at your answer (either by hand with attached photos or by providing an R script). Exams will be open-book/open-note but must be completed individually. Each exam will be on `Canvas` where you will have a 48-hour time window from when it becomes available until it is due. No late exams will be accepted. 

## **Problem Sets:** 

There will be `12` problem sets, each worth `10 points`. These problem sets can be submitted by either completing the problems by hand and attaching photos of your work on `Canvas` or by submitting a R script. Completed assignments will be awarded full credit. The first 10-15 minutes of class following a due problem set will go over the answers. Answer keys with descriptions will be available following the class period that discusses the problem set. No late problem sets will be accepted.  

## **SPSS Assignments:** 

There will be `4` SPSS assignments, each worth up to `20 points`. Collaboration is encouraged; however, every student must turn in their own assignment. Each assignment will consist of:

SPSS Assignment 1: t-test

SPSS Assignment 2: ANOVA

SPSS Assignment 3: ANOVA

SPSS Assignment 4: Regression

1. Research Question & Hypothesis
2. Recoding Variables
3. Descriptive Statistics
4. Descriptive Visualization
5. Inferential Statistic
6. Visualization of Statistic Finding
7. Write-up of Inferential Statistic

SPSS can be accessed on your personal computer by using the CPP virtual lab (https://www.cpp.edu/it/virtual-software-lab/index.shtml), downloading the free 14-day trial from IBM (https://www.ibm.com/analytics/spss-trials), or purchasing the Standard GradPack 26 6-mo rental (https://cpp.onthehub.com/). Answer keys with descriptions the day after the submission. No late SPSS assignments will be accepted. 

-----------------------------------

# **Other Policies**

## **Academic Integrity:** 

Students are expected to adhere to the University’s Student Conduct Code (https://www.cpp.edu/studentconduct/student-conduct-code.shtml). Academic integrity violations, including, but not limited to, cheating and plagiarism, may result in a 0 for an assignment or exam and will be reported to the Office of Student Conduct & Integrity.

## **Accommodations:** 

Reasonable accommodations will be provided for students with learning, physical, or other disabilities. Accommodations approved through the Disability Resource Center (909-869-3333) should be discussed with the instructor early in the semester to ensure appropriate implementation. 

## **Distribution of Course Materials:** 

Do not copy/screenshot or retain any exam questions. Violation of this policy is grounds for disciplinary action. Distribution of other course materials, including slides, video lectures, SPSS assignments, and problem sets, is acceptable. 

## **Mandatory Reporting:**

Please note two executive orders from the CSU’s Office of the Chancellor that limit what information faculty members can keep confidential. **Executive Order 1083 relates to reporting of child abuse** and **Executive Order 1096 relates to reporting of campus sexual assaults**. CSU faculty members are considered mandatory reporters. Once we are made aware of such incidents, we are required to report the incident to our Title IX officer on campus, regardless of whether the student wants the information reported.

## **Student Health & Well-being:**

Students experiencing emotional distress may seek services through [Counseling & Psychological Services](https://www.cpp.edu/caps/index.shtml). Additionally, there is the [Student Health & Wellbeing](https://www.cpp.edu/health/student-health-services.shtml) for other health issues. 

-----------------------------------

# Still In a Pandemic

## Cameras

If you feel comfortable please turn on your camera. However, I understand that you may be in a situation where you may not be able to have your camera on. I would like to see your faces, simply because I want to gauge understanding. This is much easier by seeing the looks on your faces.

## Participation

Please interrupt at any time if you do not understand anything...and I mean ANYTHING. This class includes the essentials for a lot of research and practical applications in both the social and biological sciences. It is important to understand as much as possible regarding the statistical tests that we will learn during this semester. Communication is key for this class. I will also offer office hours to discuss class assignments. **Please email me if the office hours do not work for you.**

## Pandemic-related Accomodations

We are still in a pandemic so please communicate with me if you are not able to meet the requirements of this class. While no late assignments will be accepted, accommodations due to pandemic-related issues will be honored. If there is no prior communication about not being able to meet the requirements for an assignment, a zero for that assignment will be assigned. 

## Classes Will Be Recorded

All classes will be recorded with automated closed captions and a full automated transcription of each lecture. Please email me if you have any concerns regarding the lecture being recorded to see if there are any accomodations that can be made.

-----------------------------------

# **Tenative Course Schedule**

{{< bootstrap-table "table table-hover" >}}

| **Week** | **Dates** | **Lecture Topics** | **Activity Topics** | **Readings** | **Assignments Post Date and Due Dates (@ 11:59pm)** |
| :----------------------- | :-----------------------: | :-----------------------: | :-----------------------: | :----------------------- |  :----------------------- |
| Week 1a | 01/25/22 | Introduction and review | Learning R as a calculator | Ch.1 |  |
| Week 1b | 01/27/22 | Research methods | SPSS Practice | Ch.1 |  |
| Week 2a | 02/01/22 | Frequencies, central tendency, measures of variability | Frequencies, measures of central tendency, and variability  | Ch.1 | Problem Set 1 |
| Week 2b | 02/03/22 | z-scores | Frequencies, measures of central tendency, and variability | Ch.1 | Problem Set 2 |
| Week 3a | 02/08/22 | Modeling, populations, samples, and the standard error | Frequencies, measures of central tendency, and variability | Ch.2  | *Problem Sets 1 and 2 due* |
| Week 3b | 02/10/22 | Modeling, populations, samples, and the standard error | sampling calculations | Ch.2 | Problem Set 3 |
| Week 4a | 02/15/22 | Modeling, populations, samples, and the standard error | z-test and one-sample t-test | Ch.2 | *Problem Set 3 due* |
| Week 4b | 02/17/22 | Null-hypothesis significance testing | z-test and one-sample t-test | Ch.2 | Problem Set 4 |
| Week 5a | 02/22/22 | Study session | Study Session | No readings | *Problem Set 4 due* |
| Week 5b | 02/24/22 | Exam 1 | Exam 1 | No readings |  |
| Week 6a | 03/01/22 | Examining Bias | SPSS Assumptions Practice | Ch.6 |  |
| Week 6b | 03/03/22 | Examining Bias | SPSS Assumptions Practice | Ch.6 |  |
| Week 7a | 03/08/22 | Examining Bias | SPSS Assumptions Practice | Ch.6 | Problem Set 5 |
| Week 7b | 03/10/22 | Independent-samples t-test | t-test calculations | Ch.10 | Problem Set 6 |
| Week 8a | 03/15/22 | Independent-samples t-test | SPSS t-test Practice | Ch.10 | *Problem Sets 5 & 6 due*/SPSS Assignment 1 |
| Week 8b | 03/17/22 | Paired-samples t-test | t-test calculations | Ch.10 | Problem Set 7 |
| Week 9a | 03/22/22 | One-way ANOVA | ANOVA calculations | Ch.12 | *Problem Set 7 and SPSS Assignment 1 due*/Problem Set 8 |
| Week 9b | 03/24/22 | One-way ANOVA | ANOVA calculations/SPSS ANOVA Practice | Ch.12 | Problem Set 9/SPSS Assignment 2 |
| Week 10a | 03/29/22 | Spring Break | Study Guide Provided |  |  |
| Week 10b | 03/31/22 | Spring Break |  |  | |
| Week 11a | 04/05/22 | Exam 2 | Exam 2 | No readings | *Problem Set 8 & 9 due* |
| Week 11b | 04/07/22 | Two-way ANOVA | ANOVA calculations | Ch.14 | *SPSS Assignment 2 due* |
| Week 12a | 04/12/22 | Two-way ANOVA | ANOVA calculations/SPSS ANOVA Practice | Ch.14 | SPSS Assignment 3 |
| Week 12b | 04/14/22 | Two-way ANOVA/Repeated Measures | SPSS ANOVA Practice | Ch.14 and Ch.15  |  |
| Week 13a | 04/19/22 | Repeated Measures & Mixed-design ANOVA | ANOVA Calculations | Ch.15 and Ch.16 |  |
| Week 13b | 04/21/22 | Repeated Measures & Mixed-design ANOVA | SPSS ANOVA Practice | Ch.15 and Ch.16 | Problem Set 10/*SPSS Assignment 3 due* |
| Week 14a | 04/26/22 | Correlation | Correlation calculations | Ch.8 | Problem Set 11/*Problem Set 10 due* |
| Week 14b | 04/28/22 | Correlation and regression | Correlation and Regression SPSS Practice | Ch.8 and Ch.9 | SPSS Assignment 4 |
| Week 15a | 05/03/22 | Regression | Regression calculations | Ch.9 | Problem Set 12/*Problem Set 11 due* |
| Week 15b | 05/05/22 | Study session | Study session | No readings | *Problem Set 12 and SPSS Assignment 4 due* |
| Finals Week | 05/09/22 - 05/13/22 | Exam 3 | Exam 3 | No readings |  |

{{< /bootstrap-table >}}
