---
title: "R Download Guide"
author: "JP"
date: "8/14/2021"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Guide to Downloading R
[Download Link Here](https://ftp.osuosl.org/pub/cran/)

## Why am I recommending you use R for this course?

1. Because R is just a huge calculator.
```{r basic calculations, eval = TRUE, echo = TRUE}
options(scipen=999)

3 + 5

tictoc::tic()
423164572464815794858642 + 423164572464815794858642
tictoc::toc()

423164572464815794858642*2


```

2. It's free. 

3. It also lets you save parts in what are called `objects`. 

You can either use arrows or equal signs to create objects. The objects will allow you to save every single part of the functions you will be writing. For example, below I will show you the steps taken to get the standard deviation of some simulated test scores. But to get the standard deviation, you'll first have to do some steps that are shown in your textbook. 

First, I'm going to create some fake data so we can move forward. Below is to show that using `=` or `<-` gets you the same 100 answers. 

**Unnecessary Information:** the `rnorm` functions is just stating that I want 100 observations with an average test score of 87.5 and I want the scores to vary around 87.5 by a standard deviation of 1.87. We'll get back to this later this semester. Additionally, the `set.seed()` is simply like a save point to make sure I always get the same 100 values for the observations.
```{r test score creation, eval = TRUE, echo = TRUE}

set.seed(08152021)
test_scores = rnorm(100, mean = 87.5, sd = 1.87)

set.seed(08152021)
test_scores2 <- rnorm(100, mean = 87.5, sd = 1.87)

test_scores
test_scores2

```

Okay, now that I have test scores, let's see what the sum of those scores are. I will also create an additional object that holds all the observed test scores. 
```{r number of observations, eval = TRUE, echo = TRUE}
sum_score = sum(test_scores)
sum_score

n = 100
n
```

Let's see what the average test score was. First, we can see what the average score is, but let's also save the average/mean test score for further analyses. 
```{r mean calculation by hand, eval = TRUE, echo = TRUE}
sum_score/n

xbar = sum_score/n
xbar
```

Now that we have the average test scores, let's see how far students were away from that average score. This is the `deviance` from the average test score. We'll come back to this for class. 
```{r, eval = TRUE, echo = TRUE}
dev = test_scores - xbar
dev
```

Totaling the deviance shows that no matter what, all students' scores should zero-out. I'll show this visually shortly. 
```{r, eval = TRUE, echo = TRUE}
total_dev = sum(dev)
total_dev
```

We'll also see when scores are "standardized" by squaring the test scores. This will be important for analyses later in this class. 
```{r, eval = TRUE, echo = TRUE}
dev_sq = (test_scores - xbar)^2
dev_sq
```

Let's also collect an object of the total deviance squared. 
```{r, eval = TRUE, echo = TRUE}
total_dev_sq = sum(dev_sq)
total_dev_sq
```

Similar to the mean, we need to get the total observations. We will talk about why we subtract 1 later in this class. 
```{r, eval = TRUE, echo = TRUE}
sample_obs = n - 1
sample_obs
```

To see how much the scores collectively vary from the average test score, we will get the variance. 
```{r, eval = TRUE, echo = TRUE}
variance = total_dev_sq/sample_obs
variance
```

To get the standard deviation, we conduct the square root of the variance. 
```{r, eval = TRUE, echo = TRUE}
standard_dev = sqrt(variance)
standard_dev
```

Thanks to `R`, we can use the built-in functions to see both the mean and standard deviations of our simulated test scores. 
```{r built in functions, eval = TRUE, echo = TRUE}
mean(test_scores)
sd(test_scores)
```

To get a little fancier, we can create a database/dataset/dataframe/tibble to visualize the deviance and squared deviance.
```{r creating a dataset, eval = TRUE, echo = TRUE}
library(tidyverse)

dataset <- tibble::tibble(scores = test_scores,
                          xbar_mean = xbar,
                          dev = dev,
                          dev_sq = dev_sq) %>% 
  rowid_to_column()

ggplot(data = dataset, aes(rowid, scores)) +
  geom_point(color = "dodgerblue") +
  geom_hline(yintercept = 87.15, color = "red", size = 1.25, alpha = .5) +
  theme_minimal()

ggplot(data = dataset, aes(rowid, dev)) +
  geom_col(position = "dodge2", fill = "dodgerblue", color = "white", alpha = .5) +
  geom_text(label = dataset$rowid) +
  geom_hline(yintercept = 0, color = "black", size = 1, alpha = .5) +
  theme_minimal()

ggplot(data = dataset, aes(rowid, dev_sq)) +
  geom_col(position = "dodge2", fill = "dodgerblue", color = "white", alpha = .5) +
  geom_text(label = dataset$rowid) +
  geom_hline(yintercept = 0, color = "black", size = 1, alpha = .5) +
  theme_minimal() 

```

We can now get some more information from this. From all of these graphs, students #90, and #24 appear to need some additional help here. We can also see that both students #53 and #55 may be able to help those students as they scored higher than all the other students.  

# Additional Guide to Downloading RStudio

## Why Use RStudio?

### **You don't need Rstudio, especially for use in this class.**

1. It is more user-friendly than R. It is an easier transition from SPSS.

2. It looks better.

3. Has keyboard shortcuts.

4. You can do your analyses, write your paper, and see your visuals at the same time.

5. It's also free.

6. You can create so much with it.

* [Websites](https://jonathan-a-pedroza.netlify.app/)
* [Data visualization dashboards](https://jpedroza1228.github.io/gradschool_dashboard/)
* [Create external servers for visualizations or analyses that take forever](https://shiny.rstudio.com/gallery/)

[Here is the download link](https://www.rstudio.com/products/rstudio/download/)
