---
title: "Problem Set 1"
author: "Jonathan A. Pedroza, PhD"
date: '2022-02-01'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, eval = TRUE}
library(tidyverse)

set.seed(082521)
normal = rnorm(10000, mean = 0, sd = 1)
neg_skew = rbeta(10000, shape1 = 40, shape2 = 2)
pos_skew = rbeta(10000, shape1 = 2, shape2 = 40)

skew <- tibble::tibble(normal, 
                       pos_skew,
                       neg_skew) %>% 
  rowid_to_column()

options(scipen = 999)
theme_set(theme_minimal())
```

1. What type of distribution is below?

```{r, echo = FALSE, eval = TRUE}
skew %>% 
  ggplot(aes(normal)) + 
  geom_histogram(fill = "gray70", color = "white", bins = 20, alpha = .8) +
  labs(x = '')
```

**Answer:** Normal distribution. It looks perfectly symmetrical and has the highest point at the middle with two even tails.

------------------------------------------------------------------------

2. What type of distribution is below?

```{r, echo = FALSE, eval = TRUE}
skew %>% 
  ggplot(aes(pos_skew)) + 
  geom_histogram(fill = "gray70", color = "white", bins = 20, alpha = .8) +
  labs(x = '')
```

**Answer:** Positively skewed distribution. It has a high frequency of low scores and a low frequency of higher scores. 

------------------------------------------------------------------------

3. What type of distribution is below?

```{r, echo = FALSE, eval = TRUE}
skew %>% 
  ggplot(aes(neg_skew)) + 
  geom_histogram(fill = "gray70", color = "white", bins = 20, alpha = .8) +
  labs(x = '')
```

**Answer:** Negatively skewed distribution. It has a high frequency of high scores and a low frequency of lower scores.

------------------------------------------------------------------------

4. Describe what a bimodal distribution is in your own words.

**Answer:** A bimodal distribution is a distribution with two modes, or two most frequent points. It can often show that there are two distinct groups in your distribution. 

------------------------------------------------------------------------

5. What is the term for the formula below?

$$= \frac{f}{N}$$

**Answer:** This is the formula for relative frequency. It is the frequency/count of a value divided by the number of observations. 

------------------------------------------------------------------------

6. Calculate the relative frequency for each category. The following data will be for numbers 6-8.**If using R/RStudio, you can copy+paste what is below and run that in R/RStudio to see the values.**

```{r, echo = TRUE, eval = TRUE}
categories <- c('d', 'e', 'a', 'e', 'e', 'e',
                'b', 'd', 'e', 'c', 'a', 'e',
                'a', 'd', 'e', 'd', 'e', 'c',
                'a', 'b', 'e', 'a', 'b', 'd',
                'd', 'd', 'd', 'e', 'e', 'e')
categories
table(categories)
```

**Answer:**
5/30 - A
3/30 - B
2/30 - C
8/30 - D
12/30 - E

------------------------------------------------------------------------

7. Get the **proportions** & **percentages** of each category.

**Answer:**

```{r}
5/30
3/30
2/30
8/30
12/30
```

.17 - A
.10 - B
.07 - C
.27 - D
.40 - E

```{r}
.17*100
.10*100
.07*100
.27*100
.40*100
```

17% A
10% B
07% C
27% D
40% E

------------------------------------------------------------------------

8. Calculate the cumulative frequency of categories D and E combined.

**Answer:**

.27 - D
.40 - E

```{r}
.27 + .40
```

D and E combined = .67 or 67% of the distribution

------------------------------------------------------------------------

9.Calculate the median for the following values. The following data will be for numbers 9 & 10. **If using R/RStudio, you can copy+paste what is below and run that in R/RStudio to see the values.**

```{r, echo = TRUE, eval = TRUE}
numbers <- c(3, 5, 7, 6, 4, 9, 1, 8, 4, 6, 6, 4, 5, 2, 3)
numbers
```

```{r}
sort(numbers)
```


**Answer:** 5 is the median, as it is the 8th value in order from smallest to largest.

------------------------------------------------------------------------

10. Calculate the mean for the following values.

```{r}
1+2+3+3+4+4+4+5+5+6+6+6+7+8+9
# sum is 73
```

```{r}
73/15
```

**Answer:** The mean/average value for the values provided was 4.87.

------------------------------------------------------------------------