---
title: "Problem Set 7"
author: "JP"
date: '2022-03-15'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You are interested in whether or not your intervention to reduce depression in college students worked. You collected baseline depression data then provided your intervention and tested all students after the intervention. Check to make sure your intervention significantly reduced depression levels from baseline to after the intervention.

```{r}
time_points <- data.frame(depress_t1 = c(18, 14, 16, 13, 19, 20),
                          depress_t2 = c(7, 4, 8, 10, 6, 4))
time_points
```

## What are our two hypotheses

H0:
H1:

## Get the means for both conditions

$$
\overline{X} = \sum X
$$

```{r}
mean(time_points$depress_t1)
mean(time_points$depress_t2)
```


## Get the df and sum of squares values for both conditions

$$
SS = \sum(X - \overline{X})^2
$$

df is 5 for both conditions

```{r}
# t1
(18 - 16.67)^2 +
(14 - 16.67)^2 +
(16 - 16.67)^2 +
(13 - 16.67)^2 +
(19 - 16.67)^2 +
(20 - 16.67)^2
```

```{r}
# t2
(7 - 6.5)^2 +
(4 - 6.5)^2 +
(8 - 6.5)^2 +
(10 - 6.5)^2 +
(6 - 6.5)^2 +
(4 - 6.5)^2
```

ss for t1 = 39.33
ss for t2 = 27.5

## What are the variance and standard deviation values for both conditions

$$
S^2 = \frac{SS}{df}
$$

$$
S = \sqrt{\frac{SS}{df}}
$$

```{r}
# t1 

39.33/5

# t2
27.5/5
```

t1 variance = 7.87
t2 variance = 5.5

```{r}
# t1
sqrt(7.87)

# t2
sqrt(5.5)
```

sd t1 = 2.81
sd t2 = 2.35

## Get the difference between each value

```{r}
18 - 7
14 - 4
16 - 8
13 - 10
19 - 6
20 - 4

time_points$diff <- time_points$depress_t1 - time_points$depress_t2
time_points
```

## Get the average difference between the two values

$$
\overline{D} = \sum D
$$

```{r}
(11 + 10 + 8 + 3 + 13 + 16)/6

mean(time_points$diff)
```

Average difference is 10.17

## Get the sum of squares for the differences

$$
SS = \sum(D - \overline{D})^2
$$

```{r}
(11 - 10.17)^2 + 
(10 - 10.17)^2 + 
(8 - 10.17)^2 + 
(3 - 10.17)^2 + 
(13 - 10.17)^2 + 
(16 - 10.17)^2
```

ss for difference is 98.83

## Get the variance and standard deviation of the differences

$$
S^2 = \frac{SS}{df}
$$

$$
S = \sqrt{\frac{SS}{df}}
$$

```{r}
98.83/5

sqrt(98.83/5)
```

variance of differences is 19.77
sd of differences is 4.45

## Get the standard error

$$
\sigma_{\overline{D}} = \frac{S_{D}}{\sqrt{N}}
$$

```{r}
sqrt(6)
4.45/2.45
```

se of differences is 1.82

## Get your t-obtained value

$$
t = \frac{\overline{D} - \mu_{D}}{\sigma_{\overline{D}}}
$$

```{r}
10.17 - 0

10.17/1.82
```

t-obtained is 5.59

## Is it significant?

2.571 critical value < t-obtained value 5.59

