---
title: "SPSS Assignments"
author: "JP"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, eval = TRUE, echo = FALSE}
library(tidyverse)

options(scipen = 999)
theme_set(theme_minimal())

data <- read_csv("https://raw.githubusercontent.com/jpedroza1228/dissertation/master/final_data/ca_county.csv") %>% 
  janitor::clean_names() %>% 
  select(-x1, -x) %>% 
  filter(release_year == 2018 &
           county_name != "california")
  
```

# What Each Assignment Will Look Like

## Create a research question and a hypothesis.

RQ: Is violent crime related to physical inactivity in California counties? 

H: More violent crime will be associated with more physical inactivity in California counties. 



## Conduct some descriptive statistics

Next, you will collect the mean and median for the outcome variable that we are interested in. For this assignment, the outcome will be

```{r, include = TRUE}
psych::describe(data$violent_crime, na.rm = TRUE)
psych::describe(data$physical_inactivity, na.rm = TRUE)

ggplot(data = data, aes(violent_crime)) + 
  geom_histogram(color = "white", fill = "dodgerblue", bins = 20)

ggplot(data = data, aes(physical_inactivity)) + 
  geom_histogram(color = "white", fill = "dodgerblue", bins = 20)
```


## Test Assumptions

```{r}
reg.diagnostics::linear_residuals(data = data, x = "violent_crime", y = physical_inactivity, se = FALSE, alpha = .5, size = 1.25,
                                  line_size = 1.25, color = "red", line_color = "dodgerblue")

reg.diagnostics::uni_bi_dist(data = data, x = violent_crime, y = physical_inactivity,
                             fill = "dodgerblue", alpha = .7, bins = 15, color = "gray70",
                             loess_color = "darkgreen", line_color = "red", se = FALSE, size = 1.25)
```

## Conduct inferential statistic for the tutorial

```{r, include = TRUE}
model <- lm(physical_inactivity ~ violent_crime, data = data)
summary(model)
lm.beta::lm.beta(model)
```

## Show some type of visualization to get the point across of the data

```{r}
ggplot(data = data, aes(violent_crime, physical_inactivity)) + 
  geom_point(aes(color = county_name)) + 
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme(legend.position = "none")

```

## Write up on your findings

There was a significant association between violent crime and physical inactivity (*b* = 0.00005, *SE* = 0.00002, *\beta* = .33, *p* = .01). For every .33 average increase in violent crime, there was a 1 point average increase in physical inactivity. 

```{r, include = TRUE}
broom::tidy(model)
jtools::summ(model)
```
