---
title: "SPSS Assignment 1"
author: "JP"
date: "8/16/2021"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, eval = TRUE, echo = FALSE}
library(tidyverse)

options(scipen = 999)
theme_set(theme_minimal())

jp <- rio::import(here::here("jp_thesis_1.sav")) %>% 
  janitor::clean_names() %>% 
  filter(ccc_class_standing == 3 |
           ccc_class_standing == 4)
  
```

# What Each Assignment Will Look Like

## Create a `null` and `alternative` hypothesis



## Rename `gender` into `sex`



## Conduct descriptive statistics for the following variables



- also include the appropriate descriptive visualization for each variable below

- `sex`
- `mtuas internet q1`
- `mtuas internet q2`
- `mtuas internet q3`
- `mtuas internet q4`
- `avg internet score variable`

## Conduct inferential statistic (independent samples t-test)

- screenshot the table with the independent samples t-test

## Tell me whether or not your finding is statistically significant

## find the t-critical value for a two-tailed test

## Show some type of visualization that shows the difference in the two groups on internet use

- examples of what it should look like are below

## Reject/Retain the Null Hypothesis?

## Write up on your findings

```{r, include = TRUE}
jp <- jp %>% 
  mutate(sex = factor(ccc_gender),
         internet_avg = (mtuas_internet_q1 + mtuas_internet_q2 + mtuas_internet_q3 + mtuas_internet_q4)/4)
```

```{r}
psych::describeBy(jp$internet_avg, jp$sex, na.rm = TRUE)

psych::describe(jp$internet_avg, na.rm = TRUE)
```

```{r}
jp %>% 
  mutate(sex = recode(sex, "1" = "Male",
                      "2" = "Female")) %>% 
  ggplot(aes(sex)) + 
  geom_bar(aes(color = sex, fill = sex)) + 
  scale_fill_manual(values = c("#d74122", "#387448")) +
  theme(legend.position = "none")

jp %>% 
  group_by(sex) %>% 
  count()

jp %>% 
  ggplot(aes(internet_avg)) + 
  geom_histogram(color = "white", fill = "#d74122", bins = 10)

internet_only <- jp %>% 
  dplyr::select(mtuas_internet_q1:mtuas_internet_q4)

map(internet_only, ~ggplot(data = internet_only, aes(.x)) + geom_histogram(color = "white", fill = "#387448", bins = 10))
```

```{r}
car::leveneTest(internet_avg ~ sex, data = jp)

t_test_finding <- t.test(internet_avg ~ sex, data = jp, var.equal = TRUE)
t_test_finding
t.test(internet_avg ~ sex, data = jp)
```

```{r}
jp %>% 
  mutate(sex = recode(sex, "1" = "Male",
                      "2" = "Female")) %>% 
  group_by(sex) %>% 
  summarize(mean_internet = mean(internet_avg),
            sd_internet = sd(internet_avg)) %>% 
  ungroup() %>% 
  ggplot(aes(sex, mean_internet)) + 
  geom_col(fill = "gray70", alpha = .5) +
  geom_errorbar(aes(ymin = mean_internet - sd_internet,
                    ymax = mean_internet + sd_internet,
                    color = sex), size = 1.25) +
  scale_color_manual(values = c("#d74122", "#387448")) +
  theme(legend.position = "none") +
  labs(x = "",
       y = "Average Time on the Internet")

jp %>% 
  mutate(sex = recode(sex, "1" = "Male",
                    "2" = "Female")) %>% 
  ggplot(aes(sex, internet_avg)) +
  geom_boxplot(aes(fill = sex), alpha = .5) +
  geom_jitter(color = "black", alpha = .5) + 
  scale_fill_manual(values = c("#d74122", "#387448")) +
  theme(legend.position = "none") +
  labs(x = "",
       y = "Average Time on the Internet")

```

```{r, include = TRUE}
broom::tidy(t_test_finding)
```

We conducted an independent samples t-test to examine if there was a difference in time spent on the internet between males and females. Our t-test showed that males (*M* = 5.95, *SD* = 1.95) spent less time on the internet compared to females (*M* = 6.61, *SD* = 2.17); *t*(239) = -2.26, *p* = .025. We are 95% certain that females have values of -1.23 and -0.08 less than males; indicating less time spent on the internet.  
