---
title: "SPSS Assignments"
author: "JP"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, eval = TRUE, echo = FALSE}
library(tidyverse)

options(scipen = 999)
theme_set(theme_minimal())

data <- read_csv("https://raw.githubusercontent.com/jpedroza1228/dissertation/master/final_data/ca_county.csv") %>% 
  janitor::clean_names() %>% 
  select(-x1, -x) %>% 
  filter(release_year == 2018 &
           county_name != "california")
  
```

# What Each Assignment Will Look Like

## Create a research question and a hypothesis.

RQ: Will there be a difference in air pollution between California counties with drinking water violations and those without drinking water violations?

H: California counties with drinking water violations will have significantly higher air pollution particulate matter levels than California counties without drinking water violations. 

## Recode Variables

For this assignment, you will first need to recode the variable `county_name` to create a new variable that separates Southern California and Northern California. This new variable will be called `south_north`. Below are the counties that will be in Southern California and Northern California.

Southern California:

- san_diego_county
- imperial_county
- riverside_county
- orange_county
- los_angeles_county
- san_bernardino_county
- kern_county
- ventura_county
- santa_barbara_county
- san_luis_obispo_county

Central California:

- inyo_county
- tulare_county
- kings_county
- monterey_county
- fresno_county
- san_benito_county
- mono_county
- madera_county
- mariposa_county
- merced_county
- santa_clara_county
- santa_cruz_county
- stanislaus_county
- tuolumne_county
- alpine_county
- calaveras_county
- amador_county
- san_joaquin_county

Northern California:

- sonoma_county
- colusa_county
- sutter_county
- yuba_county
- sierra_county
- nevada_county
- placer_county
- el_dorado_county
- yolo_county
- sacramento_county
- solano_county
- napa_county
- marin_county
- alameda_county
- contra_costa_county
- san_mateo_county
- san_francisco_county

Practically Oregon California: 

- del_norte_county
- siskiyou_county
- modoc_county
- humboldt_county
- trinity_county
- shasta_county
- lassen_county
- plumas_county
- tehama_county
- butte_county
- glenn_county
- lake_county
- mendocino_county


```{r, include = TRUE}

data <- data %>% 
  filter(drinking_water_violations == 0 |
           drinking_water_violations == 1) %>% 
  mutate(drinking_water_violations = as.factor(drinking_water_violations),
         high_low_pop = case_when(population < 183254 ~ 0,
                                  population >= 183254 ~ 1),
         high_low_pop = as.factor(high_low_pop))
```

## Conduct some descriptive statistics

Next, you will collect the mean and median for the outcome variable that we are interested in. For this assignment, the outcome will be

```{r, include = TRUE}
psych::describe(data$median_household_income, na.rm = TRUE)

psych::describeBy(data$median_household_income, data$drinking_water_violations, na.rm = TRUE)
psych::describeBy(data$median_household_income, data$high_low_pop, na.rm = TRUE)
```

```{r}
aov_table <- data %>% group_by(high_low_pop, drinking_water_violations) %>% 
  summarize(count = n(),
            mean_income = mean(median_household_income),
            sd_income = sd(median_household_income)) %>% 
  ungroup()

kableExtra::kbl(aov_table) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                position = "center")
```

## Test Assumptions

```{r}

```


## Conduct inferential statistic for the tutorial

```{r, include = TRUE}
library(car)

two_way <- aov(median_household_income ~ high_low_pop*drinking_water_violations, data = data)

summary(two_way)
Anova(two_way, type = "III")
TukeyHSD(two_way, conf.level = .95)
```

## Show some type of visualization to get the point across of the data

```{r}
data %>% 
  group_by(drinking_water_violations, high_low_pop) %>% 
  summarize(avg_income = mean(median_household_income),
            sd_income = sd(median_household_income)) %>% 
  ungroup() %>% 
  ggplot(aes(drinking_water_violations, avg_income)) + 
  geom_col(aes(fill = drinking_water_violations), position = "dodge2", alpha = .3) +
  geom_errorbar(aes(ymin = avg_income - sd_income,
                    ymax = avg_income + sd_income), size = 1.25) +
  facet_wrap(~high_low_pop)

data %>% 
  ggplot(aes(drinking_water_violations, median_household_income)) +
  geom_boxplot(aes(fill = drinking_water_violations), alpha = .5) +
    geom_jitter(color = "red", alpha = .5) +
  facet_wrap(~high_low_pop)

```

## Write up on your findings

California counties with drinking water violations (*M* = 8.84, *SD* = 2.84) did not have significantly higher pollution particulate matter levels compared to California counties without drinking water violations (*M* = 7.93, *SD* = 1.16); *t*(56) = -1.12, *p* = .266. 

```{r, include = TRUE}
broom::tidy(two_way)
```
