---
title: "Final Exam"
author: "Jonathan A. Pedroza, PhD"
date: "11/29/2021"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
editor_options: 
  chunk_output_type: console
---

```{css my-header-colors, echo = FALSE}
.page-header {
  color: #FAE9A0;
  text-align: center;
  background-color: #005C55;
  background-image: linear-gradient(120deg, #7DA37B, #005C55);
}

a {
  color: #F1C40F; #color of linked inline code
  font-weight: bold;
  text-decoration: none;
}
  a:hover {
    text-decoration: underline;
    }

.main-content {
  word-wrap: break-word; }
  .main-content :first-child {
    margin-top: 0; }
  @media screen and (min-width: 64em) {
    .main-content {
      max-width: 64rem;
      padding: 2rem 6rem;
      margin: 0 auto;
      font-size: 1.1rem; } }
  @media screen and (min-width: 42em) and (max-width: 64em) {
    .main-content {
      padding: 2rem 4rem;
      font-size: 1.1rem; } }
  @media screen and (max-width: 42em) {
    .main-content {
      padding: 2rem 1rem;
      font-size: 1rem; } }
  .main-content img {
    max-width: 100%; }
  .main-content h1,
  .main-content h2,
  .main-content h3,
  .main-content h4,
  .main-content h5,
  .main-content h6 {
    margin-top: 2rem;
    margin-bottom: 1rem;
    font-weight: normal;
    color: #3F7F68; } # header color
  .main-content p {
    margin-bottom: 1em; }
  .main-content code {
    padding: 2px 4px;
    font-family: Consolas, "Liberation Mono", Menlo, Courier, monospace;
    font-size: 0.9rem;
    color: #005C55; #package color/whatever is in `` brackets
    background-color: #7DA37B; #background color of ``brackets
    border-radius: 0.3rem; }
  .main-content pre {
    padding: 0.8rem;
    margin-top: 0;
    margin-bottom: 1rem;
    font: 1rem Consolas, "Liberation Mono", Menlo, Courier, monospace;
    color: #567482;
    word-wrap: normal;
    background-color: #f3f6fa; #background color of code chunks
    border: solid 1px #005C55; #this is the border of the r code chunks
    border-radius: 0.3rem; }
    .main-content pre > code {
      padding: 0;
      margin: 0;
      font-size: 0.9rem;
      color: #5D6D7E;
      word-break: normal;
      white-space: pre;
      background: transparent;
      border: 0; }
  .main-content .highlight {
    margin-bottom: 1rem; }
    .main-content .highlight pre {
      margin-bottom: 0;
      word-break: normal; }
  .main-content .highlight pre,
  .main-content pre {
    padding: 0.8rem;
    overflow: auto;
    font-size: 0.9rem;
    line-height: 1.45;
    border-radius: 0.3rem;
    -webkit-overflow-scrolling: touch; }
  .main-content pre code,
  .main-content pre tt {
    display: inline;
    max-width: initial;
    padding: 0;
    margin: 0;
    overflow: initial;
    line-height: inherit;
    word-wrap: normal;
    background-color: transparent;
    border: 0; }
    .main-content pre code:before, .main-content pre code:after,
    .main-content pre tt:before,
    .main-content pre tt:after {
      content: normal; }
  .main-content ul,
  .main-content ol {
    margin-top: 0; }
  .main-content blockquote {
    padding: 0 1rem;
    margin-left: 0;
    color: #819198;
    border-left: 0.3rem solid #dce6f0; }
    .main-content blockquote > :first-child {
      margin-top: 0; }
    .main-content blockquote > :last-child {
      margin-bottom: 0; }
  .main-content table {
    display: block;
    width: 100%;
    overflow: auto;
    word-break: normal;
    word-break: keep-all;
    -webkit-overflow-scrolling: touch; }
    .main-content table th {
      font-weight: bold; }
    .main-content table th,
    .main-content table td {
      padding: 0.5rem 1rem;
      border: 1px solid #e9ebec; }
  .main-content dl {
    padding: 0; }
    .main-content dl dt {
      padding: 0;
      margin-top: 1rem;
      font-size: 1rem;
      font-weight: bold; }
    .main-content dl dd {
      padding: 0;
      margin-bottom: 1rem; }
  .main-content hr {
    height: 2px;
    padding: 0;
    margin: 1rem 0;
    background-color: #5E9171; #line to break up sections
    border: 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

2 points - What assumption is examined when you need to have a linear relationship between your IV and your DV?

A. Independence of residuals

B. Normality

C. Homoscedasticity

**D. Linearity**

2 points - What test accounts for the violation in independence of residuals assumption?

**A. Repeated Measures ANOVA**

B. Linear Regression

C. Correlation

D. Two-Way ANOVA

2 points - The standardized regression coefficient will result in the same value as a Pearson's correlation coefficient in a multiple linear regression

A. True

**B. False**

2 points - In a simple linear regression, a beta/standardized regression coefficient of .34 would be a __________ effect size.

A. Small

**B. Medium**

C. Large

D. Super-sized

2 points - What can be interpreted from a standardized regression coefficient in a multiple linear regression?

A. The effect size of the relationship

**B. The strength of the relationship**

C. The amount of variance accounted for in Y by X

D. The amount of change in Y for a one-unit increase in X

2 points - What is the difference between correlation and regression?

A. Correlation is scale dependent

B. Both focus on the strength of a linear relationship

C. Nothing, they are the same test

**D. Regression focuses on the direction of the relationship and can often include both categorical and continuous independent variables**

2 points - In the regression model, which of the values is the residual/error variance?

y = b0 + b1X + e

A. b0

B. b1

C. the y-intercept

**D. e**

E. y


**Not sure on the amount of points** - Match the distinction to the correct inferential statistic 

One-way ANOVA - one categorical independent variable and one continuous dependent variable

ANCOVA - two categorical independent variables, one continuous dependent variable, but no interaction between the two independent variables

Linear Regression - either one or many categorical or continuous independent variables/predictors and one continuous dependent variable

Two-way ANOVA - two categorical independent variables, one continuous dependent variable, and an interaction between the two independent variables

MANOVA - Who cares. Its not very useful.

Mixed-design ANOVA - two categorical variables, with one being a between-subjects design and the other being a within-subjects design

Bivariate Correlation - two continuous variables, none of which is explicitly stated to be an independent or dependent variable

Repeated Measures ANOVA - one or more within-subjects categorical variables with multiple conditions/time points; has a potential interaction between the variables (if more than one variable included)


2 points - In the following ANOVA, how many variables are included in the ANOVA? 

3 x 4 x 2 ANOVA

**A. 3**

B. 4

C. 2

D. 1

**Fill in the Blank Answer**

2 points - r = .35 is a stronger correlation than -.67

A. True

**B. False**

2 points - Sphericity is an issue only when you have two time points.

A. True

**B. False**

2 points - What assumption is when you want to make sure that your residuals are independent of one another? 

**A. Independence**

B. Linearity

C. Homogeneity of Variance

D. Normality

2 points - What is the difference between between-subjects ANOVA designs and within-subjects ANOVA designs?

A. In within-subjects designs, participants are separated into different levels/conditions

B. In between-subjects designs, every participant gets each condition

C. Between-subjects ANOVA designs are useful when you are looking at the same measure over time

**D. None of these are correct**

2 points - What test is used for longitudinal designs?

A. One-way ANOVA

B. Linear Regression

C. Correlation

**D. Repeated Measures ANOVA**

2 points - What is a potential problematic Cook's distance in your data?

**A. 3.45** 

B. 0.6

C. 0.3

D. 0.1

2 points - How does a repeated measures/mixed design ANOVA account for the lack of independence?

A. Collapsing

**B. Partitioning**

C. Interactions

D. It does not account for a lack of independence

2 points - Which choices below are instances that a repeated measures/mixed design ANOVA would not be the appropriate analysis for longitudinal data? (multiple answers)

**A. Missing data in the outcome**

**B. Time is treated as a continuous variable**

**C. Covariates vary by each time point**

**D. The outcome is dichotomous/categorical**

2 points - In a two-way ANOVA, what would be the order in which we look at our findings?

**A. Interaction, main effects, multiple comparisons and simple effects**

B. Main effects, multiple comparisons and simple effects, interaction

C. Main effects, interaction, multiple comparisons and simple effects

D. Just examine the multiple comparisons and simple effects

2 points - A main effect is when a relationship between an IV and DV is dependent on the other IV values in a two-way ANOVA

A. True

**B. False**

2 points - You can have a significant interaction but then nonsignificant simple effects/multiple comparisons

**A. True**

B. False

2 points - In a two-way ANOVA, if your interaction is nonsignificant you should stop looking at your findings.

A. True

**B. False**

2 points - Which is a relationship that when one variable increases the other variable increases?

A. no relationship

B. Negative relationship

C. Inverse relationship

**D. Positive relationship**

2 points - A cross-production deviation is...

A. Covariance

**B. The multiplied deviations of two variables**

C. The average combined deviations

D. Another term for variance

2 points - Covariance is a standardized measurement of the relationship between two continuous variables

A. True

**B. False**

2 points - You are interested in the relationship between time spent studying (x) and exam scores (y). Below is the raw data and the deviations from the mean. The **mean for time spent studying is 7**, while the **average exam score was 86.8**. Calculate the covariance between the two variables and choose the correct answer below.

```{r}
test_scores <- data.frame(x = c(5, 4, 7, 9, 10),
                          y = c(80, 94, 75, 100, 85))

test_scores %>% 
  mutate(x_deviations = x - mean(x),
         y_deviations = y - mean(y))
```

```{r, include = FALSE}
cov(test_scores$x, test_scores$y)
```

**A. 3.25**

B. 1.96

C. 3.14

D. 7.45

2 points - Can you use a covariance value as an effect size?

A. Yes

**B. No**

2 points - A standardized value is in ____ units?

**A. Standard deviation**

B. Variance

C. Correlation

D. Covariance

2 points - Which is an impossible correlation coefficient?

A. 0

B. -1 

C. -.9

**D. 1.1**

1 point - For this problem you are interested in the cells used for an interaction between race/ethnicity and education level. The education levels are high school graduates, college graduates, and trade school graduates. The races/ethnicities are Black/African American, Latina/o/x, and White. For the data below, we are interested in the average depression score of Black/African American college graduates. What numbers from the cell of interest would we use to get the average score? 

```{r}
race_class <- data.frame(education = c("hs_grad", "hs_grad", "hs_grad",
                                       "college_grad", "college_grad", "college_grad",
                                       "trade_school_grad", "trade_school_grad", "trade_school_grad"),
                         black_african_american = c(5, 7, 9, 8, 4, 2, 1, 9, 2),
                         latino = c(5, 4, 8, 7, 6, 2, 4, 6, 7),
                         white = c(0, 2, 1, 6, 7, 5, 8, 9, 1))

race_class
```

A. 6, 7, 5

**B. 8, 4, 2**

C. 1, 9, 2

D. 5, 7, 9, 8, 4, 2, 1, 9, 2

1 point - What is the average for our Black/African American college graduates?

Note. Please round your final answer to two decimal points. An example would be: 7.54, not 7.536

```{r}
(8 + 4 + 2)/3
```

**Answer: 4.67**

2 points - For the table above, What does the unstandardized regression coefficient for the relationship between BMI category 2/overweight (bmi_cat=2.0) compared to the reference group(BMI category 1/normal & underweight)?

A. That overweight participants have a 5.53 higher BMI on average when compared to normal and underweight participants.

**B. That overweight participants have a 5.53 higher BMI when compared to normal and underweight participants on average when holding an obese BMI, exercise levels, tv viewing levels, and phone call levels constant**

C. That normal and underweight participants have a BMI that is 5.53 points higher than overweight participants on average.

D. That normal and underweight participants have on average a 12.27 lower BMI than obese participants.

2 points - For the table above, what findings are statistically significant?

**A. BMI category = 2(overweight), BMI category = 3(obese), tv viewing**

B. exercise and phone call

C. tv only

D. all the relationships are statistically significant

2 points - In the table above, all the predictors are included in the model together. What type of linear regression was conducted?

A. Sequential

**B. Simultaneous/Standard**

C. Hierarchical

D. Automated

2 points - In the screenshot, what assumption is being violated in this visualization of the residuals?

**A. Independence**

B. Homogeneity

C. Outliers

D. Normality

2 points - An actual score is...

**A. a score that a participant answered in the survey/experiment**

B. a score that is on the regression line

C. a residual

D. the score they thought of in their head

2 points - If you have multicollinearity between two predictors in your regression model, it means that you have...

A. your model is simply wrong

B. to increase your sample size

C. too many predictors in your model

**D. predictors that are too similar and you should probably remove one from the model**

2 points - Partial-correlation is used to examine the unique relationship between an IV and a DV while controlling for one or more other variables.

**A. True**

B. False

2 points - What is the difference between Spearman's correlation coefficient and Pearson's correlation coefficient?

A. Pearson's correlation coefficient is a nonparametric test

B. Spearman's correlation coefficient is a parametric test

**C. Spearman's correlation coefficient examines the relationship between variables that represent ranked data**

D. Pearson's correlation coefficient is written out with a rho greek symbol

2 points - Eta squared and R squared are both measures of the amount of variation accounted for in your outcome. The only difference is eta squared is often used for ANOVA and R squared is more commonly used with regression.

**A. True**

B. False

2 points - What is the distance between a participant's actual score and the regression line called?

**A. residual**

B. error variance

C. intercept

D. regression

2 points - What is k in a regression?

A. Number of conditions/levels

**B. Number of predictors/IVs**

C. a measure of model fit

D. There is no k in regression

2 points - What is used to compare predictors in the strength of their relationship/association with the outcome in your regression?

A. Studentized residual

B. The intercept

C. Unstandardized regression coefficient

**D. Standardized regression coefficient/beta**

2 points - A unstandardized residual is how far a participant's score is from the regression model in the original units

**A. True**

B. False

2 points - What is needed to compare two regression models?

A. One should be a regression and the other should be an ANOVA

B. They must have the same predictors but different outcomes

C. They must have different predictors

**D. They must be nested**

1 point - 

Below is some data examining four different time points where the researchers are interested in stats knowledge on a 1-5 scale before taking a stats class, during the class, after the class, and a year after taking the class. Answer the following questions about these students.

```{r}
knowledge <- data.frame(pre = c(1, 1, 2, 1),
                        during = c(2, 2, 2, 4),
                        post = c(4, 5, 3, 5),
                        follow_up = c(3, 4, 2, 1))
knowledge
```

```{r, include = FALSE}

ss_total <- function(total_sq, total, n){
  total_sq - (((total)^2)/n)
}

ss_between <- function(pre_val, dur_val, post_val, follow_val, n_per_group, total, n){
 (((pre_val)^2)/n_per_group + ((dur_val)^2)/n_per_group + ((post_val)^2)/n_per_group + ((follow_val)^2)/n_per_group) - (((total)^2)/n)
}

ss_subject <- function(p1, p2, p3, p4, k, total, n){
  (((p1^2 + p2^2 + p3^2 + p4^2))/k) - (((total)^2)/n)
}

knowledge %>% 
  summarize(total_sum = (1 + 1 + 2 + 1 +
                           2 + 2 + 2 + 4 + 
                           4 + 5 + 3 + 5 + 
                           3 + 4 + 2 + 1),
            pre_mean = mean(pre),
            dur_mean = mean(during),
            post_mean = mean(post),
            fol_mean = mean(follow_up),
            pre_sum = 1 + 1 + 2 + 1,
            dur_sum = 2 + 2 + 2 + 4,
            post_sum = 4 + 5 + 3 + 5,
            fol_sum = 3 + 4 + 2 + 1,
            p1_sum = 1 + 2 + 4 + 3,
            p2_sum = 1 + 2 + 5 + 4,
            p3_sum = 2 + 2 + 3 + 2,
            p4_sum = 1 + 4 + 5 + 1,
            pre_sq = 1^2 + 1^2 + 2^2 + 1^2,
            dur_sq = 2^2 + 2^2 + 2^2 + 4^2,
            post_sq = 4^2 + 5^2 + 3^2 + 5^2,
            fol_sq = 3^2 + 4^2 + 2^2 + 1^2,
            total_sq = pre_sq + dur_sq + post_sq + fol_sq,
            k = 4,
            n_per = 4,
            total_n = 16,
            ss_t = ss_total(total_sq = total_sq, total = total_sum, n = total_n),
            ss_b = ss_between(pre_val = pre_sum, dur_val = dur_sum, post_val = post_sum, follow_val = fol_sum, n_per_group = 4, total = total_sum, n = total_n),
            ss_s = ss_subject(p1 = p1_sum, p2 = p2_sum, p3 = p3_sum, p4 = p4_sum, k = k, total = total_sum, n = total_n),
            ss_e = ss_t - ss_b - ss_s,
            df_b = k - 1,
            df_e = (k - 1)*(4 - 1),
            ms_b = ss_b/df_b,
            ms_e = ss_e/df_e,
            f_obt = ms_b/ms_e,
            f_crit = 3.8625
            )

```

2 points - What test will you be running

A. Correlation

**B. Repeated-measures ANOVA**

C. Mixed-design ANOVA

D. Two-way ANOVA

1 point - Get the sum of squares total

1 point - Get the sum of squares between groups

1 point - Get the sum of squares for the participants

1 point - What is the F critical value (check the F table; you should have two degrees of freedom for both within/error and between)

1 point - What is your F obtained value

2 points - Is your test statistically significant?

A. Yes

B. No




