---
title: "Homework Problem Set 3"
author: "JP"
date: "9/8/2021"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Calculate the standard deviation using either formula discussed in class.

```{r, echo = TRUE, eval = TRUE}
set.seed(090821)

test_scores = c(81, 78, 91, 88, 95, 97, 99, 79, 73, 74)

N = 10

```

```{r}
test_scores
```

```{r}
81^2 + 78^2 + 91^2 + 88^2 + 95^2 + 97^2 + 99^2 + 79^2 + 73^2 + 74^2
# 73951 is sum of squared Xs

81 + 78 + 91 + 88 + 95 + 97 + 99 + 79 + 73 + 74
# sum of Xs is 855

855^2
# squared sum of Xs is 731025

731025/10
# 73102.5 is the right side of the numerator

73951.0 - 73102.5
# numerator is 848.5

10 - 1
# denominator is 9

848.5/9
# variance is 94.28

sqrt(94.27778)
# sd is 9.71

sd(test_scores)

```

**Answer** 

# SPSS Practice

**For SPSS Answers, please see the syntax file in the `Answer Key` folder.**

1. Gather descriptive statistics using the `descriptives` tab in SPSS for the following variables. (Get a screenshot of this table and put it on this word document.)
  - ffq_sausage
  - ffq_pizza_portion
  - dass_stress_q1
  - dass_depress_q7
  
**Picture Answer**

2. Get the z-transformed variables from problem 1. (Screenshot the variables in either the `Data View` or the `Variable View`.)

**Picture Answer**

3. Get the descriptive statistics and the histograms for the z-transformed variables from problem 2. Use the `frequencies` tab for the descriptive statistics (Screenshot the table and one variable histogram and put it on this word document.)

**Picture Answer**
  
4. Also, get either the bar graph of each variable from the previous problem using the `frequencies` tab. I do not need the descriptive statistics for this section. (Get a screenshot of one of the variables and put it on this word document.)

**Picture Answer**

5. Reverse score the `ffq_sausage` variable. To show that you did it correctly, get the frequencies of both the original variable and the new variable you created. (Get a screenshot the frequency tables and put it on this word document.)

**Picture Answer**

6. Dummy code the `ccc_class_standing` variable. Rename the variables in the `Variable View` to the names that correspond to the numbers. For example, 1 = Freshmen, 2 = Sophomore, 3 = Junior, 4 = Senior, 5 = Senior_plus. (Get a screenshot of the variables in the data or variable view and put it on this word document.)

**Picture Answer**

7. Create a composite score of the variables that start with `cpaq`. There should be 4 questions. Then get the descriptive statistics and histogram for the new variable you created. (Screenshot the table and the histogram and put it on this word document.)

**Picture Answer**

