---
title: "PSY 3307"
subtitle: "Analysis of Variance (ANOVA)"
author: "Jonathan A. Pedroza PhD"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 99999)

library(tidyverse)
```

# Agenda

- Post-hoc tests
  - Tukey HSD
  - Bonferroni Correction
- Effect sizes
- Proportion of Variance Accounted For
- What is an ANCOVA

  
---

# Post-hoc Tests

- **Tukey HSD (Honestly Significant Difference)** is a post hoc test for ANOVA to compare mean for each level in a factor
  - used only when the `ns` in each level of the factor are equal
  
- it computes the the minimum difference between two means that is required for them to be significantly different

- Bonferroni Correction
  - based on your alpha, you get that value and divide by the number of tests you need to conduct

---

# From last class

$$\left[\begin{array}{cc} 
Source\\
Between\\
Within\\
Total
\end{array}\right] 
\left[\begin{array}{cc} 
Sum\;of\;Squares\\
63.33\\
84\\
147.33\\
\end{array}\right] 
\left[\begin{array}{cc} 
df\\
2\\
12\\
14\\
\end{array}\right] 
\left[\begin{array}{cc} 
Mean Square\\
31.67\\
7\\
\\
\end{array}\right] 
\left[\begin{array}{cc} 
F-ratio\\
4.52\\
\\
\\
\end{array}\right]$$

---

# Tukey Steps

1. Find `q_{k}`, which are values of studentized range statistic
  - We'll come back to studentized values
  - locate the `k`
  - find the `df_{wm}`

2. Compute the HSD
  
$$HSD = (q_{k})(\sqrt{\frac{MS_{wn}}{n}})$$

---

3. Determine the differences between each pair of means

4. Compare each difference between means to the Tukey HSD value
  - if the absolute difference between two means is greater than the HSD then these differences are significantly different
    
    + similar to a independent samples t-test with a significant t-obtained value

---

# Example from last slides

MS within = 7
n = 5

For `q_{k}`, we need k and `df_{wn}`
df_wn = 12
k = 3
q_k = 3.77

```{r}
3.77*(sqrt(7/5))
```

---

Our Tukey HSD is `4.46`

So now we compare the means of our three groups/levels

```{r}
8 - 6 # easy - medium
8 - 3 # easy - hard
6 - 3 # medium - hard
```

2 < 4.46
5 > 4.46
3 < 4.46

So only the difference between easy and the hard groups was a statistically significant finding at an alpha of .05

---

# Effect Size

- cohen's d only really used for two-sample designs (t-tests)

- rather, for ANOVA and regression, we will rely on **Proportion of variance accounted for**
  - there are better effect sizes (most would argue that proportion of variance accounted for in regression models is not an effect size)

- greater the eta squared, the more that our IV accounts for the DV

- eta squared is the same as R squared

- **eta squared** is the proportion of variance in the DV that is accounted for with the change in levels in the factor/IV

---

# Formula

$$\eta^2 = \frac{SS_{bn}}{SS_{total}}$$

- SS between is the differences between all the scores in the experiment
  - it is also what we know our model is actually looking at
  
- because it is proportion, we can easily interpret it as the percentage that the change in groups/levels accounts for the DV score difference

$$ANOVA:\;eta^{squared} = \eta^2$$

$$Regression:\;R^2$$

---

# ANCOVA

- conceptually the same thing as ANOVA but it includes covariates/control variables

- ANCOVA = Analysis of Covariance

- The same thing as multiple regression

- You are still interested in differences between levels on your DV values, but now you include additional variables to help account for some of the variance in your DV

---


$$\left[\begin{array}{cc} 
Source\\
Between\\
Within\\
Total
\end{array}\right] 
\left[\begin{array}{cc} 
Sum\;of\;Squares\\
SS_{bn}\\
SS_{wn}\\
SS_{total}\\
\end{array}\right] 
\left[\begin{array}{cc} 
df\\
df_{bn}\\
df_{wn}\\
df_{total}\\
\end{array}\right] 
\left[\begin{array}{cc} 
Mean Square\\
MS_{bn}\\
MS_{wn}\\
\\
\end{array}\right] 
\left[\begin{array}{cc} 
F-ratio\\
F_{obt}\\
\\
\\
\end{array}\right]$$

---

# Context

You are trying out four different methods for helping participants with their depression. You decide to give out: medication only, individual therapy, group, and medication and individual therapy. Below are each level's depression levels at the end of the experiment. You are interested in a difference between the four groups. 

```{r}
data <- data.frame(meds = c(4, 5, 4, 6, 5),
                   ind_therapy = c(5, 4, 3, 5, 5),
                   group = c(8, 8, 6, 9, 10),
                   med_ind = c(3, 1, 3, 2, 1))

data
```

---

# Hypotheses

H0: There will be no differences between the therapeutic approaches

H1: There will be differences between the therapeutic approaches

Even better,

H1: The group that receives medication and individual therapy will have significantly better or worse depression rates than every other group.

---

# Step-by-step

What is the k?

```{r, include = FALSE}
k = 4
k
```

---

What is the number of participants per level?

```{r, include = FALSE}
meds_n = 5
ind_n = 5
group_n = 5
both = 5

meds_n
ind_n
group_n
both
```

---

What is the total number of participants?

```{r, include = FALSE}
total_n = meds_n + ind_n + group_n + both
total_n
```

---

What is the sum of each group?

```{r, include = FALSE}
meds_sum = sum(data$meds)
ind_sum = sum(data$ind_therapy)
group_sum = sum(data$group)
med_ind_sum = sum(data$med_ind)

meds_sum
ind_sum
group_sum
med_ind_sum
```

---

What is the mean of each group?

```{r, include = FALSE}
meds_mean = mean(data$meds)
ind_mean = mean(data$ind_therapy)
group_mean = mean(data$group)
med_ind_mean = mean(data$med_ind)

meds_mean
ind_mean
group_mean
med_ind_mean
```

---

What is the sum of squared Xs for reach group?

```{r, include = FALSE}
meds_sum2 = 4^2 + 5^2 + 4^2 + 6^2 + 5^2
ind_sum2 = 5^2 + 4^2 + 3^2 + 5^2 + 5^2
group_sum2 = 8^2 + 8^2 + 6^2 + 9^2 + 10^2
med_ind_sum2 = 3^2 + 1^2 + 3^2 + 2^2 + 1^2

meds_sum2
ind_sum2
group_sum2
med_ind_sum2

```

---

What is the total sum?

```{r, include = FALSE}
total_sum = meds_sum + ind_sum + group_sum + med_ind_sum
total_sum
```

---

What is the total sum of all squared values?

```{r, include = FALSE}
total_sum2 = meds_sum2 + ind_sum2 + group_sum2 + med_ind_sum2
total_sum2
```

---

What is the sum of squares total?

$$SS_{total} = \Sigma(X_{ij} - \overline{X})^2$$

$$SS_{total} = \Sigma X^2_{total} - \frac{(\Sigma X_{total})^2}{N}$$

```{r, include = FALSE}
587 - ((97^2)/20)
# 116.55
```

---

What is the sum of squares between/treatment?

$$SS_{treat} = n \Sigma(\overline{X_{j}} - \overline{X})^2$$

$$SS_{bn} = \Sigma(\frac{(\Sigma X\;in\;each\;column)^2}{n\;in\;each\;column}) - \frac{(\Sigma X_{total})^2}{N}$$

```{r, include = FALSE}
24^2
576/5

22^2
484/5

41^2
1681/5
 
10^2
100/5

97^2
9409/20

(115.2 + 96.8 + 336.2 + 20) - 470.45
# 97.75
```

---

What is the sum of squares within/error?

$$SS_{wn} = SS_{total} - SS_{bn}$$


$$SS_{error} = SS_{total} - SS_{treat}$$

```{r, include = FALSE}
116.55 - 97.75
# 18.8
```

---

What are the degrees of freedom between?

$$df_{bn} = k - 1$$

```{r, include = FALSE}
4 - 1
# 3
```

---

What are the degrees of freedom within? (N - k)

```{r, include = FALSE}
20 - 4
# 16
```

---

What are the degrees of freedom total? (N - 1)

```{r, include = FALSE}
20 - 1
# 19
```

---

What is the mean squares between value?

$$MS_{bn} = \frac{SS_{bn}}{df_{bn}}$$

```{r, include = FALSE}
97.75/3
# ms between = 32.58
```


---

What is the mean squares within value?

$$MS_{wn} = \frac{SS_{wn}}{df_{wn}}$$

```{r, include = FALSE}
18.8/16
# 1.18
```

---

What is the F-obtained value?

$$F_{obt} = \frac{MS_{bn}}{MS_{wn}}$$

```{r, include = FALSE}
32.58/1.18
# f value is 27.61
```

---

Calculate the Tukey HSD

$$HSD = (q_{k})(\sqrt{\frac{MS_{wn}}{n}})$$

```{r, include = FALSE}
# 16 df_wn
# ms_wn = 1.18
# n = 5
# k = 4
# qk = 4.046

4.046*(sqrt(1.18/5))
# HSD = 1.97
```

---

```{r, include = FALSE}
# meds_mean is 4.8
# ind_mean is 4.4
# group_mean is 8.2
# med_ind_mean 2

4.8 - 4.4 #ns
8.2 - 4.8 #s
8.2 - 4.4 #s
8.2 - 2 #s
4.8 - 2 #s
4.4 - 2 #s

```

---

# Effect Size

$$\eta^2 = \frac{SS_{bn}}{SS_{total}}$$

```{r, include = FALSE}
eta2 = 97.75/116.55
eta2
```

---

What can we say?

- There is no significant difference between medications and individual therapy in depression scores
- There are lower depression scores in the group with individual therapy compared to group therapy
- There are lower depression scores in the group with medications compared to group therapy
- There are lower depression scores in the group with individual therapy and medications compared to group therapy
- There are lower depression scores in the group with individual therapy and medications compared to medications alone
- There are lower depression scores in the group with individual therapy and medications compared to individual therapy alone

- Effect size: The differences in therapeutic styles accounted for 83.87% of the variance in depression scores in our sample.

---

```{r}
data2 <- data.frame(meds = c(4, 5, 4, 6, 5),
                   ind_therapy = c(5, 4, 3, 5, 5),
                   group = c(8, 8, 6, 9, 10),
                   med_ind = c(3, 1, 3, 2, 1),
                   control = c(10, 9, 8, 7, 9))

data2
```

---

# Step-by-step

What is the k?

```{r, include = FALSE}
k = 5
k
```

---

What is the number of participants per level?

```{r, include = FALSE}
meds_n = 5
ind_n = 5
group_n = 5
both = 5
control = 5

meds_n
ind_n
group_n
both
control
```

---

What is the total number of participants?

```{r, include = FALSE}
total_n = meds_n + ind_n + group_n + both + control
total_n
```

---

What is the sum of each group?

```{r, include = FALSE}
meds_sum = sum(data2$meds)
ind_sum = sum(data2$ind_therapy)
group_sum = sum(data2$group)
med_ind_sum = sum(data2$med_ind)
control_sum = sum(data2$control)

meds_sum
ind_sum
group_sum
med_ind_sum
control_sum
```

---

What is the mean of each group?

```{r, include = FALSE}
meds_mean = mean(data2$meds)
ind_mean = mean(data2$ind_therapy)
group_mean = mean(data2$group)
med_ind_mean = mean(data2$med_ind)
control_mean = mean(data2$control)

meds_mean
ind_mean
group_mean
med_ind_mean
control_mean
```

---

What is the sum of squared Xs for reach group?

```{r, include = FALSE}
meds_sum2 = 4^2 + 5^2 + 4^2 + 6^2 + 5^2
ind_sum2 = 5^2 + 4^2 + 3^2 + 5^2 + 5^2
group_sum2 = 8^2 + 8^2 + 6^2 + 9^2 + 10^2
med_ind_sum2 = 3^2 + 1^2 + 3^2 + 2^2 + 1^2
control_sum2 = 10^2 + 9^2 + 8^2 + 7^2 + 9^2 

meds_sum2
ind_sum2
group_sum2
med_ind_sum2
control_sum2
```

---

What is the total sum?

```{r, include = FALSE}
total_sum = meds_sum + ind_sum + group_sum + med_ind_sum + control_sum
total_sum
```

---

What is the total sum of all squared values?

```{r, include = FALSE}
total_sum2 = meds_sum2 + ind_sum2 + group_sum2 + med_ind_sum2 + control_sum2
total_sum2
```

---

What is the sum of squares total?

$$SS_{total} = \Sigma(X_{ij} - \overline{X})^2$$

$$SS_{total} = \Sigma X^2_{total} - \frac{(\Sigma X_{total})^2}{N}$$

```{r, include = FALSE}
962 - ((140^2)/25)
# 178
```

---

What is the sum of squares between/treatment?

$$SS_{treat} = n \Sigma(\overline{X_{j}} - \overline{X})^2$$

$$SS_{bn} = \Sigma(\frac{(\Sigma X\;in\;each\;column)^2}{n\;in\;each\;column}) - \frac{(\Sigma X_{total})^2}{N}$$

```{r, include = FALSE}
24^2
576/5

22^2
484/5

41^2
1681/5
 
10^2
100/5

43^2
1849/5

130^2
16900/25

(115.2 + 96.8 + 336.2 + 20 + 369.8) - 784
# 154
```

---

What is the sum of squares within/error?

$$SS_{wn} = SS_{total} - SS_{bn}$$


$$SS_{error} = SS_{total} - SS_{treat}$$

```{r, include = FALSE}
178 - 154
# 24
```

---

What are the degrees of freedom between?

$$df_{bn} = k - 1$$

```{r, include = FALSE}
5 - 1
# 4
```

---

What are the degrees of freedom within? (N - k)

```{r, include = FALSE}
25 - 5
# 20
```

---

What are the degrees of freedom total? (N - 1)

```{r, include = FALSE}
25 - 1
# 24
```

---

What is the mean squares between value?

$$MS_{bn} = \frac{SS_{bn}}{df_{bn}}$$

```{r, include = FALSE}
154/4
# ms between = 38.5
```


---

What is the mean squares within value?

$$MS_{wn} = \frac{SS_{wn}}{df_{wn}}$$

```{r, include = FALSE}
24/20
# 1.2
```

---

What is the F-obtained value?

$$F_{obt} = \frac{MS_{bn}}{MS_{wn}}$$

```{r, include = FALSE}
38.5/1.2
# f value is 32.08
```

---

Calculate the Tukey HSD

$$HSD = (q_{k})(\sqrt{\frac{MS_{wn}}{n}})$$

```{r, include = FALSE}
# df_wn = 20
# ms_wn = 1.2
# n = 5
# k = 5
# qk = 4.232

4.232*(sqrt(1.2/5))
# HSD = 2.07
```

---

```{r, include = FALSE}
# meds_mean is 4.8
# ind_mean is 4.4
# group_mean is 8.2
# med_ind_mean 2
# control_mean 8.6

4.8 - 4.4 #ns
8.2 - 4.8 #s
4.8 - 2 #s
8.6 - 4.8 #s

8.2 - 4.4 #s
4.4 - 2 #s
8.6 - 4.4 #s

8.2 - 2 #s
8.6 - 8.2 #ns

8.6 - 2 #s

```

---

$$\eta^2 = \frac{SS_{bn}}{SS_{total}}$$

```{r, include = FALSE}
eta2 = 154/178
eta2
```

---

What can we say?

- There is no significant difference between medications and individual therapy in depression scores
- There are lower depression scores in the group with medications compared to group therapy
- There are lower depression scores in the group with individual therapy and medications compared to medications alone
- There are lower depression scores in the group with medications compared to the control group
- There are lower depression scores in the group with individual therapy compared to group therapy

---

- There are lower depression scores in the group with individual therapy and medications compared to individual therapy alone
- There are lower depression scores in the group with individual therapy compared to the control group
- There are lower depression scores in the group with individual therapy and medications compared to group therapy
- There are no significant differences between the group therapy group and the control group
- There are lower depression scores in the group with individual therarpy and medications compared to the control group

- Effect size: The differences in therapeutic styles accounted for 91.61% of the variance in depression scores in our sample.


