---
title: "problem_set_11_and_12"
author: "JP"
date: "11/16/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

Problem Set 11

1. What are the three types of relationships?

2. If one variable deviates above its mean and another variable deviates above its mean, what is the direction of the relationship?

3. Is covariance a standardized measure?

4. What can a correlation coefficient value range from?

5. What are the effect sizes of r for small, medium, and large?

6. What would an effect size of r = .15 be?

7. What is the difference between a bivariate and partial correlation?

8. What are the two issues of correlational designs that cannot show cause and effect?

9. Draw what a partial correlation looks like. 



Problem Set 12

1. Create a composite score of smartphone use

2. Create a composite score of stress

3. Conduct a correlation between bmi and smartphone use

4. Report the correlation in writing.

There was a (significant/nonsignificant) relationship between ______ and ______ (r = #, p = #). 

5. Conduct a Spearman's correlation between stress levels and `mtuas_anxiety_q1`

6. Report the correlation in writing.

There was a (significant/nonsignificant) relationship between ______ and ______ (rho = #, p = #). 

$$\rho$$

```{r, include = FALSE}
jp <- rio::import(here::here("jp_thesis_1.sav")) %>% 
  janitor::clean_names() %>% 
  rowid_to_column() %>% 
  rename(sex = ccc_gender) %>% 
  mutate(bmi = ccc_bmi,
         stress = (dass_stress_q1 + dass_stress_q2 + dass_stress_q3 + dass_stress_q4 + dass_stress_q5 + dass_stress_q6 + dass_stress_q7)/7,
         email = (mtuas_email_q1 + mtuas_email_q2 + mtuas_email_q3 + mtuas_email_q4)/4,
         smartphone = (mtuas_smartphone_q1 + mtuas_smartphone_q2 + mtuas_smartphone_q3 + mtuas_smartphone_q4 + mtuas_smartphone_q5 + mtuas_smartphone_q6 + mtuas_smartphone_q7 + mtuas_smartphone_q8 + mtuas_smartphone_q9)/9) 

cor.test(jp$bmi, jp$smartphone)

cor.test(jp$stress, jp$mtuas_anxiety_q1, method = "spearman")
```


