---
title: "problem_set_10"
author: "JP"
date: "11/16/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
tired <- data.frame(time1 = c(1, 2, 4, 3, 2),
                     time2 = c(4, 5, 6, 4, 3),
                     time3 = c(9, 6, 8, 7, 10))
tired
```

1. Get the sum for each participant (adding up all of their time points)

```{r}
# participants
1+4+9
2+5+6
4+6+8
3+4+7
2+3+10
```


2. Get the sum of each level of our IV/factor

```{r}
time1_sum = sum(tired$time1)
time1_sum
time2_sum = sum(tired$time2)
time2_sum
time3_sum = sum(tired$time3)
time3_sum
```


3. Get the total sum of values

```{r}
total_sum = time1_sum + time2_sum + time3_sum
total_sum
```


4. Get the squared sum of each level of our IV/factor

```{r}
t1_square = 1^2 + 2^2 + 4^2 + 3^2 + 2^2
t1_square
t2_square = 4^2 + 5^2 + 6^2 + 4^2 + 3^2
t2_square
t3_square = 9^2 + 6^2 + 8^2 + 7^2 + 10^2
t3_square
```


5. Get the total sum of squared values

```{r}
total_square = t1_square + t2_square + t3_square
total_square
```


6. Get the means of each condition/level

```{r}
t1_m = mean(tired$time1)
t2_m = mean(tired$time2)
t3_m = mean(tired$time3)

t1_m
t2_m
t3_m
```


7. Get the n, the N, and the k

```{r}
total_n = 15
k = 3

total_n 
k
```

8. Get the sum of squares total

$$SS_{total} = \Sigma X^2_{total} - (\frac{(\Sigma X_{total})^2}{N})$$

$$SS_{total} = 466 - (\frac{(74)^2}{15})$$

```{r}
74^2
```

$$SS_{total} = 466 - (\frac{5476}{15})$$

```{r}
5476/15
```

$$SS_{total} = 466 - 365.07$$

```{r}
466-365.07
```

$$SS_{total} = 100.93$$

9. Get Sum of Squares Between Groups 

$$SS_{time} = \Sigma(\frac{(\Sigma X\;in\;each\;column)^2}{n\;in\;each\;column}) - (\frac{(\Sigma X_{total})^2}{N})$$

$$SS_{time} = \Sigma(\frac{12^2}{5} + \frac{22^2}{5} + \frac{40^2}{5}) - 365.07$$

```{r}
12^2
22^2
40^2
```

$$SS_{time} = \Sigma(\frac{144}{5} + \frac{484}{5} + \frac{1600}{5}) - 365.07$$

```{r}
144/5
484/5
1600/5
```

$$SS_{time} = (28.8 + 96.8 + 320) - 365.07$$

```{r}
28.8+96.8+320
```

$$SS_{time} = 445.6 - 365.07$$

```{r}
445.6 - 365.07
```

$$SS_{time} = 80.53$$

10. Get the Sum of Squares for the participants

$$SS_{subj} = \frac{(\Sigma X_{subj\;1}^2 + \Sigma X_{subj\;2}^2 + \Sigma X_{subj\;3}^2 + \Sigma X_{subj\;4}^2 + \Sigma X_{subj\;5}^2)}{k_{A}} - (\frac{(\Sigma X_{total})^2}{N})$$

$$SS_{subj} = \frac{(14^2 + 13^2 + 18^2 + 14^2 + 15^2)}{3} - 365.07$$

```{r}
14*14
13*13
18*18
14*14
15*15
```

$$SS_{subj} = \frac{(196 + 169 + 324 + 196 + 225)}{3} - 365.07$$

```{r}
196 + 169 + 324 + 196 + 225
```

$$SS_{subj} = \frac{(1110)}{3} - 365.07$$

```{r}
1110/3
```

$$SS_{subj} = 370 - 365.07$$

```{r}
370-365.07
```

$$SS_{subj} = 4.93$$

11. Get the error of our factor by participants

$$SS_{error} = SS_{total} - SS_{time} - SS_{subj}$$

$$SS_{error} = 100.93 - 80.53 - 4.93$$

```{r}
100.93 - 80.53 - 4.93
```

$$SS_{error} = 15.47$$

12. Get the Degrees of Freedom Between Groups

$$df_{time} = k_{time} - 1$$

$$df_{time} = 3 - 1$$

```{r}
3 - 1
```


$$df_{time} = 2$$

13. Get the Degrees of Freedom For the Error (this is a combination of the df for between groups and for subjects/participants)

$$df_{error} = (k_{time} - 1)(k_{subj} - 1)$$

$$df_{error} = (3 - 1)(5 - 1)$$

```{r}
3-1
5-1
```

$$df_{error} = (2)(4)$$

```{r}
2*4
```

$$df_{error} = 8$$

14. Get the Mean Square for The Factor/IV

$$MS_{time} = \frac{SS_{time}}{df_{time}}$$

$$MS_{time} = \frac{80.53}{2}$$

```{r}
80.53/2
```

$$MS_{time} = 40.27$$

15. Get the Mean Square of the Error of the Factor by the Participants

$$MS_{error} = \frac{SS_{error}}{df_{error}}$$

$$MS_{error} = \frac{15.47}{8}$$

```{r}
15.47/8
```

$$MS_{error} = 1.93$$


16. Within-Subjects F-statistic

$$F_{obt} = \frac{MS_{time}}{MS_{error}}$$

$$F_{obt} = \frac{40.27}{1.93}$$

```{r}
40.27/1.93
```

$$F_{obt} = 20.87$$

Is your finding statistically significant?
