---
title: "PSY 3307"
subtitle: "Repeated Measures ANOVA"
author: "Jonathan A. Pedroza PhD"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 99999)

library(tidyverse)
library(rstatix)
```

# Agenda

- Within-Subjects Design  
  - Repeated Measures ANOVA
  - Differences Between Paired-Samples t-test & Repeated Measures ANOVA
- Longitudinal Designs
- Why They Are Useful
- Assumptions of Repeated Measures ANOVA
- Calculations for One-way Within-Subjects ANOVA
  - Repeated Measures ANOVA
- Repeated Measures in SPSS

---

# Previously With ANOVA

- IV should be two more more groups
  - preferably three groups or more

- no outliers
  - **Cook's distance** is a method to find influential cases/outliers in your model
    
    + used for both ANOVA and regression models

- **homogeneity of variance** is the assumption that each population has the same variance

- **normality** DV values are normally distributed

- **independence** observations are independent of one another
  - it really is that the residual/error is independent but for now we'll keep it as observations are different from one another
  
---

# Within-Subjects Design

- Every participant gets each condition
  - the way we will be using this design is with a **repeated measures** design
  
    + or that every participant will see the same measure multiple times
    
- An example would be to look at a depression scale before the intervention, after the intervention, and then 6 months after the end of the intervention

- Outside a researcher perspective, we could think of this as a program evaluator
  - you could be evaluating a community on what they think about their safety
    
    + so you can ask before, during the program promoting safety in the community, and months afterward to see if your program is still working
    
---

# Comparisons to paired-samples t-test

- Repeated measures is similar to the paired-samples t-test in that we can test multiple time points for every participant

- however, we can test more than two time points with repeated measures ANOVA

- we can also looked at mixed designs (between-subjects x within-subjects)
  - we'll get to this soon
---

# Longitudinal Designs

- this is a major reason why this statistical test is so useful
  - this test accounts for independence of observations/residuals

- **longitudinal designs** are when we are following our sample for a period of time
  - these designs get much more complex than this (same group of people, similar groups over different years)
  
---

# Repeated Measures ANOVA

- no outliers
  - **Cook's distance** is a method to find influential cases/outliers in your model
    
    + used for both ANOVA and regression models

- **homogeneity of variance** is the assumption that each population has the same variance

- **normality** DV values are normally distributed

- **independence** is not observed for repeated measures ANOVA because the test accounts for repeating values (same construct different time points)

---

# Repeated Measures ANOVA

- the lack of independence would be a problem but repeated measures ANOVA **partitions** or removes the dependence imposed by multiple measurements on the same participants

- this is due by partialling out or by getting rid of the overlap in the proportion of variance explained
  - we'll get to this when we get to regression

- normal linear regression would actually be worse than repeated measures ANVOA in this case

---

# When won't it work

- missing data in the DV

- unbalanced amount of participants

- time is continuous

- covariates that vary by time

- nested/hierarchical models (students in classrooms in schools)

- non-continuous DV

---

# Example 

```{r}
hungry <- data.frame(time1 = c(9, 7, 8, 6, 10),
                     time2 = c(4, 5, 4, 3, 5),
                     time3 = c(6, 5, 5, 6, 7))
hungry
```

---

# Calculations

### Let's get the sum for each participant (adding up all of their time points)

```{r}
participant1_sum = 9+4+6
participant2_sum = 7+5+5
participant3_sum = 8+4+5
participant4_sum = 6+3+6
participant5_sum = 10+5+7

participant1_sum
participant2_sum
```

---

```{r}
participant3_sum
participant4_sum
participant5_sum
```


---

### Now, let's get the sum of each level of our IV/factor

```{r}
time1_sum = 9+7+8+6+10
time2_sum = 4+5+4+3+5
time3_sum = 6+5+5+6+7

time1_sum
time2_sum
time3_sum

```

---

### Let's also get the total sum of values

```{r}
total_sum = time1_sum + time2_sum + time3_sum
total_sum
```

---

### Now we can also get the squared sum of each level of our IV/factor

```{r}
time1_sum_square = 9^2+7^2+8^2+6^2+10^2
time2_sum_square = 4^2+5^2+4^2+3^2+5^2
time3_sum_square = 6^2+5^2+5^2+6^2+7^2

time1_sum_square
time2_sum_square
time3_sum_square
```

---

### And the total sum of squared values

```{r}
total_sum_square = time1_sum_square + time2_sum_square + time3_sum_square
total_sum_square
```

---

### Let's also get the mean

```{r}
time1_mean = time1_sum/5
time2_mean = time2_sum/5
time3_mean = time3_sum/5

time1_mean
time2_mean
time3_mean
```

---

### Lastly, let's get the n, the N, and the k

```{r}
time1_n = 5
time2_n = 5
time3_n = 5

time1_n
time2_n
time3_n
```

---

```{r}
k = 3
k

total_n = time1_n + time2_n + time3_n
total_n
```


---

### Let's get the sum of squares total

$$SS_{total} = \Sigma X^2_{total} - (\frac{(\Sigma X_{total})^2}{N})$$

$$SS_{total} = 592 - (\frac{(90)^2}{15})$$

---

```{r}
90^2
```


$$SS_{total} = 592 - (\frac{8100}{15})$$

---

```{r}
8100/15
```


$$SS_{total} = 592 - 540$$

---

```{r}
592 - 540
```

$$SS_{total} = 52$$
---

### Sum of Squares Between Groups 

$$SS_{time} = \Sigma(\frac{(\Sigma X\;in\;each\;column)^2}{n\;in\;each\;column}) - (\frac{(\Sigma X_{total})^2}{N})$$

$$SS_{time} = (\frac{40^2}{5} + \frac{21^2}{5} + \frac{29^2}{5}) - 540$$

---

```{r}
40^2
21^2
29^2
```

$$SS_{time} = (\frac{1600}{5} + \frac{441}{5} + \frac{841}{5}) - 540$$

---

```{r}
1600/5
441/5
841/5
```

$$SS_{time} = (320 + 88.2 + 168.2) - 540$$

---

```{r}
320 + 88.2 + 168.2
```

$$SS_{time} = 576.4 - 540$$

---

```{r}
576.4 - 540
```

$$SS_{time} = 36.4$$

---

### Now on to the Sum of Squares for the participants

$$SS_{subj} = \frac{(\Sigma X_{subj\;1}^2 + \Sigma X_{subj\;2}^2 + \Sigma X_{subj\;3}^2 + \Sigma X_{subj\;4}^2 + \Sigma X_{subj\;5}^2)}{k_{A}} - (\frac{(\Sigma X_{total})^2}{N})$$

$$SS_{subj} = \frac{(19^2 + 17^2 + 17^2 + 15^2 + 22^2)}{3} - 540$$

---

```{r}
19^2
17^2
17^2
15^2
22^2
```

$$SS_{subj} = \frac{(361 + 289 + 289 + 225 + 484)}{3} - 540$$

---

```{r}
361 + 289 + 289 + 225 + 484
```


$$SS_{subj} = \frac{1648}{3} - 540$$

---

```{r}
1648/3
```

$$SS_{subj} = 549.33 - 540$$

---

```{r}
549.33 - 540
```

$$SS_{subj} = 9.33$$

---

### Lastly the interaction of our factor by participants

$$SS_{error} = SS_{total} - SS_{time} - SS_{subj}$$

$$SS_{error} = 52 - 36.4 - 9.33$$

---

```{r}
52 - 36.4 - 9.33
```

$$SS_{error} = 6.27$$

---

### Degrees of Freedom Between Groups

$$df_{time} = k_{time} - 1$$

$$df_{time} = 3 - 1$$

```{r}
3 - 1
```

$$df_{time} = 2$$

---

### Degrees of Freedom For the Interaction

$$df_{error} = (k_{time} - 1)(k_{subj} - 1)$$

$$df_{error} = (3 - 1)(5 - 1)$$

---

```{r}
(3 - 1)
(5 - 1)
```

$$df_{error} = (2)(4)$$

---

```{r}
2*4
```


$$df_{error} = 8$$

---

### Mean Square for The Factor/IV

$$MS_{time} = \frac{SS_{time}}{df_{time}}$$

$$MS_{time} = \frac{36.4}{2}$$

---

```{r}
36.4/2
```

$$MS_{time} = 18.2$$

---

### Mean Square of the Interaction of Factor by the Participants

$$MS_{error} = \frac{SS_{error}}{df_{error}}$$

$$MS_{error} = \frac{6.27}{8}$$

---

```{r}
6.27/8
```

$$MS_{error} = .78$$

---

### Within-Subjects F-statistic

$$F_{obt} = \frac{MS_{time}}{MS_{error}}$$

$$F_{obt} = \frac{18.2}{.78}$$

---

```{r}
18.2/.78
```

$$F_{obt} = 23.33$$

---

```{r}
tired <- data.frame(time1 = c(1, 2, 4, 3, 2),
                     time2 = c(4, 5, 6, 4, 3),
                     time3 = c(9, 6, 8, 7, 10))
tired
```

