---
title: "SPSS Assignment 1"
author: "JP"
date: "10/1/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Context (20 points)

You are interested in whether there are differences in the amount of phone calls that are being made between male and female students. You think that there will be a difference between the two groups but you're not sure which group will have a higher score. Create hypotheses and conduct all the analyses and visualizations below. You will submit your answers to anything that is not SPSS specific (hypotheses, t critical value, write up) on this **word document and will turn in this document** as well as your **SPSS output file**. Email me if you have any questions about submitting both documents. 

## Create a `null` and `alternative` hypothesis (in words) (3 points)

H0:

H1: 

## Rename `gender` into `sex` (1 point)

## Get a composite average score for phone calls (2 points)

```{r}
library(tidyverse)

jp <- rio::import(here::here("jp_thesis_1.sav")) %>% 
  janitor::clean_names() %>% 
  rowid_to_column() %>% 
  mutate(phone = (mtuas_phonecall_q1 + mtuas_phonecall_q2)/2,
         sex = ccc_gender)
```

- there are two variables that will make up the composite score

## Conduct descriptive statistics for the following variables (2 points)

- please write out what the mean and standard deviation for the composite score to get full credit

- make sure to include your initials after the composite score

```{r}
jp %>% 
  summarize(mean = mean(phone),
            sd = sd(phone))
```


- `sex`
- `mtuas_phonecall_q1`
- `mtuas_phonecall_q2`
- `avg phonecall variable`

```{r}
jp %>% 
  ggplot(aes(sex)) + 
  geom_bar()

jp %>% 
  ggplot(aes(phone)) + 
  geom_histogram(color = "white", bins = 15) 
```

## Get a descriptive visual for the composite score (1 point)

- is the skewness acceptable for the composite score

- is the distribution "normal" enough


## Conduct inferential statistic (independent samples t-test) (3 points)

```{r}
t.test(phone ~ sex, data = jp)
t.test(phone ~ sex, data = jp, var.equal = TRUE)
```

- report the correct inferential statistics in the form of a t-test (e.g., t(df) = t value, p value, 95% CI [lower confidence interval, upper confidence interval]) 

## Tell me whether or not your finding is statistically significant (1 point)

## find the t-critical value for a two-tailed test (2 points)

- look at your t table to see the t critical value for your test

## Show some type of visualization that shows the difference in the two groups on internet use (2 points)

```{r}
jp %>% 
  ggplot(aes(factor(sex), phone)) + 
  geom_boxplot(aes(color = factor(sex)))
```


## Reject/Retain the Null Hypothesis? (1 point)

## Write up on your findings (2 points)