---
title: "PSY 3307"
subtitle: "t-test Combined"
author: "Jonathan A. Pedroza PhD"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 99999)

library(tidyverse)
library(psych)
library(ggmap)
library(maps)
library(RColorBrewer)
library(ggrepel)
library(gganimate)
library(agricolae)
library(ggridges)
```

# What is a One-Sample t-test

- It's pretty similar to a z-test
  - t-test used more often in behavioral research

- z-test requires we know population standard deviation
  - often not possible in behavioral research

- uses unbiased estimators (N - 1 formulas)

- computes something like the z-score for our sample mean
  - t-score
  
---

# One-Sample t-test

- parametric test for when the population standard deviation is unknown

- still compares the sample mean to the population mean

---

# Steps to One-Sample t-test

1. Statistical Hypotheses
  - what is the population mean and is your sample mean different from that population mean
    
    - H0: sample mean equals the population mean
    
    - H1: sample mean is different from the population mean
    
2. Select an alpha

3. Check assumptions
  
  - Outcome needs to be continuous (interval or ratio scale)
  
  - Population score forms a normal distribution
  
  - variability of raw score population is estimated from the sample
  
---

# Steps to One-Sample t-test

- All we need to know is the t critical value and if the t obtained value is within the regions of rejection

---

# Steps to a z-test/One-Sample t-test

- get population standard deviation (z-test)

- get estimated standard deviation (t-test)

--

- get the standard error (SE) of the mean (z-test)

- get the **estimated** SE (t-test)

--

- calculate the score by subtracting the population mean from the sample mean and dividing by the SE
  - either obtained z or t value

---

# Changes between the z-test and t-test

$$S^2_{X} = \frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N - 1}$$

$$S_{x} = \sqrt{\frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N - 1}}$$

.pull-left[
$$\sigma_{\overline{X}} = \frac{\sigma_{X}}{\sqrt{N}}$$

$$z_{obt} = \frac{\overline{X} - \mu}{\sigma_{\overline{X}}}$$
]

.pull-right[
$$S_{\overline{X}} = \frac{S_{X}}{\sqrt{N}}$$

$$t_{obt} = \frac{\overline{X} - \mu}{S_{\overline{X}}}$$
]

---

# Small Change in Formulas

- SE calculation will start to look slightly different as it will use the variance squared

- Due to future formulas using slightly different notation, we will adopt that for our SE

$$S_{\overline{X}} = \sqrt{\frac{S^2_{x}}{N}}$$

---

# Example

```{r}
set.seed(092221)

numbers = rnorm(10, mean = 5, sd = 1.2)

numbers
```


1. Calculate the Variance

2. Calculate the SE

3. Compute t

---

```{r}
# population mean is 10

4.770670 + 5.271329 + 6.899812 + 5.598090 + 4.219022 + 6.828034 + 6.046497 + 2.921249 + 4.993870 + 5.948126
# 53.50

53.50/10
# 5.35

4.770670^2 + 5.271329^2 + 6.899812^2 + 5.598090^2 + 4.219022^2 + 6.828034^2 + 6.046497^2 + 2.921249^2 + 4.993870^2 + 5.948126^2
# 299.33

53.50^2
# 2862.25
```

---

```{r}
2862.25/10
# 286.23

299.33 - 286.23
# 13.1

13.1/9
# 1.46 variance
```

---

```{r}
# se is...
1.46/10

sqrt(1.46/10)
sqrt(.146)

# compute t
(5.35 - 10)/.38

# t value of -12.24
```

---

# t-distribution & Degrees of Freedom (df)

- we will now be working with the t-distribution
  - this also means we'll be working with a t-table
  
- **t-distribution** is the sampling distribution of all values of t when samples of a particular size (differing N size) are selected from the raw score population in the null hypothesis

---
  
# t-distribution & Degrees of Freedom (df)

- higher values on the t-distribution are to the right of the population mean, lower values to the left of the population mean

- t-tests also have regions of rejection

- doesn't always represent a perfectly normal distribution
  - dependent on N value
  
    + larger the sample the more normal the distribution looks

- the different shapes are important because our regions of rejection will look different dependent on the sample size

---

# t-distribution & Degrees of Freedom (df)

- the distribution changes based on the sample size, which then means that the 5% of the regions of rejection and critical value change

- remember to be conservative about estimating variance and SD, we have been using `N - 1`

- the name of that is the **degrees of freedom** or df
  - number of scores in a sample that reflect the variability in the population
  - determines shape of sampling distribution when estimating standard deviation for the population
  
---

# t-distribution & Degrees of Freedom (df)

- since the df is the sample size - 1, the larger the df, the closer to resembling a normal distribution our data becomes
  - df of 120+ is the same as a z-distribution
  
---

# Using the t-table

- the t-table is different from the z-table

- has df, \alpha = .05 and \alpha = .01
  - this is dependent on our sample size - 1, and what our alpha is `a priori` 
  
---

# t-table

- we need to figure out our `t critical value`

- we need our sample size, and a decision on what alpha we want to use (.05 or .01)

- since not all df are listed, if your df is between two values, a statistically significant finding is a t-value larger than the larger df and smaller than the smaller df

---

# Examples

- sample size = 200
  - alpha = .05
  
- sample size = 90  
  - alpha = .05
  
- sample size = 37
  - alpha = .01

---

# t-test Interpretation

- If a statistically significant finding is found
  - your sample is significantly different from the population in whatever the outcome was

---

# One-tailed one-sample t-test

- if you know if your sample will do better or worse than the population, you'd use a one-tailed test
  
- Example: you know that your sample will get higher grades than the population

---

# Confidence Intervals

- **point estimation** is a way to estimate a point where you think the population's outcome value will be
  - this is why we can't say we're certain mu is a specific number and have to say *around* that number

- **interval estimation** is when we state that mu will fall within a range of values
  - margin of error, such as getting an exam and stating that the average test score was 84 plus or minus 3 points
    - due to sampling error
  
- **confidence intervals** are a range of values which we are certain our value falls within
  - when we say *around* a value, we are saying that we got one value but we are certain it is within a range of values
  - around 84 points on an exam, but we are certain the correct value is between 80 and 87
  
---

# Confidence Intervals

- We're choosing a range of values that are not significantly different from our sample mean

- we compute confidence intervals after we have a statistically significant finding

- It is often stated as:
  - We got a statistically significant finding where our sample scored ___ points compared to the population's score ___; *t*(df) = t-value, p-value
    - Example: *t*(31) = 4.7, p = .037

---

```{r}
coffee <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-07/coffee_ratings.csv') %>% 
  mutate(species = as.factor(species),
         process = recode(processing_method, "Washed / Wet" = "washed",
                          "Semi-washed / Semi-pulped" = "not_washed",
                          "Pulped natural / honey" = "not_washed",
                          "Other" = "not_washed",
                          "Natural / Dry" = "not_washed",
                          "NA" = NA_character_),
         process = as.factor(process),
         species = as.factor(species),
         country_of_origin = as.factor(country_of_origin),
         variety = as.factor(variety)) %>% 
  drop_na(process, color)
```

---
```{r}
psych::describe(coffee$total_cup_points, na.rm = TRUE)
# mean is 82.03
# SE is .08
# sample size is 1071
```


---

```{r}
t.test(coffee$total_cup_points, mu = 85) #conf int only works for two tailed test

```

---

```{r}
t.test(coffee$total_cup_points, mu = 85, alternative = "less")

```

---

```{r}
t.test(coffee$total_cup_points, mu = 85, alternative = "greater")
```

---

# Confidence Interval Calculations

$$(s_{x})(-t_{crit}) + \overline{X} \; \leq \;  \mu \; \leq \; (s_{x})(t_{crit}) + \overline{X}$$

```{r}
# t critical value is 1.96 since we have such a large sample and df

# mu = 85
# sample mean = 82.03
# SE = .08
# df = 1070

# lower
.08*-1.96 + 82.03
# 81.8732

# higher
.08*1.96 + 82.03
# 82.1868
```

---

```{r}
t.test(coffee$total_cup_points, mu = 85)
```

*t*(1070) = -36.39, *p* < .05, 95% CI [81.87, 82.19]

Our one-sample t-test comparing a sample of coffee ratings (*M* = 82.03, *SD* = 2.67) to the population of coffee ratings (*M* = 85) showed evidence of a statistically significant difference. Specifically, the sample's average coffee rating was significantly lower than the population's average coffee rating; *t*(1070) = -36.39, *p* < .05, 95% CI [81.87, 82.19]. We are 95% certain that the actual sample mean is between 81.87 and 82.19. 

---

# Independent-samples t-test

---

# Between & Within Designs

- Experiments can be broken down into two different types of designs

- **Between-subject/group** design is when you are interested in comparing two (for now) or more groups on an outcome variable

- **Within-subject/group** design is when you have the same participants but you test them twice (either with two different variables or two different time points)

---

# Two tests we are talking about

- **independent samples t-test** is when there are two groups of participants are separated into two different conditions to compare based on that condition
	- comparing the physical activity levels (DV) of sexes (Condition 1 = Male, Condition 2 = Female)
	- parametric test

- **paired-samples t-test** is when there are two experimental conditions that the same participants take part in
	- interested in two variables in the same sample of participants
	- can be the same variable and two different time points
	- bmi levels before an experiment and after the experiment for all participants
	- parametric test

---

# Independent Samples t-test

- JP note: probably the most often used t-test

- because it is a parametric test, it has assumptions

- Assumptions are
  - DV is normally distributed interval/ratio scores
  - populations have homogeneous variance
  - not a true assumption but something important to note is that your groups should be equal in `n` (condition) size
  
---

# Homogeneity of Variance

- **homogeneity of variance** is when the variances of the populations represented in a study have "equal" variances

- in order to test that the variances are equal, we can look at it through visuals
  - however, a better option is to use the Levene's test

---

# Independent samples t-test

- hypotheses are now focused on the differences between the two groups/conditions

$$H0: \mu_{1} - \mu_{2} = 0$$

H0: There will be no difference in DV scores between group 1 and group 2.
  - both samples/groups represent the population

$$H1: \mu_{1} - \mu_{2} \neq 0$$

H1: There will be differences in DV scores between group 1 and group 2.
  - the groups represent a different population or don't represent the current population

---

# t-distribution for independent samples t-test

- we are interested in the difference between our group/sample means

- we have two samples from one raw score population 

- **sampling distribution of differences between means** show all differences between two means that occur when random samples are drawn from a population of scores

- the mean of the sampling distribution is zero because both sample means will equal the population mean of the raw score population

---

# Independent samples t-test

- determines the probability of obtaining our difference between our means when H0 is true

- Term changes
  - N is now the full sample size
  - n is the size of each group/sample/condition
    * so for each group/sample/condition, we have an n
    
---

# Performing the indepdendent samples t-test

$$s^2_{x} = \frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N - 1}$$

```{r}
male_scores = c(4, 6, 2, 3, 5, 1, 2, 4, 3, 5)
female_scores = c(4, 6, 9, 6, 5, 8, 2, 5, 3, 7)

male_scores
female_scores
```

---

First we'll calculate the variance

```{r}
# male sum
4+6+2+3+5+1+2+4+3+5
# sum is 35
```

$$\;s^2_{x_{1}} = \frac{\Sigma X^2 - \frac{(35)^2}{N}}{N - 1}$$

---

```{r}

# female sum
4+6+9+6+5+8+2+5+3+7
# sum 55
```

$$\;s^2_{x_{2}} = \frac{\Sigma X^2 - \frac{(55)^2}{N}}{N - 1}$$

---

```{r}
35/10
# male mean 3.5
```


```{r}
55/10
# female mean 5.5
```

---

```{r}
# male sum of squared Xs
4^2+6^2+2^2+3^2+5^2+1^2+2^2+4^2+3^2+5^2
# 145 
```

$$\;s^2_{x_{1}} = \frac{145 - \frac{(35)^2}{10}}{10 - 1}$$

---

```{r}
# female sum of squared Xs
4^2+6^2+9^2+6^2+5^2+8^2+2^2+5^2+3^2+7^2
# female 345
```

$$\;s^2_{x_{2}} = \frac{345 - \frac{(55)^2}{10}}{10 - 1}$$

---

```{r}
# male sum of X squared and divided by N
35^2
1225/10
# 122.5
```

$$\;s^2_{x_{1}} = \frac{145 - \frac{1225}{10}}{10 - 1}$$

---

```{r}
# female sum of X squared and divided by N
55^2
3025/10
# 302.5
```

$$\;s^2_{x_{2}} = \frac{345 - \frac{302.5}{10}}{10 - 1}$$

---

```{r}
# male variance calculations
(145 - 122.5)/(10-1)
# variance is 2.5
```

$$\;s^2_{x_{1}} = \frac{145 - 122.5}{10 - 1}$$

---

```{r}
# female variance calculations
(345 - 302.5)/(10 - 1)
# variance is 4.72

```

$$\;s^2_{x_{2}} = \frac{345 - 302.5}{10 - 1}$$
---

```{r}
sd(male_scores)^2
sd(female_scores)^2
```

---

# New Terms

- **pooled variance** is the weighted average variance of the groups'/samples' variances in a independent samples t-test

- **standard error of the difference** is the estimated standard deviation of the sampling distribution of differences between the means

---

Now we can calculate the pooled variance
n1 = 10
n2 = 10
variance of group 1 = 2.5
variance of group 2 = 4.72

$$S^2_{pool} = \frac{(n_{1} - 1)S^2_{1} + (n_{2} - 1)S^2_{2}}{(n_{1} - 1) + (n_{2} - 1)}$$

---

```{r}
# start with the numerator
(10 - 1)*2.5 + (10 - 1)*4.72
# numerator is 64.98

# denominator
(10 - 1) + (10 - 1)
# denominator is 18
```

$$S^2_{pool} = \frac{(10 - 1)2.5 + (10 - 1)4.72}{(10 - 1) + (10 - 1)}$$

---

```{r}
9*2.5 + 9*4.72
# 64.98

9+9
# 18
```

$$S^2_{pool} = \frac{(9)2.5 + (9)4.72}{9 + 9}$$

---

```{r}
64.98/18
# pooled variance is 3.61
```

$$S^2_{pool} = \frac{64.98}{18}$$

$$S^2_{pool} = 3.61$$

---

Let's calculate for the standard error of the difference

$$S_{\overline{X_{1}} - \overline{X_{2}}} = \sqrt{(S^2_{pool})(\frac{1}{n_{1}} + \frac{1}{n_{2}})}$$

---

```{r}
1/10

3.61*(.1 + .1)

sqrt(.72)
```

$$S_{\overline{X_{1}} - \overline{X_{2}}} = \sqrt{(3.61)(\frac{1}{10} + \frac{1}{10})}$$

$$S_{\overline{X_{1}} - \overline{X_{2}}} = \sqrt{(3.61)(.1 + .1)}$$
---

```{r}
3.61*(.1+.1)
# se is .72
```

$$S_{\overline{X_{1}} - \overline{X_{2}}} = \sqrt{.72}$$

---

```{r}
sqrt(.72)
# se of the difference is .85
```
 
$$S_{\overline{X_{1}} - \overline{X_{2}}} = .85$$ 

---

Now we can calculate the independent samples t-test obtained value

$$t_{obt} = \frac{(\overline{X_{1}} - \overline{X_{2}}) - (\mu_{1} - \mu_{2})}{S_{\overline{X_{1}} - \overline{X_{2}}}}$$

**Note** the population mean 1 minus the population mean 2 is what is specified in the null hypothesis, so it will be zero

```{r}
((3.5 - 5.5) - 0)/.85
# t obtained value is -2.35
```

$$t_{obt} = \frac{(3.5 - 5.5) - 0}{.85}$$

---

Let's now calculate the degrees of freedom

$$df = (n_{1} - 1) + (n_{2} - 1)$$
---

```{r}
(10 - 1) + (10 - 1)
# df is 18

# t critical is +-2.101
```

$$df = (10 - 1) + (10 - 1)$$

$$df = 18$$

---

So we get a value of -2.35 and the t-critical value is -2.101

Is there a statistically significant difference between the two groups?

--

-2.35 > -2.101

---

Now let's get confidence intervals

$$Lower\;Bound: (\overline{X}_{1} - \overline{X}_{2}) - t_{\alpha/2}\;*\;\sqrt{\frac{S^2_{1}}{n_{1}} + \frac{S^2_{2}}{n_{2}}}$$

$$Upper\;Bound: (\overline{X}_{1} - \overline{X}_{2}) + t_{\alpha/2}\;*\;\sqrt{\frac{S^2_{1}}{n_{1}} + \frac{S^2_{2}}{n_{2}}}$$

---

```{r}
# group 1 mean = 3.5
# group 2 mean = 5.5
# t critical value is 2.101
# n1 = 10
# n2 = 10
# variance of group 1 = 2.5
# variance of group 2 = 4.72

# lower
(3.5 - 5.5) - 2.101 * sqrt((2.5/10) + (4.72/10))
# -3.79

# upper
(3.5 - 5.5) + 2.101 * sqrt((2.5/10) + (4.72/10))
# -.21
```

---

# Effect Sizes

- Reminder: r effect sizes are .1 = small, .3 = medium, .5 = large

- Reminder: cohen's d effect sizes are .2 = small, .5 = medium, .8 = large

- these are both measures of the strength of a relationship
  - better than simply using p value alone

- cohen's d can never be negative so the value you get is the absolute value (.e.g., its always positive)

- if unequal sample sizes in groups/conditions then you'll use `Hedges' g`
  - same formula and will be roughly the same once sample sizes get larger than 20 (N = 20)
  
---

$$d = \frac{(\overline{X_{1}} - \overline{X_{2}})}{\sqrt{S^2_{pool}}}$$

Small sample sizes use the following formula (samples under 50)

$$d = \frac{(\overline{X_{1}} - \overline{X_{2}})}{\sqrt{S^2_{pool}}} * (\frac{N - 3}{N - 2.25}) * \sqrt{\frac{N - 2}{N}}$$

---

```{r}
(3.5 - 5.5)/sqrt(3.61)

# each step below
3.5 - 5.5

-2/sqrt(3.61)

# cohen's d is .92 or the number of standard deviations between the means
```

$$d = \frac{(3.5 - 5.5)}{\sqrt{3.61}}$$

---

```{r}
(3.5 - 5.5)/sqrt(3.61)

(10-3/10 -2.25)

sqrt((10-2)/10)

-1.05*7.45*.89
```


$$d = \frac{(3.5 - 5.5)}{\sqrt{3.61}} * (\frac{10 - 3}{10 - 2.25}) * \sqrt{\frac{10 - 2}{10}}$$

---

# Steps for independent samples t-test

1. Get the means of both groups/samples

2. Get the variances of both groups/samples

3. Get the group/sample sizes (n)

4. Get the pooled variance by getting the groups'/samples' variances averaged

5. Get the standard error of the differences

6. Calculate the t-obtained value

7. Get the degrees of freedom

8. Calculate the confidence intervals

9. Get the effect size

---

# Independent samples t-test Example

```{r, eval = TRUE, echo = FALSE}
set.seed(093021)

df <- data.frame(
  sex=factor(rep(c("F", "M"), each=200)),
  weight=round(c(rnorm(200, mean=55, sd=5),
                 rnorm(200, mean=65, sd=5)))
  )

describeBy(df$weight, df$sex, na.rm = TRUE)
```

---

```{r, eval = TRUE, echo = FALSE}
df %>% 
  ggplot(aes(weight, color = sex, fill = sex)) +
  geom_density(alpha = .5) +
  geom_vline(xintercept = 54.7, color = "#d74122", size = 1.25) +
  geom_vline(xintercept = 65.36, color = "#387448", size = 1.25) +
  scale_fill_manual(values = c("#d74122", "#387448")) +
  scale_color_manual(values = c("#d74122", "#387448")) +
  theme_minimal()

```

---

```{r, eval = TRUE, echo = FALSE}
car::leveneTest(weight ~ sex, data = df)

t.test(weight ~ sex, data = df, var.equal = TRUE)
```

---

# Real life independent samples t-test

```{r, eval = TRUE, echo = FALSE}
jp <- rio::import(here::here("jp_thesis_1.sav")) %>% 
  janitor::clean_names()

jp <- jp %>% 
  mutate(stress_avg = (dass_stress_q1 + dass_stress_q2 + dass_stress_q3 + dass_stress_q4 + dass_stress_q5 + dass_stress_q6 + dass_stress_q7)/7,
         exercise_avg = (cpaq_light_exercise + cpaq_moderate_exercise + cpaq_vigorous_exercise + cpaq_muscle_exercise)/4,
         depression_avg = (dass_depress_q1 + dass_depress_q2 + dass_depress_q3 + dass_depress_q4 + dass_depress_q5 + dass_depress_q6 + dass_depress_q7)/7,
         unhealthy_snacks_avg = (ffq_cake_portion + ffq_saltysnacks_portion + ffq_chocolate_portion + ffq_icecream_portion)/4,
         games = (mtuas_videogaming_q1 + mtuas_videogaming_q2 + mtuas_videogaming_q3)/3,
         gender = ccc_gender,
         gender = factor(gender),
         class_standing = ccc_class_standing,
         bmi = ccc_bmi)
```

---

```{r, eval = TRUE, echo = FALSE}
jp %>% 
  group_by(gender) %>% 
  summarize(games = mean(games))
```

---

```{r, eval = TRUE, echo = FALSE}
jp %>% 
  ggplot(aes(games)) + 
  geom_density(aes(color = gender, fill = gender), alpha = .3) +
  theme_minimal()
```

---

```{r, eval = TRUE, echo = FALSE}
jp %>% 
  mutate(gender = recode(gender, "1" = "Male",
                         "2" = "Female")) %>% 
  ggplot(aes(games, gender)) + 
  geom_density_ridges(aes(fill = gender), alpha = .7, scale = 1) +
  geom_vline(xintercept = 4.23, color = "#d74122", size = 1.25) +
  geom_vline(xintercept = 3.3, color = "#387448", size = 1.25) +
  scale_fill_manual(values = c("#d74122", "#387448")) +
  labs(x = "Average Time Spent Playing Video Games",
       y = "") +
  theme_minimal()
```

---

```{r, eval = TRUE, echo = FALSE}
car::leveneTest(games ~ gender, data = jp)

t.test(games ~ gender, data = jp, var.equal = TRUE)
```

---

```{r, eval = TRUE, echo = FALSE}
describe(jp$games, na.rm = TRUE)
```

---

# One-tail independent-samples t-test

- used when we are confident that the direction of the relationship

- if you think one group will have a higher score/value then you would make that statement in your alternative hypothesis

---

# Steps for one-tailed test

- Variable(Groups) = Intervention(Got Intervention/Control)
- Outcome = Fruit & Vegetable Intake

1. decide which sample/group score you think will be larger
  - I think that the intervention group will eat more fruits and vegetables

2. decide which condition/group to subtract from the other
  - `intervention group - control group`

3. decide whether the difference will be positive or negative
  - a positive value should be returned from the previous equation
  
---

4. create hypotheses
  - The intervention group will eat the same or less fruits and vegetables than the control group
    - H0: mu(intervention group) - mu(control group) ≤ 0
  - The intervention group will eat more fruits and vegetables than the control group
    - H1: mu(intervention group) - mu(control group) > 0
    
---

5. locate regions of rejection
  - since we are interested in a positive difference in our hypothesis, we'll only be looking at the upper tail of the distribution

6. conduct your independent samples t-test
  - make sure your groups/samples are subtracted the same way as your hypotheses
    - `intervention group - control group`

---

# Paired Samples t-test

- within-subjects/groups design

- also called related-samples t-test
  - each participant gets measured in each condition

- an example would be an intervention for weight loss where everyone's weight is measured before (first time point to measure) the intervention and after (second time point to measure)

- **matched-samples design** is when a participant in one condition is matched with a participant in the other condition

- **repeated-measures design** is when each participant gets measured twice
  - can be more for more rigorous statistics, but not the ones we'll learn about in this class

---

# Why we use paired-samples t-tests

- pretest/posttest design
  - before everyone gets the intervention/experiment (pretest or first time point) and then after the intervention/experiment (posttest or second time point)

---

- we subtract every person's before/after score from the before/score score to get a `difference` score

- you can subtract whatever score you want from the other (before - after or after - before)

- **mean differences** the mean of the differences between the paired scores
  - repesented as dbar

- add all the values of differences for before/after to get a sum and then divide by the number of participants (each person has a before and after score or the number of difference scores)

$$Sample\;mean\;difference = \overline{D}$$

$$Population\;mean\;difference = \mu_{D}$$

- now because we have a sample mean from a sample, we can now perform a one-sample t-test

---

# Hypotheses

- our null hypothesis is that there will be no change in scores on both occasions so there should be a difference of zero

$$H_0: \mu_{D} = 0$$

- our alternative hypothesis is that either our before or after scores should be higher
  - our hypothesis supports either a positive or negative change

$$H_1: \mu_{D} \neq 0$$

- similar to the independent samples t-test, our sampling distribution is of mean differences

---

# Steps to paired-samples t-test

Calculate the estimated variance of the difference scores

$$S^2_{D} = \frac{\Sigma D^2 - \frac{(\Sigma D)^2}{N}}{N - 1}$$

**Note** the notation changes from X to D because we are working with differences in scores and not means

---

Calculate the standard error of the mean differences

$$S_{\overline{D}} = \sqrt{\frac{S^2_{D}}{N}}$$

---

Find the obtained t-value

$$t_{obt} = \frac{\overline{D} - \mu_{D}}{S_{\overline{D}}}$$

The population difference will be zero unless you are testing for a nonzero difference

---

Calculate the degrees of freedom

$$df = N - 1$$

---

Calculate effect size

$$d = \frac{\overline{D}}{\sqrt{S^2_{D}}}$$

OR

$$r^2_{pb} = \sqrt{\frac{(t_{obt})^2}{(t_{obt})^2 + df}}$$
---

Discuss Significance

- Is the difference between time point 1 and time point 2 statistically significant?

---

# Example

```{r}
set.seed(100521)
t1 = rnorm(10, mean = 10, sd = 1.5)

t2 = rnorm(10, mean = 5.8, sd = 4)

df <- data.frame(participant = 1:10,
                 t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"),
                 score_difference_squared = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"))

```

---

```{r}
df
```


---

```{r}
df$t2

df$t1
```

---

```{r, include = FALSE}
df2 <- df %>% 
  mutate(score_difference = t2 - t1,
         score_difference_squared = score_difference^2)

df2 %>% 
  summarize(mean = mean(score_difference),
            variance = sd(score_difference)^2,
            sd = sd(score_difference))
```

---

```{r}
12.75 - 8.32
10.54 - 13.96
7.51 - 8.17
```

---

```{r}
4.32 - 8.40
9.95 - 8.90
3.64 - 7.58
```

---

```{r}
6.45 - 9.67
3.40 - 11.51
3.43 - 10.13 
13.77 - 11.49
```

---

To get your dataframe to include the numbers you just included, you need to run this again. Then it should fill in the blanks with the numbers you provided. 

```{r}

df <- data.frame(participant = 1:10,
  t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c(4.43, -3.42, -.66, -4.08, 1.05, -3.94, -3.22, -8.11, -6.7, 2.28),
                 score_difference_squared = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"))
```

---

```{r}
df$score_difference
```

---

```{r}
(4.43)^2
(-3.42)^2
(-.66)^2
```

---

```{r}
(-4.08)^2
(1.05)^2
(-3.94)^2
```

---

```{r}
(-3.22)^2
(-8.11)^2
(-6.7)^2
(2.28)^2
```

---

```{r}
df <- data.frame(participant = 1:10,
  t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c(4.43, -3.42, -.66, -4.08, 1.05, -3.94, -3.22, -8.11, -6.7, 2.28),
                 score_difference_squared = c(19.62, 11.70, .44, 16.64, 1.10, 15.52, 10.37, 65.77, 44.89, 5.20))

df
```

---

```{r}
df$score_difference

4.43 + (-3.42) + (-0.66) + (-4.08) + 1.05 + (-3.94) + (-3.22) + (-8.11) + (-6.70) + 2.28
# sum difference is -22.37

-22.37/10
# mean difference is -2.24
```

---

## Calculate the estimated variance of the difference scores

$$s^2_{D} = \frac{\Sigma D^2 - \frac{(\Sigma D)^2}{N}}{N - 1}$$

$$s^2_{D} = \frac{\Sigma D^2 - \frac{(-22.37)^2}{10}}{10 - 1}$$

---

```{r}
19.62+ 11.70+ .44+ 16.64+ 1.10+ 15.52+ 10.37+ 65.77+ 44.89+ 5.20
```

$$s^2_{D} = \frac{191.25 - \frac{(-22.37)^2}{10}}{10 - 1}$$

---

```{r}
(-22.37)^2
```

$$s^2_{D} = \frac{191.25 - \frac{500.42}{10}}{10 - 1}$$

---

```{r}
500.42/10
```

$$s^2_{D} = \frac{191.25 - 50.04}{9}$$

---

```{r}
191.25 - 50.04
```

$$s^2_{D} = \frac{141.21}{9}$$

---

```{r}
141.21/9
```

$$s^2_{D} = 15.69$$

---

## Calculate the standard error of the mean differences

$$S_{\overline{D}} = \sqrt{\frac{S^2_{D}}{N}}$$

$$S_{\overline{D}} = \sqrt{\frac{15.69}{10}}$$

---

```{r}
15.69/10
```

$$S_{\overline{D}} = \sqrt{1.57}$$

---

```{r}
sqrt(1.57)
```

$$S_{\overline{D}} = 1.25$$

---

## Find the obtained t-value

$$t_{obt} = \frac{\overline{D} - \mu_{D}}{S_{\overline{D}}}$$

$$t_{obt} = \frac{-2.24 - 0}{1.25}$$

---

```{r}
-2.24 - 0
```

$$t_{obt} = \frac{-2.24}{1.25}$$

---

```{r}
-2.24/1.25
```

$$t_{obt} = -1.79$$

---

## Calculate the degrees of freedom

$$df = N - 1$$

```{r}
10 - 1
```

$$df = 9$$

---

## Calculate effect size

$$d = \frac{\overline{D}}{\sqrt{S^2_{D}}}$$

---

```{r}
mean(df$score_difference)
sd(df$score_difference)^2


-2.24
15.69

```

$$d = \frac{-2.24}{\sqrt{15.69}}$$

---

```{r}
sqrt(15.69)
sd(df$score_difference)
```

$$d = \frac{-2.24}{3.96}$$

---

```{r}
-2.24/3.96
```

$$d = -.57$$

so really this means

$$d = .57$$

---

$$r^2_{pb} = \sqrt{\frac{(t_{obt})^2}{(t_{obt})^2 + df}}$$

$$r^2_{pb} = \sqrt{\frac{(-1.79)^2}{(-1.79)^2 + df}}$$

---

```{r}
(-1.79)^2
```

$$r^2_{pb} = \sqrt{\frac{3.20}{3.20 + 9}}$$

---

```{r}
3.20 + 9
```

$$r^2_{pb} = \sqrt{\frac{3.20}{12.20}}$$

---

```{r}
3.2/12.2
```

$$r^2_{pb} = \sqrt{.26}$$

---

```{r}
sqrt(.26)
```

$$r^2_{pb} = .51$$

---

## Discuss Significance

t-obtained value of -1.79

t-critical value of 

---

# One-tailed paired-samples t-test

- choose whether you think your `after` score would be lower/higher than the `before` score

- have your hypotheses reflect that

- if you think after scores should be higher than before (learning intervention) then the difference scores should be positive
  - because you subtracted before scores from after scores

$$H_{a}: \mu_{D} > 0$$

$$H_{0}: \mu_{D} \leq 0$$

---

# Reporting

- report in a similar style to all other t-test
  - t(df) = t value, p value
  
- you'll also report the means of the before and after scores
  - you won't report the difference between the two means
  
---

# Effect Sizes

- **effect size** is the amount of influence that changing the conditions of the IV has on the DV

- **cohen's d** is a measure of effect size in two-sample studies that reflects the magnitude of difference

  - small = .2
  
  - medium = .5
  
  - large = .8

- larger effect size means stronger the strength of the association/relationship between the IV and DV

---

# Effect Size using Proportion of Variance Accounted For

- used to determine how consistently scores change

- **proportion of variance accounted for** is the proportion of differences in DV scores assocaited with changing the conditions of the IV
  - effect size using **squared point-biserial correlation coefficient**, which indicates the proprotion of variance in DV scores that is accounted fro by IV variable in two-sample studies, are below
  
    - small = .09
    
    - moderate = .10-.25
    
    - large = .25
