---
title: "PSY 3307"
subtitle: "Two-Way Analysis of Variance"
author: "Jonathan A. Pedroza PhD"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 99999)

library(tidyverse)
library(effectsize)
```

# Agenda

- Go over the problem set
- Talk about two-way ANOVA
- Talk about how to do one-way ANOVA in SPSS
  - small introduction into two-way ANOVA

---

```{r}
jp <- rio::import(here::here("jp_thesis_1.sav")) %>% 
  janitor::clean_names() %>% 
  rowid_to_column() %>% 
  mutate(sex = ccc_gender,
         sex = recode(sex, "1" = "Male",
                      "2" = "Female"),
         sex = factor(sex),
         race = recode(ccc_ethnicity, "0" = "other", 
                       "1" = "other", 
                       "2" = "black",
                       "3" = "white",
                       "4" = "latino",
                       "5" = "other",
                       "6" = "other"),
         race = factor(race),
         bmi = ccc_bmi,
         games_avg = (mtuas_videogaming_q1 + mtuas_videogaming_q2 + mtuas_videogaming_q3)/3,
         tv_avg = (mtuas_tv_q1 + mtuas_tv_q2)/2,
         race = relevel(race, ref = "white"),
         sex = relevel(sex, ref = "Male"))
```

---

```{r, echo = FALSE, eval = TRUE}
jp %>% 
  ggplot(aes(sex, games_avg)) + 
  geom_boxplot(aes(fill = race)) +
  labs(x = "",
       y = "Video Games") +
  theme_minimal()
```

---

```{r}
games_aov <- aov(games_avg ~ sex*race, data = jp)
summary(games_aov)
```

---

```{r}
TukeyHSD(games_aov)
```

---

```{r}
effectsize::eta_squared(games_aov, alternative = "two.sided", ci = .95)
```


---

```{r}
games_lm <- lm(games_avg ~ sex*race, data = jp)
summary(games_lm)
```

---

```{r}
effectsize::eta_squared(games_lm, alternative = "two.sided", ci = .95)
```

---

```{r}
effectsize::eta_squared(games_aov, alternative = "two.sided", ci = .95)
.03 + .02 + 6.93e-03

summary(games_lm)$r.squared
```


---

```{r}
games_lm2 <- lm(games_avg ~ tv_avg*sex, data = jp)
summary(games_lm2)
```

---

```{r, echo = FALSE, eval = TRUE}
jp %>% 
  ggplot(aes(tv_avg, games_avg)) + 
  geom_jitter(alpha = .7, color = "gray70") + 
  geom_smooth(method = "lm", se = FALSE, color = "#116DA9") + 
  theme_minimal()
```

---


```{r, echo = FALSE, eval = TRUE}
jp %>% 
  ggplot(aes(tv_avg, games_avg)) + 
  geom_jitter(alpha = .7, color = "gray70") + 
  geom_smooth(method = "lm", se = FALSE, aes(color = sex)) + 
  scale_color_manual(values = c("#A91151", "#116DA9")) + 
  theme_minimal()
```

---

```{r, echo = FALSE, eval = TRUE}
jp %>% 
  ggplot(aes(tv_avg, games_avg)) + 
  geom_jitter(alpha = .7, color = "gray70") + 
  geom_smooth(method = "lm", se = FALSE, aes(color = sex)) + 
  scale_color_manual(values = c("#A91151", "#116DA9")) +
  xlim(8, 10) + 
  ylim(2.25, 5) + 
  theme_minimal()
```

---

```{r, warning = FALSE}
interactions::probe_interaction(games_lm2, pred = tv_avg, modx = sex)
```


---

$$\left[\begin{array}{cc}
-\\
Male\\
Female\\
\end{array}\right] 
\left[\begin{array}{cc} 
White\\
Male\;White\\
Female\;White\\
\end{array}\right] 
\left[\begin{array}{cc} 
Black\\
Male\;Black\\
Female\;Black\\
\end{array}\right] 
\left[\begin{array}{cc} 
Latino\\
Male\;Latino\\
Female\;Latina\\
\end{array}\right]
\left[\begin{array}{cc} 
Other\\
Male\;Other\\
Female\;Other\\
\end{array}\right]$$

---

```{r}
set.seed(102621)

table <- data.frame(sex = c("male", "male", "male", "female", "female", "female"),
                    white = rnorm(6, 10, 1.5),
                     black = rnorm(6, 5, 2.4),
                     latino = rnorm(6, 15, 3.1),
                    other = rnorm(6, 12, 5))

table
```

---

# Two-way ANOVA

- **two-way ANOVA** is a parametric ANOVA test that involves two IVs
	- one-way ANOVA is with one IV

- we will be covering a **two-way between-subjects ANOVA**, or two IVs that involve independent samples 

- we will come back to within-subjects ANOVA
	- repeated measures ANOVA
	- **two-way within-subjects ANOVA** is when every participant receives the same two IVs

- lastly, we will cover **two-way mixed-design ANOVA**, which is when one IV is between-subjects and the other IV is within-subjects

---

# Two-way ANOVA

- to understand two-way ANOVAs we will refer to the amount of levels each variable has
	- Ex: Variable 1: Sex (Male and Female) & Variable 2: Generation (baby boomers, generation X, millennials)
	- The design is a 3 x 2 (3 by 2) ANOVA, is also a 2 x 3 ANOVA (just know which variable has each number of levels)

- our two-way ANOVAs are not the same thing as an ANCOVA because we are interested in the **interaction** of our two IVs

- in a two-way ANOVA, we need to think about the **cells**, or the conbination of the two factors/IVs' levels
	- the columns are one IV
	- the rows are the other IV
	
---

# Two-way ANOVA

- when each level of both IVs are combined, this is seen as a factorial design

- assumptions of a two-way ANOVA
	- pretty much exactly the same assumptions as a one-way ANOVA
	
		+ homogenous variance
		+ interval/ratio DV
		+ normally distributed
		
---

# Main Effects

- JP Note: Just because it states it is a main effect, it is not automatically cause --> effect
	- this is just a term

- the main effect is the association between one IV and the DV
	- we have been doing this with one-one ANOVA

- In a two-way ANOVA, we get the mean for each IV
	- this is referred to as **collapsing**, where we average the scores from each level in one factor to calculate the main effect mean for the other factor
	
---

# Main Effects

- a **main effect mean** is the mean of the level of one factor after collapsing the other factor

- *to see a main effect, look at how the main effect means change as the levels of the factor change*

- simply, when we are interested in one IV and the association with the DV, we either look at our data horizontally or vertically

---

# Interactions

- the difference between a one-way ANCOVA and a two-way ANOVA is the inclusion of an interaction

- two IVs combined and teh influence of both on the DV is referred to as a two-way interaction
	- less common is a three-way interaction, which is three variables together

- rather than collapsing, we look at the cell means
	- **cell means** are the means of the scores from one cell
	
---

```{r}
set.seed(102621)

table <- data.frame(sex = c("male", "male", "male", "female", "female", "female"),
                    white = rnorm(6, 10, 1.5),
                     black = rnorm(6, 5, 2.4),
                     latino = rnorm(6, 15, 3.1))

table
```
	
---

# Interactions

- Ex: we would look at the average for white males, black males, latino males and white females, black females, latina females

- For our main effects, we compare the level means, for the interaction, we look at the cell means

- **two-way interaction effect** is when a IV-DV relationship is dependent on the other IV values

- the book has a mention of an interaction with one value going up as the other goes down
	- **but** there are times where the visual looks like an interaction but the interaction is cancelled out (I'll show an example)

- when including an interaction, the order of caring is:
	- is the interaction is significant, good
	- if not, then look at the main effects

---

# Recap

- In a two-way between-subjects ANOVA we have
	- the first main effect (IV 1 and its association with DV)
	- the second main effect (IV 2 and its association with DV)
	- the interaction effect (IV 1 x IV2 and this association with DV)

- this means that we are essentially running three different F tests 
	- we'll show this in SPSS
	- one for the first main effect
	- another for the second main effect
	- a third for the interaction
	- then whatever is leftover in the error/within

- look at the interaction first, then the main effects

---

# Looking at Main Effects

$$HSD = (q_{k})(\sqrt{\frac{MS_{wn}}{n}})$$

- we need the `k`, `df_wn`, `MS_wn`, and the `n`
	- make sure to remember that the n may be different for groups

$$\eta^2 = \frac{SS_{bn}}{SS_{total}}$$

---

# Looking at the Interaction

- along with working out the calculations, we will also visualize the interaction

- you can visualize it several different ways
	- line graph
	- boxplots
	-bar graphs

- I'll show all of these in SPSS

- parallel lines in a visualization indicates there is no interaction

---

# Tukey for Interaction

- can be used to see what specific cell means are significantly different

- it is slightly different for SPSS

- we don't compare every single cell mean though

- **counfounded comparisons** are comparisons between two cells that differ along more than one factor

- **unconfounded comparisons** are comparisons that differ along only one factor
  - this is the preferred way of looking at comparisons 
  - we'll only look at vertical or horizontal comparisons

- because we're not looking at all comparisons, we'll examine adjusted k

---

```{r}
set.seed(102621)

table <- data.frame(sex = c("male", "male", "male", "female", "female", "female"),
                    white = rnorm(6, 10, 1.5),
                     black = rnorm(6, 5, 2.4),
                     latino = rnorm(6, 15, 3.1))

table
```

---

# Interpreting Two-way ANOVA

- We'll look at the interaction first

- Then look at each main effect if there is a nonsignificant interaction

1. look at the F tests

2. look at post-hoc comparisons

3. look at effect size of each component of the model (main effects and interaction)

---

# Final Note

Because we're talking about generalized linear models, we will begin to talk about parameter estimates and what they mean

- Something super important to note is how to interpret single associations

- we will have more exposure to this in regression, but it is super helpful for interpretation

For example, if we see a significant association between sex and our outcome, we would interpret it as:

For a one point increase in X, there is a ____ increase/decrease in Y (continuous)

For a one point difference Between Males and Females, there is a ____ increase/decrease in Y (categorical)




