---
title: "PSY 3307"
subtitle: "Two Sample t-test"
author: "Jonathan A. Pedroza, MS, MA"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 99999)

library(tidyverse)
library(ggridges)
library(psych)
library(nFactors)

```

# Agenda

- Review terms
  - one-sample t test, t-distribution, df, confidence intervals
- Understanding a two-sample study
- Between & Within Designs
- Independent-samples t-test
- Performing an independent samples t-test
- Effect Size

---

# Review

- One-sample t-test

- t-distribution

- df

- confidence intervals

---

# Between & Within Designs

- Experiments can be broken down into two different types of designs

- **Between-subject/group** design is when you are interested in comparing two (for now) or more groups on an outcome variable

- **Within-subject/group** design is when you have the same participants but you test them twice (either with two different variables or two different time points)

---

# Two tests we are talking about

- **independent samples t-test** is when there are two groups of participants are separated into two different conditions to compare based on that condition
	- comparing the physical activity levels (DV) of sexes (Condition 1 = Male, Condition 2 = Female)
	- parametric test

- **paired-samples t-test** is when there are two experimental conditions that the same participants take part in
	- interested in two variables in the same sample of participants
	- can be the same variable and two different time points
	- bmi levels before an experiment and after the experiment for all participants
	- parametric test

---

# Independent Samples t-test

- JP note: probably the most often used t-test

- because it is a parametric test, it has assumptions

- Assumptions are
  - DV is normally distributed interval/ratio scores
  - populations have homogeneous variance
  - not a true assumption but something important to note is that your groups should be equal in `n` (condition) size
  
---

# Homogeneity of Variance

- **homogeneity of variance** is when the variances of the populations represented in a study have "equal" variances

- in order to test that the variances are equal, we can look at it through visuals
  - however, a better option is to use the Levene's test

---

# Independent samples t-test

- hypotheses are now focused on the differences between the two groups/conditions

$$H0: \mu_{1} - \mu_{2} = 0$$
H0: There will be no difference in DV scores between group 1 and group 2.
  - both samples/groups represent the population

$$H1: \mu_{1} - \mu_{2} \neq 0$$

H1: There will be differences in DV scores between group 1 and group 2.
  - the groups represent a different population or don't represent the current population

---

# t-distribution for independent samples t-test

- we are interested in the difference between our group/sample means

- we have two samples from one raw score population 

- **sampling distribution of differences between means** show all differences between two means that occur when random samples are drawn from a population of scores

- the mean of the sampling distribution is zero because both sample means will equal the population mean of the raw score population

---

# Independent samples t-test

- determines the probability of obtaining our difference between our means when H0 is true

- Term changes
  - N is now the full sample size
  - n is the size of each group/sample
    * so for each group/sample, we have an n
    
---

# Performing the indepdendent samples t-test

$$s^2_{x} = \frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N - 1}$$

```{r}
male_scores = c(4, 6, 2, 3, 5, 1, 2, 4, 3, 5)
female_scores = c(4, 6, 9, 6, 5, 8, 2, 5, 3, 7)

male_scores
female_scores
```

---

First we'll calculate the variance

```{r}
# male sum
4+6+2+3+5+1+2+4+3+5
# sum is 35
```

$$\;s^2_{x_{1}} = \frac{\Sigma X^2 - \frac{(35)^2}{N}}{N - 1}$$

---

```{r}

# female sum
4+6+9+6+5+8+2+5+3+7
# sum 55
```

$$\;s^2_{x_{2}} = \frac{\Sigma X^2 - \frac{(55)^2}{N}}{N - 1}$$

---

```{r}
35/10
# male mean 3.5
```


```{r}
55/10
# female mean 5.5
```

---

```{r}
# male sum of squared Xs
4^2+6^2+2^2+3^2+5^2+1^2+2^2+4^2+3^2+5^2
# 145 
```

$$\;s^2_{x_{1}} = \frac{145 - \frac{(35)^2}{10}}{10 - 1}$$

---

```{r}
# female sum of squared Xs
4^2+6^2+9^2+6^2+5^2+8^2+2^2+5^2+3^2+7^2
# female 345
```

$$\;s^2_{x_{2}} = \frac{345 - \frac{(55)^2}{10}}{10 - 1}$$

---

```{r}
# male sum of X squared and divided by N
35^2
1225/10
# 122.5
```

$$\;s^2_{x_{1}} = \frac{145 - \frac{1225}{10}}{10 - 1}$$

---

```{r}
# female sum of X squared and divided by N
55^2
3025/10
# 302.5
```

$$\;s^2_{x_{2}} = \frac{345 - \frac{302.5}{10}}{10 - 1}$$

---

```{r}
# male variance calculations
(145 - 122.5)/(10-1)
# variance is 2.5
```

$$\;s^2_{x_{1}} = \frac{145 - 122.5}{10 - 1}$$

---

```{r}
# female variance calculations
(345 - 302.5)/(10 - 1)
# variance is 4.72

```

$$\;s^2_{x_{2}} = \frac{345 - 302.5}{10 - 1}$$
---

```{r}
sd(male_scores)^2
sd(female_scores)^2
```

---

# New Terms

- **pooled variance** is the weighted average variance of the groups'/samples' variances in a independent samples t-test

- **standard error of the difference** is the estimated standard deviation of the sampling distribution of differences between the means

---

Now we can calculate the pooled variance
n1 = 10
n2 = 10
variance of group 1 = 2.5
variance of group 2 = 4.72

$$S^2_{pool} = \frac{(n_{1} - 1)S^2_{1} + (n_{2} - 1)S^2_{2}}{(n_{1} - 1) + (n_{2} - 1)}$$

---

```{r}
# start with the numerator
(10 - 1)*2.5 + (10 - 1)*4.72
# numerator is 64.98

# denominator
(10 - 1) + (10 - 1)
# denominator is 18
```

$$S^2_{pool} = \frac{(10 - 1)2.5 + (10 - 1)4.72}{(10 - 1) + (10 - 1)}$$

---

```{r}
9*2.5 + 9*4.72
# 64.98

9+9
# 18
```

$$S^2_{pool} = \frac{(9)2.5 + (9)4.72}{9 + 9}$$

---

```{r}
64.98/18
# pooled variance is 3.61
```

$$S^2_{pool} = \frac{64.98}{18}$$

---

Let's calculate for the standard error of the difference

$$S_{\overline{X_{1}} - \overline{X_{2}}} = \sqrt{(S^2_{pool})(\frac{1}{n_{1}} + \frac{1}{n_{2}})}$$

---

```{r}
1/10

3.61*(.1 + .1)

sqrt(.72)
```

$$S_{\overline{X_{1}} - \overline{X_{2}}} = \sqrt{(3.61)(\frac{1}{10} + \frac{1}{10})}$$

$$S_{\overline{X_{1}} - \overline{X_{2}}} = \sqrt{(3.61)(.1 + .1)}$$
---

```{r}
3.61*(.1+.1)
# se is .72
```

$$S_{\overline{X_{1}} - \overline{X_{2}}} = \sqrt{.72}$$

---

```{r}
sqrt(.72)
# se of the difference is .85
```
 
---

Now we can calculate the independent samples t-test obtained value

**Note** the population mean 1 minus the population mean 2 is what is specified in the null hypothesis, so it will be zero

```{r}
((3.5 - 5.5) - 0)/.85
# t obtained value is -2.35
```

$$t_{obt} = \frac{(3.5 - 5.5) - 0}{.85}$$

---

Let's now calculate the degrees of freedom

$$df = (n_{1} - 1) + (n_{2} - 1)$$
---

```{r}
(10 - 1) + (10 - 1)
# df is 18

# t critical is +-2.101
```

$$df = (10 - 1) + (10 - 1)$$

---

So we get a value of -2.35 and the t-critical value is -2.101

Is there a statistically significant difference between the two groups?

--

-2.35 > -2.101

---

Now let's get confidence intervals

```{r}
# group 1 mean = 3.5
# group 2 mean = 5.5
# t critical value is 2.101
# n1 = 10
# n2 = 10
# variance of group 1 = 2.5
# variance of group 2 = 4.72

# lower
(3.5 - 5.5) - 2.101 * sqrt((2.5/10) + (4.72/10))
# -3.79

# upper
(3.5 - 5.5) + 2.101 * sqrt((2.5/10) + (4.72/10))
# -.21
```

---

# Effect Sizes

- Reminder: r effect sizes are .1 = small, .3 = medium, .5 = large

- Reminder: cohen's d effect sizes are .2 = small, .5 = medium, .8 = large

- these are both measures of the strength of a relationship
  - better than simply using p value alone

- cohen's d can never be negative so the value you get is the absolute value (.e.g., its always positive)
  
---

$$\hat{d} = \frac{(\overline{X_{1}} - \overline{X_{2}})}{S_{2}}$$

---

```{r}
(3.5 - 5.5)/sqrt(4.72)

# each step below
3.5 - 5.5

-2/sqrt(4.72)

# cohen's d is .92 or the number of standard deviations between the means
```

$$\hat{d} = \frac{(3.5 - 5.5)}{\sqrt{4.72}}$$

---

```{r}
(-2.78)^2
(-2.78)^2 + 18
sqrt(7.73/25.73)
# r value of .55
```

---

# Steps for independent samples t-test

1. Get the means of both groups/samples

2. Get the variances of both groups/samples

3. Get the group/sample sizes (n)

4. Get the pooled variance by getting the groups'/samples' variances averaged

5. Get the standard error of the differences

6. Calculate the t-obtained value

7. Get the degrees of freedom

8. Calculate the confidence intervals

9. Get the effect size

---

# Independent samples t-test Example

```{r, eval = TRUE, echo = FALSE}
set.seed(093021)

df <- data.frame(
  sex=factor(rep(c("F", "M"), each=200)),
  weight=round(c(rnorm(200, mean=55, sd=5),
                 rnorm(200, mean=65, sd=5)))
  )

describeBy(df$weight, df$sex, na.rm = TRUE)
```

---

```{r, eval = TRUE, echo = FALSE}
df %>% 
  ggplot(aes(weight, color = sex, fill = sex)) +
  geom_density(alpha = .5) +
  geom_vline(xintercept = 54.7, color = "#d74122", size = 1.25) +
  geom_vline(xintercept = 65.36, color = "#387448", size = 1.25) +
  scale_fill_manual(values = c("#d74122", "#387448")) +
  scale_color_manual(values = c("#d74122", "#387448")) +
  theme_minimal()

```

---

```{r, eval = TRUE, echo = FALSE}
car::leveneTest(weight ~ sex, data = df)

t.test(weight ~ sex, data = df, var.equal = TRUE)
```

---

# Real life independent samples t-test

```{r, eval = TRUE, echo = FALSE}
jp <- rio::import(here::here("jp_thesis_1.sav")) %>% 
  janitor::clean_names()

jp <- jp %>% 
  mutate(stress_avg = (dass_stress_q1 + dass_stress_q2 + dass_stress_q3 + dass_stress_q4 + dass_stress_q5 + dass_stress_q6 + dass_stress_q7)/7,
         exercise_avg = (cpaq_light_exercise + cpaq_moderate_exercise + cpaq_vigorous_exercise + cpaq_muscle_exercise)/4,
         depression_avg = (dass_depress_q1 + dass_depress_q2 + dass_depress_q3 + dass_depress_q4 + dass_depress_q5 + dass_depress_q6 + dass_depress_q7)/7,
         unhealthy_snacks_avg = (ffq_cake_portion + ffq_saltysnacks_portion + ffq_chocolate_portion + ffq_icecream_portion)/4,
         games = (mtuas_videogaming_q1 + mtuas_videogaming_q2 + mtuas_videogaming_q3)/3,
         gender = ccc_gender,
         gender = factor(gender),
         class_standing = ccc_class_standing,
         bmi = ccc_bmi)
```

---

```{r, eval = TRUE, echo = FALSE}
jp %>% 
  group_by(gender) %>% 
  summarize(games = mean(games))
```

---

```{r, eval = TRUE, echo = FALSE}
jp %>% 
  ggplot(aes(games)) + 
  geom_density(aes(color = gender, fill = gender), alpha = .3) +
  theme_minimal()
```

---

```{r, eval = TRUE, echo = FALSE}
jp %>% 
  mutate(gender = recode(gender, "1" = "Male",
                         "2" = "Female")) %>% 
  ggplot(aes(games, gender)) + 
  geom_density_ridges(aes(fill = gender), alpha = .7, scale = 1) +
  geom_vline(xintercept = 4.23, color = "#d74122", size = 1.25) +
  geom_vline(xintercept = 3.3, color = "#387448", size = 1.25) +
  scale_fill_manual(values = c("#d74122", "#387448")) +
  labs(x = "Average Time Spent Playing Video Games",
       y = "") +
  theme_minimal()
```

---

```{r, eval = TRUE, echo = FALSE}
car::leveneTest(games ~ gender, data = jp)

t.test(games ~ gender, data = jp, var.equal = TRUE)
```

---

```{r, eval = TRUE, echo = FALSE}
describe(jp$games, na.rm = TRUE)
```

---

# Practice Time

```{r}
set.seed(093021)

mistakes_made_tutor = rnorm(10, mean = 1.5, sd = 1.4)
mistakes_made_control = rnorm(8, mean = 4.1, sd = 1)

mistakes_made_tutor
mistakes_made_control

```

---

```{r}
set.seed(093021)

translating_native_speaker = rnorm(9, mean = 20, sd = 4.7)
translating_non_native = rnorm(14, mean = 10, sd = .99)

translating_native_speaker
translating_non_native
```

---

```{r}
set.seed(093021)

first_gen_bmi = rnorm(6, mean = 22, sd = 2.2)
second_gen_bmi = rnorm(9, mean = 28, sd = 5)

first_gen_bmi
second_gen_bmi
```

