---
title: "PSY 3307"
subtitle: "Two Sample t-test Pt2"
author: "Jonathan A. Pedroza PhD"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 99999)

library(tidyverse)
library(ggridges)
library(psych)
library(nFactors)

```

# Agenda

- Review terms
  - independent samples t-test, homogeneity of variance, pooled variance, standard error of the difference
- Between & Within Designs
- One Tailed Independent Samples t-test
- Paired Sample t-test
- Performing a Paired Sample t-test
- Effect Size

---

# One Sample t-test

- used when we are confident that the direction of the relationship

- if you think one group will have a higher score/value then you would make that statement in your alternative hypothesis

---

# Steps for one-tailed test

- Variable(Groups) = Intervention(Got Intervention/Control)
- Outcome = Fruit & Vegetable Intake

1. decide which sample/group score you think will be larger
  - I think that the intervention group will eat more fruits and vegetables

2. decide which condition/group to subtract from the other
  - `intervention group - control group`

3. decide whether the difference will be positive or negative
  - a positive value should be returned from the previous equation
  
---

4. create hypotheses
  - The intervention group will eat the same or less fruits and vegetables than the control group
    - H0: mu(intervention group) - mu(control group) ≤ 0
  - The intervention group will eat more fruits and vegetables than the control group
    - H1: mu(intervention group) - mu(control group) > 0
    
---

5. locate regions of rejection
  - since we are interested in a positive difference in our hypothesis, we'll only be looking at the upper tail of the distribution

6. conduct your independent samples t-test
  - make sure your groups/samples are subtracted the same way as your hypotheses
    - `intervention group - control group`

---

# Paired Samples t-test

- within-subjects/groups design

- also called related-samples t-test
  - each participant gets measured in each condition

- an example would be an intervention for weight loss where everyone's weight is measured before (first time point to measure) the intervention and after (second time point to measure)

- **matched-samples design** is when a participant in one condition is matched with a participant in the other condition

- **repeated-measures design** is when each participant gets measured twice
  - can be more for more rigorous statistics, but not the ones we'll learn about in this class

---

# Why we use paired-samples t-tests

- pretest/posttest design
  - before everyone gets the intervention/experiment (pretest or first time point) and then after the intervention/experiment (posttest or second time point)

---

- we subtract every person's before/after score from the before/score score to get a `difference` score

- you can subtract whatever score you want from the other (before - after or after - before)

- **mean differences** the mean of the differences between the paired scores
  - repesented as dbar

- add all the values of differences for before/after to get a sum and then divide by the number of participants (each person has a before and after score or the number of difference scores)

$$ Sample\;mean\;difference = \overline{D} $$

$$population\;mean\;difference = \mu_{D} $$

- now because we have a sample mean from a sample, we can now perform a one-sample t-test

---

# Hypotheses

- our null hypothesis is that there will be no change in scores on both occasions so there should be a difference of zero

$$H_0: \mu_{D} = 0$$

- our alternative hypothesis is that either our before or after scores should be higher
  - our hypothesis supports either a positive or negative change

$$H_1: \mu_{D} \neq 0$$

- similar to the independent samples t-test, our sampling distribution is of mean differences

---

# Steps to paired-samples t-test

Calculate the estimated variance of the difference scores

$$s^2_{D} = \frac{\Sigma D^2 - \frac{(\Sigma D)^2}{N}}{N - 1}$$

**Note** the notation changes from X to D because we are working with differences in scores and not means

---

Calculate the standard error of the mean differences

$$S_{\overline{D}} = \sqrt{\frac{S^2_{D}}{N}}$$

---

Find the obtained t-value

$$t_{obt} = \frac{\overline{D} - \mu_{D}}{S_{\overline{D}}}$$

The population difference will be zero unless you are testing for a nonzero difference

---

Calculate the degrees of freedom

$$df = N - 1$$

---

Calculate effect size

$$d = \frac{\overline{D}}{\sqrt{S^2_{D}}}$$

OR

$$r^2_{pb} = \sqrt{\frac{(t_{obt})^2}{(t_{obt})^2 + df}}$$
---

Discuss Significance

- Is the difference between time point 1 and time point 2 statistically significant?

---

# Example

```{r}
set.seed(100521)
t1 = rnorm(10, mean = 10, sd = 1.5)

t2 = rnorm(10, mean = 5.8, sd = 4)

df <- data.frame(participant = 1:10,
                 t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"),
                 score_difference_squared = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"))

```

---

```{r}
df
```


---

```{r}
df$t2

df$t1
```

---

```{r, include = FALSE}
df2 <- df %>% 
  mutate(score_difference = t2 - t1,
         score_difference_squared = score_difference^2)

df2 %>% 
  summarize(mean = mean(score_difference),
            variance = sd(score_difference)^2,
            sd = sd(score_difference))
```

---

```{r}
12.75 - 8.32
10.54 - 13.96
7.51 - 8.17
```

---

```{r}
4.32 - 8.40
9.95 - 8.90
3.64 - 7.58
```

---

```{r}
6.45 - 9.67
3.40 - 11.51
3.43 - 10.13 
13.77 - 11.49
```

---

To get your dataframe to include the numbers you just included, you need to run this again. Then it should fill in the blanks with the numbers you provided. 

```{r}

df <- data.frame(participant = 1:10,
  t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c(4.43, -3.42, -.66, -4.08, 1.05, -3.94, -3.22, -8.11, -6.7, 2.28),
                 score_difference_squared = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"))
```

---

```{r}
df$score_difference
```

---

```{r}
(4.43)^2
(-3.42)^2
(-.66)^2
```

---

```{r}
(-4.08)^2
(1.05)^2
(-3.94)^2
```

---

```{r}
(-3.22)^2
(-8.11)^2
(-6.7)^2
(2.28)^2
```

---

```{r}
df <- data.frame(participant = 1:10,
  t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c(4.43, -3.42, -.66, -4.08, 1.05, -3.94, -3.22, -8.11, -6.7, 2.28),
                 score_difference_squared = c(19.62, 11.70, .44, 16.64, 1.10, 15.52, 10.37, 65.77, 44.89, 5.20))

df
```

---

```{r}
df$score_difference

4.43 + (-3.42) + (-0.66) + (-4.08) + 1.05 + (-3.94) + (-3.22) + (-8.11) + (-6.70) + 2.28
# sum difference is -22.37

-22.37/10
# mean difference is -2.24
```

---

## Calculate the estimated variance of the difference scores

$$s^2_{D} = \frac{\Sigma D^2 - \frac{(\Sigma D)^2}{N}}{N - 1}$$

$$s^2_{D} = \frac{\Sigma D^2 - \frac{(-22.37)^2}{10}}{10 - 1}$$

---

```{r}
19.62+ 11.70+ .44+ 16.64+ 1.10+ 15.52+ 10.37+ 65.77+ 44.89+ 5.20
```

$$s^2_{D} = \frac{191.25 - \frac{(-22.37)^2}{10}}{10 - 1}$$

---

```{r}
(-22.37)^2
```

$$s^2_{D} = \frac{191.25 - \frac{500.42}{10}}{10 - 1}$$

---

```{r}
500.42/10
```

$$s^2_{D} = \frac{191.25 - 50.04}{9}$$

---

```{r}
191.25 - 50.04
```

$$s^2_{D} = \frac{141.21}{9}$$

---

```{r}
141.21/9
```

$$s^2_{D} = 15.69$$

---

## Calculate the standard error of the mean differences

$$S_{\overline{D}} = \sqrt{\frac{S^2_{D}}{N}}$$

$$S_{\overline{D}} = \sqrt{\frac{15.69}{10}}$$

---

```{r}
15.69/10
```

$$S_{\overline{D}} = \sqrt{1.57}$$

---

```{r}
sqrt(1.57)
```

$$S_{\overline{D}} = 1.25$$

---

## Find the obtained t-value

$$t_{obt} = \frac{\overline{D} - \mu_{D}}{S_{\overline{D}}}$$

$$t_{obt} = \frac{-2.24 - 0}{1.25}$$

---

```{r}
-2.24 - 0
```

$$t_{obt} = \frac{-2.24}{1.25}$$

---

```{r}
-2.24/1.25
```

$$t_{obt} = -1.79$$

---

## Calculate the degrees of freedom

$$df = N - 1$$

```{r}
10 - 1
```

$$df = 9$$

---

## Calculate effect size

$$d = \frac{\overline{D}}{\sqrt{S^2_{D}}}$$

---

```{r}
mean(df$score_difference)
sd(df$score_difference)^2


-2.24
15.69

```

$$d = \frac{-2.24}{\sqrt{15.69}}$$

---

```{r}
sqrt(15.69)
sd(df$score_difference)
```

$$d = \frac{-2.24}{3.96}$$

---

```{r}
-2.24/3.96
```

$$d = -.57$$

so really this means

$$d = .57$$

---

$$r^2_{pb} = \sqrt{\frac{(t_{obt})^2}{(t_{obt})^2 + df}}$$

$$r^2_{pb} = \sqrt{\frac{(-1.79)^2}{(-1.79)^2 + df}}$$

---

```{r}
(-1.79)^2
```

$$r^2_{pb} = \sqrt{\frac{3.20}{3.20 + 9}}$$

---

```{r}
3.20 + 9
```

$$r^2_{pb} = \sqrt{\frac{3.20}{12.20}}$$

---

```{r}
3.2/12.2
```

$$r^2_{pb} = \sqrt{.26}$$

---

```{r}
sqrt(.26)
```

$$r^2_{pb} = .51$$

---

## Discuss Significance

t-obtained value of -1.79

t-critical value of 

---

# One-tailed paired-samples t-test

- choose whether you think your `after` score would be lower/higher than the `before` score

- have your hypotheses reflect that

- if you think after scores should be higher than before (learning intervention) then the difference scores should be positive
  - because you subtracted before scores from after scores

$$H_{a}: \mu_{D} > 0$$

$$H_{0}: \mu_{D} \leq 0$$

---

# Reporting

- report in a similar style to all other t-test
  - t(df) = t value, p value
  
- you'll also report the means of the before and after scores
  - you won't report the difference between the two means
  
---

# Effect Sizes

- **effect size** is the amount of influence that changing the conditions of the IV has on the DV

- **cohen's d** is a measure of effect size in two-sample studies that reflects the magnitude of difference

  - small = .2
  
  - medium = .5
  
  - large = .8

- larger effect size means stronger the strength of the association/relationship between the IV and DV

---

# Effect Size using Proportion of Variance Accounted For

- used to determine how consistently scores change

- **proportion of variance accounted for** is the proportion of differences in DV scores assocaited with changing the conditions of the IV
  - effect size using **squared point-biserial correlation coefficient**, which indicates the proprotion of variance in DV scores that is accounted fro by IV variable in two-sample studies, are below
  
    - small = .09
    
    - moderate = .10-.25
    
    - large = .25

---

# Practice Time

```{r}
set.seed(093021)

mistakes_made_tutor = rnorm(10, mean = 1.5, sd = 1.4)
mistakes_made_control = rnorm(8, mean = 4.1, sd = 1)

mistakes_made_tutor
mistakes_made_control

```

---

```{r}
set.seed(093021)

translating_native_speaker = rnorm(9, mean = 20, sd = 4.7)
translating_non_native = rnorm(14, mean = 10, sd = .99)

translating_native_speaker
translating_non_native
```

---

```{r}
set.seed(093021)

first_gen_bmi = rnorm(6, mean = 22, sd = 2.2)
second_gen_bmi = rnorm(9, mean = 28, sd = 5)

first_gen_bmi
second_gen_bmi
```

---

```{r}
set.seed(100521)
t1 = rnorm(10, mean = 15, sd = 2.7)

t2 = rnorm(10, mean = 12, sd = 2.1)

df1 <- data.frame(participant = 1:10,
                 t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"),
                 score_difference_squared = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"))

df1$t2
df1$t1
```

---

```{r}
set.seed(100521)
t1 = rnorm(10, mean = 8, sd = 1.2)

t2 = rnorm(10, mean = 10.5, sd = 4.8)

df11 <- data.frame(participant = 1:10,
                 t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"),
                 score_difference_squared = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"))

df11$t2
df11$t1
```

---

```{r}
set.seed(100521)
t1 = rnorm(10, mean = 15, sd = 5.6)

t2 = rnorm(10, mean = 4.3, sd = 5)

df111 <- data.frame(participant = 1:10,
                 t1 = round(t1, 2),
                 t2 = round(t2, 2),
                 score_difference = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"),
                 score_difference_squared = c("_____________", "_____________", "_____________", "_____________", "_____________",
                                     "_____________", "_____________", "_____________", "_____________", "_____________"))

df111$t2
df111$t1
```







