---
title: "PSY 3307"
subtitle: "Hypothesis Testing & The z-Test"
author: "Jonathan A. Pedroza, MS, MA"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Agenda

* The Role of Inferential Statistics in Research
* Setting Up Inferential Procedures
* Performing the z-Test
* Interpreting **STATISTICALLY** Significant and Nonsignificant Results
* Summary of the z-Test
* The One-Tailed Test
* Statistics in Research Literature: Reporting the Results
* Errors in Statistical Decision Making

---

# First

* None of this information will be on Exam 1

* This information, however, will be on Exam 2

---

# Role of Inferential Statistics

* Sampling error is always a worry when trying to compare samples to a population in a sampling distribution

* **Inferential Statistics** procedures for deciding whether sample data represents particular relationships in a given population

* **Parametric Statistics/Tests** inferential statistical tests that require assumptions; used when computing the mean
  - data should be close to normally distributed
  - scores of variables should be on an interval or ratio scale

---

* **Nonparametric Statistics/Tests** inferential statistics that do not require stringent assumptions; used with the median and mode
  - severely skewed data for interval/ratio scales
  - scores of variables are nominal or ordinal scores
    + not always the case for all nominal or ordinal scores, there are parametric tests that handle these scores (e.g., logistic regression & multinominal regression)
    
* Parametric and nonparametric tests are for deciding whether there is a relationship between variables or if sampling error is misleading us into thinking there is a relationship

---

# Setting up Inferential Procedures

For experiments, there are 4 steps in the process of conducting an experiment
  1. Create experimental hypothesis/es
  2. Design and conduct the experiment
  3. Translate the experimental hypothesis/es into statistical hypothesis/es
  4. Test the statistical hypothesis/es

* **Experimental Hypotheses** two statements describing the predicted relationship that may be demonstrated in the study
  - Don't worry about knowing this

---

# JP's Guide To Actually Learning Hypotheses

* Do your research using peer-reviewed articles and other resources (e.g., meta-analyses, systematic reviews)

* Create a hypothesis/es relevant to your study's design

* Conduct your experiment/collect your data

* Conduct your statistical analyses

* Present findings

---

* **One-tailed test** when researchers predict the direction of scores in the DV will change
  - Females will have higher test scores than males.

* **Two-tailed test** when researchers don't predict the direction of scores in the DV will change

---

# One-Sample Experiment

* testing to see if a sample's average outcome measured is significantly different from the population's average outcome
  - Giving an experimental drug for blood pressure to the sample and comparing it to the population's average blood pressure

---

# The Null Hypothesis

* **Null Hypothesis** is a statement declaring that there is not a relationship between variables
 - Often seen as `HO`

* Starting point for research, as we don't know if a relationship will be present without any additional information

* **Research Hypothesis** is a statement declaring that there is a relationship between variables
  *Alternative Hypothesis* often see as `H1`
  
---

# Examples

* There will be no differences between males and females in IQ test scores

* Non-Latina/o White adults will have more access to parks than Latina/o adults

* There will be a positive relationship between age and wisdom

* There will not be a relationship between hand size and social support

---

# Logic of Hypothesis Testing

* At least for the current statistical testing we will be conducting (z-test)

* We are interested whether our DV score in our sample is representative of the population (HO) or if our sample's DV score is significantly different from the population DV score (H1)

* This means we will be using:
  - probability
  - region of rejection
  - criterion
  - critical value
  
---

# Example 

* We are interested in knowing if CPP students eat more or less fast food than the average CSU student
  H0: 
  H1: 

* The average number of a times a week that a CSU student eats fast food is 3.7, sd is 1.2

* Sampling from CPP, we find out the average CPP student eats fast food 2.4 times a week

* We would then conduct a z-test to see if our sample is statistically significantly different from CSU students

* Before doing any analyses, we would create our criterion for what counts as a significant finding

---

# The z-Test

1. Randomly selected a sample
2. DV is somewhat normally distributed in population
  - and is ratio or interval scale
3. Know population mean of raw scores under another condition of the IV
4. Know population standard deviation

---

# Sampling Distribution of Two-tailed Test

1. Create sampling distribution of means from population raw scores
  - This will be what our H0 states
2. Identify what the population mean is for H0
3. Select an alpha
  - **alpha** greek letter for criterion probability (e.g., .05)
4. Identify regions of rejection
  - One-tailed or two-tailed test
5. Determine critical value
  - two-tailed is z = 1.96

---

# Compute the z-score

$$ \sigma_{x} = sig

$$ z_{obt} = \frac{\overline{X} - \mu}{\sigma_{\overline{X}}} $$

* Need the standard error of the mean before getting the z-score

* z is now obtained from the data, that is why it is z obt in the formula above

* From the previous example
  - mu is 3.7
  - sigma is 1.2
  
* We would calculate xbar from our sample, let's just call it 2.4

---

```{r}
# standard error calculation
1.2/sqrt(100)
# se is .12

(2.4 - 3.7)/.12
# z-score is -10.83
```

---

# Comparing Obtained z-value to Critical Calue

* We know that the critical value is +-1.96

* Since our value is outside of the critical value in the region of rejection
  - we can reject that CPP students eat fast food as much as the rest of the CSU system students

* We have rejected the null hypothesis

---

# Interpretation of STATISTICALLY Significant & Nonsignificant Results

* If you reject the null, you have statistically significant findings
  - H1 is supported, there is a relationship/there are differences

* If you retain the null, you don't have statistically significant findings
  - H0 supported, there is no relationship/there are no differences
  
---

# What Does a Statistically Significant z-test Finding Represent

* Significant MEANS NOTHING

* Statistically significant indicates you reject the null and your sample is different from the population (for z-tests)

* We can't prove that H0 is false

* Book: We do not prove that our IV caused DV scores to change

* We don't know the exact population mean represented by our sample
  - potential sampling error
  
---

# What Does a Nonsignificant z-test Finding Represent

* our sample is representative of the population
  - CPP students eat fast food as much as CSU students

* Don't say insignificant
  - best way of stating this is: "there was no evidence supporting that CPP students are statistically different from CSU students"

* We didn't find evidence of a statistically significant difference/relationship

* We simply have failed to reject the null hypothesis

---

# z-Test Summary

* Create hypothesis/es

* Set up sampling distribution, select alpha, location region of rejection, determine critical value

* compute xbar, standard error and z-score obtained from population mean and standard deviation

* compare obtained z-score to critical value
  - statistically significant = reject H0
  - nonsignificnat = retain H0
    + don't draw conclusions

---

# The One-Tailed Test

* predict scores in a direction

* interested in whether DV scores are higher or lower, not both

* the alpha still needs to be determined prior to analyses

---

# Example 

* H0: 
* H1: CSUDH students will eat more fast food than CSU students 

* CSUDH students eat 4.1 times a week

---

```{r}
# standard error calculation
1.2/sqrt(100)
# se is .12

(4.1 - 3.7)/.12
# z-score is 3.33
```

---

# Errors in Statistical Decision Making

* Make sense of the sampling error issues we've been talking about

* **Type I Errors** rejecting the null when the null hypothesis is true
  - false positive

* Know that with an alpha of 5%, 1 out of 20 statistical tests will be due to chance
  - at 1%, 1 out of 100 statistical tests will be due to chance

---

* **Note** This is why you don't test every single question of a scale. You won't know whether your finding is actually an extreme score or if you got your statistically significant finding by chance.

* Book: APA format to report *p* < .05
  - With current computing power, you should always report the exact *p* value

* **Type II Errors** retaining the null when the null hypothesis is false
  - false negative

---

* Summary
  - When null is true
    + and we reject the null = type I error
    + and we retain the null = accurate conclusion
  - when null is false
    + and we reject the null = accurate conclusion
    + and we retain the null = type II error
    
---

# Power

* **Power** is the probability that we will detect a relationship correctly by avoiding a type II error

* Power can be determined a priori (before analyses) or post hoc (after analyses)

* Ways to improve power
  - increase sample size
  - potentially use a one-tailed test over a two-tailed test
  - use parametric tests over nonparametric tests
