---
title: "PSY 3307"
subtitle: "Variability"
author: "Jonathan A. Pedroza, MS, MA"
institute: "Cal Poly Pomona"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(xaringanthemer)
style_duo(primary_color = "#1F4257", secondary_color = "#F97B64")

library(tidyverse)
library(kableExtra)
library(ggpubr)

coffee <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-07/coffee_ratings.csv') %>% 
  mutate(species = as.factor(species),
         process = recode(processing_method, "Washed / Wet" = "washed",
                          "Semi-washed / Semi-pulped" = "not_washed",
                          "Pulped natural / honey" = "not_washed",
                          "Other" = "not_washed",
                          "Natural / Dry" = "not_washed",
                          "NA" = NA_character_),
         process = as.factor(process),
         species = as.factor(species),
         country_of_origin = as.factor(country_of_origin),
         variety = as.factor(variety)) %>% 
  drop_na(process, color)

data <- read_csv("https://raw.githubusercontent.com/jpedroza1228/dissertation/master/final_data/ca_county.csv") %>% 
  janitor::clean_names() %>% 
  select(-x1, -x) %>% 
  filter(county_name != "california")

options(scipen = 999)
theme_set(theme_minimal())
```

# Agenda

* Go over problem set 1
* Understanding What Variability is
* The Range
* Sample Variance
* Sample Standard Deviation
* Standard Deviation In a Normal Distribution
* Population Variance
* Population Standard Deviation
* Calculating Variance & Standard Deviation
  - Sample Variance
  - Sample Standard Deviation
  - Population Variance
  - Population Standard Deviation
* Statistics in Research (Means & Variability)

---

# Problem Set 1

* Any questions regarding the assignment

* What parts need more clarification?

* Rating your own understanding of frequencies

* Rating your own understanding of measures of central tendency

* Who is ready to bring on the statistics?

---

# A Little Review

```{r, echo = TRUE, eval = TRUE}
numbers1 = c(7 , 6 , 3 , 3 , 1)
numbers2 = c(3 , 4 , 4 , 5 , 4)
numbers3 = c(4 , 4 , 4 , 4 , 4)

(7 + 6 + 3 + 3 + 1)/5
(3 + 4 + 4 + 5 + 4)/5
(4 + 4 + 4 + 4 + 4)/5

number_ex <- tibble(numbers1, numbers2, numbers3)

```

* What is the mean of these three different sets of numbers?
* How are they different?

---

```{r, echo = TRUE, eval = TRUE}
psych::describe(number_ex)

numbers1 - 4.0
numbers2 - 4.0
numbers3 - 4.0
```

---

# What are Measures of Variability?

* **Measures of Variability** are measures that examine how much scores differ from one another in a distribution
  - So if numbers are farther from one another, they will have more variability
  - It is also referred to as dispersion or spread; I refer to it as variation sometimes
  
* Low variability means that most values are close together, high variability means that values are spread out farther from one another

* Variability tells you how much your measure of central tendency is accurately measuring your distribution

---

# Several Measures of Variability

* **Range**

* **Variance**

* **Standard Deviation**

---

# The Range

$$ Range = highest\;score - lowest\;score $$

* **Range** tells you the distance between your highest score/value and your lowest score/value

* The book states that the use of the range is mostly for nominal or ordinal data.
  - JP disagrees. The range or more importantly, the highest and lowest scores are very important for knowing how extreme some observations are
  - Also important for seeing right away that there are outliers
  
---

# Sample Variance

* Variance tells you how different scores are from one another collectively

* This is measured by calculating how much all the scores vary/differ from the mean

* When we were looking at individual scores and how they differed from the mean, we looked at how much they deviated from the mean

* Now we are looking collectively at the sample

* We can think of variance as collective distance of the deviations from the mean

---
# Some more review

$$ X - \overline{X} $$

$$ \Sigma(X - \overline{X}) $$

* The Sum of deviations will always equal zero.

---

# What is Sample Variance

* Because the sum will always equal zero, one way to get away from that is to square the sum of deviations

$$ S^2_{x} = \frac{\Sigma(X - \overline{X})^2}{N} $$

* **Sample Variance** is the average of the squared deviations of the scores around the sample mean

* To get the variance, we will calculate the sum of the deviations squared by the total observations.

* This gives us the distance scores are on average from the mean

* One issue is that the variance is a large number, which can get confusing at times

* Also, interpretation is difficult due to being squared

---

# Sample Standard Deviation

* **Sample Standard Deviation** is the square root of the variance. It is a better measure of variability that better shows the average of the deviations from the mean

$$ S_{x} = \sqrt{\frac{\Sigma(X - \overline{X})^2}{N}} $$

* Standard deviation and variance are very similar measures of variability
  - Both measure how much scores spread out from the mean

* Best interpretation is that the scores are "on average" this far from the mean

---

# Standard Deviation and Area Under the Curve

* We can use the normal distribution/bell-shaped curve/normal curve to then add/subtract standard deviations from the mean

* If the average test score is 92 and the standard deviation is 2.5, then stating you are one standard deviation above the average score means you got a 94.5

* Going back to proportions and frequencies, knowing standard deviations means you can understand how much of a normal distribution is accounted for

* When you are looking at +1SD you have 34% of the distribution accounted for, when +-1SD you then account for 68% of the distribution

---

# Population Standard Deviation & Variance

* **Population standard deviation** is the square root of the average squared deviation of scores around the population mean

$$ \sigma_{x} = \sqrt{\frac{\Sigma(X - \mu)^2}{N}} $$

* **Population variance** the average squared deviation of scores around the population mean

$$ \sigma^2_{x} = \frac{\Sigma(X - \mu)^2}{N} $$

* These values are for the true population's measures of variability

---

# Population Standard Deviation & Variance

* Examination of the population standard deviation and variance allows for us to interpret findings by using the normal distribution

---

# Estimating the Population Variance & Standard Deviation

* The sample variance and sample standard deviations are only used to describe the variability of the sample

* They are **biased estimators**, which tend to underestimate the population variability by only using N

* We need a random sample to estimate a population, so we need a random sample of deviations

* We therefore need to get a sample of deviations, and N minus 1 reflects the variability of a population

* **Unbiased estimators** use sample data to estimate the population variability 

---

# Unbiased Estimators

$$ s^2_{x} = \frac{\Sigma(X - \overline{X})^2}{N - 1} $$

* **Estimated population variance** gives us the estimated amount of variation based on the sample we have

$$ s_{x} = \sqrt{\frac{\Sigma(X - \overline{X})^2}{N - 1}} $$

* **Estimated Population Standard Deviation** is the standardized estimated amount of variation

---

# Interpretation of Estimated Measures of Variability

* Interpretation is how we would expect the distribution to be. 

* We might expect that 68% of the distribution is within the +-1SD of the mean

---

# Summary

* Descriptive variance and standard deviation is for the sample 

* When wanting to know how much scores vary from the population, we use the population variance and standard deviation

* When we are inferring about the population based on the sample we have, we compute the unbiased estimators, essentially meaning that we calculate N - 1 in our calculations

---

# Computing the Formulas for Variance & Standard Deviation

* **Sum of Squared Xs** is calculating by squaring each score and then adding those values together

$$ \Sigma X^2 $$

* **Squared Sum of X** is calculated by adding all scores then squaring the sum

$$ (\Sigma X)^2 $$


---

* Sample Variance

$$ S^2_{x} = \frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N} $$

Step 1: Find the sum of X

Step 2: Find the sum of squared X

Step 3: Square the value in step 1

Step 4: Divide the value in step 3 by N

Step 5: Subtract Step 2 from Step 4 values

Step 6: Divide by N

---

* Sample Standard Deviation

$$ S_{x} = \sqrt{\frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N}} $$


Step 1: Find the sum of X

Step 2: Find the sum of squared X

Step 3: Square the value in step 1

Step 4: Divide the value in step 3 by N

Step 5: Subtract Step 2 from Step 4 values

Step 6: Divide by N

Step 7: get the square root of the value in Step 6

---

* Estimated Population Variance

$$ s^2_{x} = \frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N - 1} $$

Step 1: Find the sum of X

Step 2: Find the sum of squared X

Step 3: Square the value in step 1

Step 4: Divide the value in step 3 by N

Step 5: Subtract Step 2 from Step 4 values

Step 6: Subtract 1 from N

Step 7: Divide by N

---

* Estimated Population Standard Deviation

$$ s_{x} = \sqrt{\frac{\Sigma X^2 - \frac{(\Sigma X)^2}{N}}{N - 1}} $$

Step 1: Find the sum of X

Step 2: Find the sum of squared X

Step 3: Square the value in step 1

Step 4: Divide the value in step 3 by N

Step 5: Subtract Step 2 from Step 4 values

Step 6: Subtract 1 from N

Step 6: Divide by N

Step 7: get the square root of the value in Step 6



---

# Calculating with the previous formulas

$$ s^2_{x} = \frac{\Sigma(X - \overline{X})^2}{N - 1} $$

Step 1: Get the mean

Step 2: Calculate the deviates from the mean

Step 3: Square the deviates

Step 4: Calculate the sum of the squared deviates

Step 5: Subtract 1 from N

Step 6: Divide numerator by the denominator

---

$$ s_{x} = \sqrt{\frac{\Sigma(X - \overline{X})^2}{N - 1}} $$

Step 1: Get the mean

Step 2: Calculate the deviates from the mean

Step 3: Square the deviates

Step 4: Calculate the sum of the squared deviates

Step 5: Subtract 1 from N

Step 6: Divide numerator by the denominator

Step: Get the square root of the answer






